---
title: "TMT MS results plotting"
author: "Sam Siljee"
output:
  pdf_document: default
  html_document: default
---

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Packages
library(dplyr) # Data wrangling
library(tidyr) # For `pivot_longer` in labelling efficiency graph
library(stringr) # For editing strings in annotations files for matrixes
library(readxl) # Read in IPA results
library(ggplot2) # Plotting
library(ComplexHeatmap) # Making the heatmaps
library(RCy3) # Make network plots
library(biomaRt) # Uniprot/gene ID mappings
library(pathview) # Pathview plots
library(evoGO) # Got term enrichment, note package is from github devtools::install_github("Evotec-Bioinformatics/evoGO")
library(fgsea) # GSEA analysis
library(msigdbr) # GSEA gene sets
library(rvest) # Scrape hsa codes from MsigDB webpage from enriched KEGG pathways

# User variables
log_2_FC_cutoff <- log2(1.3)
adj_p_value_cutoff <- 0.05
input_dir <- paste0(getwd(), "/data/TMT/")
output_dir <- paste0(getwd(), "/../produced_plots/MS_results/TMT/")

# Files to load
processed_file <- "TMT_processed.rda"
results_file <- "MSstatsTMT_results.rda"
plot_width <- 8
plot_height <- 5.5
IPA_file <- "TMT_summary_processed.xls"
annotations_file <- "annotations.tsv"
FASTA_file <- "Human_swissprot_no_iso_20250915.fasta"
contaminants_FASTA_file <- "0602_Cell_Culture_Contaminants_GFP_to_eGFP_edit.fasta"
prots_of_interest_file <- "TMT_proteins_of_interest.txt"

# Get the number of available cores, reserve 20% for other tasks
available_cores <- floor(0.8 * parallel::detectCores())
# Set GO term domains to use
GO_domains <- c("BP", "CC", "MF")
```

## Load the data
```{r load data, include = FALSE}
# Load custom functions
source(paste0(input_dir, "../../../Functions.R"))

# Contaminant species vector
cont_species <- readLines(paste0(input_dir, contaminants_FASTA_file)) %>%
  grep(pattern = ">", value = TRUE) %>% # Extract only lines with info
  str_match(pattern = "OS=(.*?)\\s*OX=") %>% # Extract just the species
  .[, 2]

# Get list of contaminant IDs, filtering out human proteins
cont_ids <- read_FASTA_names(paste0(input_dir, contaminants_FASTA_file))[cont_species != "Homo sapiens"]

# Vector of proteins of interest - manually curated for biological interest
prots_of_interest <- readLines(paste0(input_dir, prots_of_interest_file))

# Loading the results
load(paste0(input_dir, results_file))

# Process results, filter etc
results <- as.data.frame(MSstatsTMT_results) %>%
  filter(Label == "P53 KD vs control") %>% # Only keep one comparison
  filter(!Protein %in% cont_ids) %>% # Remove contaminants
  mutate(Protein = str_remove(Protein, "Cont_"), Dif = NA) # Remove "Cont_" tag from remaining human contaminants, and reset Dif column - not properly calculated in the MSstats workflow

# Add column for up/downregulated or non-significant
results$Dif <- ifelse(
  results$log2FC > log_2_FC_cutoff & results$adj.pvalue < adj_p_value_cutoff,
  "Upregulated",
  ifelse(
    results$log2FC < -log_2_FC_cutoff & results$adj.pvalue < adj_p_value_cutoff,
    "Downregulated",
    "Not significant"
  )
)

# Add list of entrez IDs to results data
results <- merge(
  mutate(results, Protein_first = sub(";.*", "", Protein)),
  gene_map,
  by.x = "Protein_first",
  by.y = "uniprotswissprot",
  all.x = TRUE
)

# Load IPA results
sheet_names <- excel_sheets(paste0(input_dir, IPA_file)) # Get sheet names
IPA_results <- lapply(sheet_names, function(x) { # Read all sheets to list
  as.data.frame(read_excel(paste0(input_dir, IPA_file), sheet = x))
})
names(IPA_results) <- sheet_names # Rename list elements

# Load processed protein lists
load(paste0(input_dir, processed_file))

# Load annotations for matrixes
annotations <- read.delim(paste0(input_dir, annotations_file), header = TRUE) %>%
  # Add sample column to use as labels
  mutate(Sample = paste(BioReplicate, Condition, sep = "_")) %>%
  group_by(Sample) %>%
  mutate(Sample = paste0(Sample, "_", row_number())) %>%
  ungroup() %>%
  mutate(
    across(
      c(
        Condition,
        BioReplicate,
        Run,
        Sex,
        Smoker,
        Sample
      ),
      as.factor
    )
  )

# Feature level data matrix
feat_mat <- ({
  df <- MSstatsTMT_processed$FeatureLevelData %>%
    filter(BioReplicate != "Norm") %>%
    mutate(
      Sample = paste(BioReplicate, str_split_i(Condition, "_", i = 1), Run, sep = "_")
    ) %>%
    dplyr::select(PSM, Sample, log2Intensity) %>%
    pivot_wider(names_from = Sample, values_from = log2Intensity)

  df_mat <- as.matrix(df[, -1])

  row.names(df_mat) <- df$PSM

  df_mat
})

# Protein matrix for TMT data
prot_mat <- ({
  df <- MSstatsTMT_processed$ProteinLevelData %>%
    filter(BioReplicate != "Norm") %>%
    mutate(
      Sample = paste(BioReplicate, str_split_i(Condition, "_", i = 1), Run, sep = "_")
    ) %>%
    dplyr::select(Protein, Sample, Abundance) %>%
    pivot_wider(names_from = Sample, values_from = Abundance)

  df_mat <- as.matrix(df[, -1])

  row.names(df_mat) <- df$Protein

  df_mat
})

# Protein matrix by donor rather than technical replicate
donor_prot_mat <- ({
  df <- MSstatsTMT_processed$ProteinLevelData %>%
    filter(BioReplicate != "Norm") %>%
    mutate(
      Sample = paste(BioReplicate, str_split_i(Condition, "_", i = 1), sep = "_")
    ) %>%
    group_by(Sample, Protein) %>%
    summarise(Abundance = mean(Abundance, na.rm = TRUE)) %>%
    dplyr::select(Protein, Sample, Abundance) %>%
    pivot_wider(names_from = Sample, values_from = Abundance)

  df_mat <- as.matrix(df[, -1])

  row.names(df_mat) <- df$Protein

  df_mat
})

# List of differentially expressed genes for enrichment analysis
DEGs <- results %>%
  filter(Dif == "Upregulated" | Dif == "Downregulated") %>%
  .$ensembl_gene_id %>%
  unique() %>%
  na.omit()

# List of differentially expressed Uniprot IDs for network analysis
Dif_prots <- results %>%
  filter(Dif == "Upregulated" | Dif == "Downregulated") %>%
  .$Protein_first %>%
  unique() %>%
  na.omit()

# Set reference set for enrichment analysis - use only quantified proteins
reference_set <- results[!is.na(results$log2FC),]$ensembl_gene_id %>% unique() %>% na.omit()
```

## Quant numbers
Note that in the text I use numbers including the potential contaminants
```{r}
print(paste("Quantified proteins:", sum(!is.na(results$log2FC))))
print(paste("Present in only one condition:", nrow(filter(results, log2FC == Inf | log2FC == -Inf))))
print(paste("Upregulated:", sum(results$Dif == "Upregulated", na.rm = TRUE)))
print(paste("Downregulated:", sum(results$Dif == "Downregulated", na.rm = TRUE)))

# Save table for top 15 (manual) proteins of interest
write.csv(filter(results, Protein %in% prots_of_interest), paste0(output_dir, "proteins_of_interest.csv"))
# Save table for top 15 proteins by adjusted p-value
results %>%
  filter(!is.na(pvalue)) %>%
  arrange(adj.pvalue) %>%
  .[1:15,] %>%
  write.csv(paste0(output_dir, "top_dif_proteins.csv"))

# Save some spread sheets for appendixes
# .csv of protein-level abundances
write.csv(MSstatsTMT_processed$ProteinLevelData, paste0(output_dir, "protein_level_TMT_abundances.csv"))
# .csv of feature-level abundances
write.csv(MSstatsTMT_processed$FeatureLevelData, paste0(output_dir, "feature_level_TMT_abundances.csv"))
# .csv of differential abundances
write.csv(results, paste0(output_dir, "differential_abundances_TMT.csv"))
```

## Volcano plot
```{r Volcano plot, include = FALSE}
# Set colours as a named vector
colours <- c("red", "blue", "black")
names(colours) <- c("Upregulated", "Downregulated", "Not significant")

# Make the plot - TMT results
results %>%
  ggplot(aes(x = log2FC, y = -log10(adj.pvalue), col = Dif)) +
  geom_vline(xintercept = c(-log_2_FC_cutoff, log_2_FC_cutoff), linetype = "dashed", colour = "black") +
  geom_hline(yintercept = -log10(adj_p_value_cutoff), linetype = "dashed", colour = "red") +
  geom_point(alpha = 0.25, show.legend = FALSE) +
  scale_color_manual(values = colours) +
  labs(
    x = "Log 2 fold-change",
    y = "-Log10(adjusted p-value)",
    title = "Volcano plot - TMT results"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "TMT volcano plot.png"), width = plot_width, height = plot_height)

# Volcano plot of all sub-group analyses
as.data.frame(MSstatsTMT_results) %>%
  mutate(
    Dif = ifelse( # Add in Dif column for colouring
      log2FC > log_2_FC_cutoff & adj.pvalue < adj_p_value_cutoff,
      "Upregulated",
      ifelse(
        log2FC < -log_2_FC_cutoff & adj.pvalue < adj_p_value_cutoff,
        "Downregulated",
        "Not significant"
      )
    )
  ) %>%
  filter(!Protein %in% cont_ids & !is.na(Dif)) %>%
  ggplot(aes(x = log2FC, y = -log10(adj.pvalue), col = Dif)) +
  geom_vline(xintercept = c(-log_2_FC_cutoff, log_2_FC_cutoff), linetype = "dashed", colour = "black") +
  geom_hline(yintercept = -log10(adj_p_value_cutoff), linetype = "dashed", colour = "red") +
  geom_point(alpha = 0.25, show.legend = FALSE) +
  facet_wrap("Label") +
  scale_color_manual(values = colours) +
  labs(
    x = "Log 2 fold-change",
    y = "-Log10(adjusted p-value)",
    title = "Volcano plot - TMT results"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "TMT volcano subgroup plot.png"), width = plot_width, height = plot_height)
```

## PCA analysis
PCA analysis and plots, along with eigen value plots
```{r PCA}
# Run PCA analysis on feature-level data
# Run PCA analysis
TMT_PCA <- prcomp(t(na.omit(prot_mat)), center = TRUE, scale. = TRUE)

PCA_data <- as.data.frame(TMT_PCA$x) %>%
  mutate(
    Sample = rownames(TMT_PCA$x),
    BioReplicate = str_split_i(Sample, "_", i = 1),
    Condition = str_split_i(Sample, "_", i = 2),
    Sample = paste(BioReplicate, Condition, sep = "_")
  ) %>%
  merge(
    unique(
      dplyr::select(
        mutate(
          filter(annotations, BioReplicate != "Norm"),
          Sample = paste(BioReplicate, Condition, sep = "_")
        ),
        c(Sample, Smoker, Sex)
      )
    ),
    by = "Sample"
  )

# PCA plot PC1 vs PC2
PCA_data %>%
  ggplot(aes(x = PC1, y = PC2, colour = Condition, shape = BioReplicate)) +
  geom_point() +
  scale_shape_manual(values = 1:8) +
  stat_ellipse(aes(group = Condition)) +
  labs(
    x = paste0(
      "PC1 (",
      round(summary(TMT_PCA)$importance[2, 1] * 100, digits = 1),
      "%)"
    ),
    y = paste0(
      "PC2 (",
      round(summary(TMT_PCA)$importance[2, 2] * 100, digits = 1),
      "%)"
    ),
    title = "PCA plot - TMT data",
    color = "Condition",
    shape = "Donor"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "TMT PCA plot.png"), width = plot_width, height = plot_height)

# Eigen value plot = first 15 values
data.frame(
  Proportion = summary(TMT_PCA)$importance[2, 1:15],
  PC = 1:15
) %>%
  ggplot(aes(x = PC, y = Proportion)) +
  geom_col(stat = "identity") +
  labs(
    x = "Principle component",
    y = "Proportion",
    title = "Eigenvalue plot - top 15 principle components"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "TMT eigenvalue plot.png"), width = plot_width, height = plot_height)

# PCA plots by smoking status
PCA_data %>%
  ggplot(aes(x = PC1, y = PC2, colour = Smoker, shape = BioReplicate)) +
  geom_point() +
  scale_shape_manual(values = 1:8) +
  stat_ellipse(aes(group = Smoker)) +
  labs(
    x = paste0(
      "PC1 (",
      round(summary(TMT_PCA)$importance[2, 1] * 100, digits = 1),
      "%)"
    ),
    y = paste0(
      "PC2 (",
      round(summary(TMT_PCA)$importance[2, 2] * 100, digits = 1),
      "%)"
    ),
    title = "PCA plot - TMT data",
    color = "Smoker",
    shape = "Donor"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "TMT PCA smoking plot.png"), width = plot_width, height = plot_height)

# Plot by Sex
PCA_data %>%
  ggplot(aes(x = PC1, y = PC2, colour = Sex, shape = BioReplicate)) +
  geom_point() +
  scale_shape_manual(values = 1:8) +
  stat_ellipse(aes(group = Sex)) +
  labs(
    x = paste0(
      "PC1 (",
      round(summary(TMT_PCA)$importance[2, 1] * 100, digits = 1),
      "%)"
    ),
    y = paste0(
      "PC2 (",
      round(summary(TMT_PCA)$importance[2, 2] * 100, digits = 1),
      "%)"
    ),
    title = "PCA plot - TMT data",
    color = "Condition",
    shape = "Donor"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "TMT PCA sex plot.png"), width = plot_width, height = plot_height)
```

## Heatmaps
Add in a heatmap of TMT data combined to check how they cluster together. Do the same for PCA plot perhaps?
```{r Heatmaps, include = FALSE}
# TMT unfiltered heatmap - including NA's
make_custom_heatmap(
  input_matrix = prot_mat,
  include_NAs = TRUE,
  condition_annotations = str_split_i(colnames(prot_mat), "_", 2),
  donor_annotations = str_split_i(colnames(prot_mat), "_", 1),
  na_col = "gray",
  annot_col = custom_colours,
  column_title = "Tandem mass tagged proteomics heatmap",
  output_dir = paste0(output_dir, "heatmaps/"),
  file_name = "TMT_heatmap_all_missing_values.png"
)

# TMT filtered heatmap - by donor
make_custom_heatmap(
  input_matrix = donor_prot_mat,
  include_NAs = FALSE,
  filter_list = Dif_prots,
  condition_annotations = str_split_i(colnames(donor_prot_mat), "_", 2),
  donor_annotations = str_split_i(colnames(donor_prot_mat), "_", 1),
  annot_col = custom_colours,
  column_title = "TMT donors - differential proteins",
  output_dir = paste0(output_dir, "heatmaps/"),
  file_name = "TMT_heatmap_filtered_by_donor.png"
)
```

## GO term enrichment
Use evoGO to remove redundant terms, alternatively topGO could be used.
```{r evoGO}
# Get GO term annotations
if (file.exists(paste0(input_dir, "evogo__go20251010__ensembl115__specieshsapiens.rds"))) { # Load downloaded GO annotations
  GO_annotation <- loadGOAnnotation(
    species = "hsapiens",
    goRelease = "2025-10-10",
    ensemblRelease = 115,
    path = input_dir
  )
} else { # Get GO annotations
  GO_annotation <- getGOAnnotation(
    species = "hsapiens",
    goRelease = "2025-10-10",
    ensemblRelease = 115,
    nCores = available_cores,
    save = TRUE,
    path = input_dir
  )
}

# Initialise list for GO term enrichment results
evoGO_results <- list()

# Perform GO enrichment analysis - loop through domains
for (d in GO_domains) {
  evoGO_results[[d]] <- calcGOenrichment(
    goAnnotation = GO_annotation,
    deGenes = DEGs,
    universe = reference_set,
    domain = d,
    minGenes = 5,
    k = 1,
    nCores = available_cores
  )
}

# Summarise the numbers
evoGO_summary <- data.frame()

for (d in GO_domains) {
  evoGO_summary <- bind_rows(
    evoGO_summary,
    data.frame(
      Domain = d,
      Total_IDs = nrow(evoGO_results[[d]]),
      Sig_IDs = sum(evoGO_results[[d]]$fisher.pvalue < adj_p_value_cutoff),
      Reduntant_IDs = sum(evoGO_results[[d]]$fisher.pvalue < adj_p_value_cutoff) - sum(evoGO_results[[d]]$evogo.pvalue < adj_p_value_cutoff)
    )
  )
}

evoGO_summary

# Make plots for each domain
for (d in GO_domains) {
  evoGO_results[[d]] %>%
    mutate(
      Ratio = significant / annotated
    ) %>% # Add useful columns
    arrange(evogo.pvalue) %>% # Sort by p-value - evoGO algorythm
    .[1:15, ] %>% # Select the top 15
    mutate(name = as.factor(str_trunc(name, 40))) %>%
    ggplot(aes(x = Ratio, y = name, size = annotated, colour = evogo.pvalue)) +
    geom_point() +
    labs(
      title = paste0("Top 15 enriched GO terms - ", d),
      x = "Ratio",
      y = "GO term",
      colour = "evoGO algorithm p-value",
      size = "Gene count"
    ) +
    theme_bw()

  # Save the plot
  ggsave(paste0(output_dir, d, " GO term plot.png"), width = plot_width, height = plot_height)
}

# Export table of all GO term results
GO_results_tab <- data.frame()
for(d in GO_domains){
  GO_results_tab <- bind_rows(
    GO_results_tab,
    mutate(evoGO_results[[d]], domain = d)
  )
}
write.csv(GO_results_tab, paste0(output_dir, "TMT_GO_term_results.csv"))
```

## GSEA
Gene set enrichment analysis. Using fgsea.
Based on the method shown in this tutorial:
https://biostatsquid.com/fgsea-tutorial-gsea/
```{r}
# Get gene sets, use all pathways
gene_sets_df <- msigdbr(db_species = "HS", species = "Homo sapiens")

# Convert to named list format required by fgsea
gene_sets <- gene_sets_df %>%
  split(x = .$ensembl_gene, f = .$gs_name)

# Prepare list of ranked genes. Use sign from foldchange, and magnitude from p-value
rankings <- results %>%
  mutate(ranking = sign(log2FC) * -log10(pvalue)) %>%
  group_by(ensembl_gene_id) %>%
  summarise(ranking = mean(ranking, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(ensembl_gene_id) & !is.na(ranking)) %>%
  arrange(desc(ranking)) %>%
  pull(ranking, name = ensembl_gene_id)

# Run GSEA
GSEA_res <- fgsea(
  pathways = gene_sets,
  stats = rankings,
  scoreType = "std",
  minSize = 10,
  maxSize = 500,
  nPermSimple = 20000
)

# Print some stats
print("Total number of pathways:")
nrow(GSEA_res)

print("Total number of statistically significant pathways")
sum(GSEA_res$padj < 0.05)

print("Total number of statistically significant upregulated pathways")
sum(GSEA_res$padj < 0.05 & GSEA_res$ES > 0)

print("Total number of statistically significant downregulated pathways")
sum(GSEA_res$padj < 0.05 & GSEA_res$ES < 0)

top_pathways_up <- GSEA_res[ES > 0][head(order(padj), n = 15), pathway]
top_pathways_down <- GSEA_res[ES < 0][head(order(padj), n = 15), pathway]
top_pathways <- c(top_pathways_up, rev(top_pathways_down))

# Get significant pathways related to p53
p53_pathways <- GSEA_res %>%
  filter(padj < 0.05 & grepl("P53", pathway)) %>%
  .$pathway

# Plot table of results
plotGseaTable(
  pathways = gene_sets[top_pathways],
  stats = rankings,
  fgseaRes = GSEA_res,
  gseaParam = 0.5
)

# Save the plot
ggsave(paste0(output_dir, "GSEA tab.png"), width = plot_width, height = plot_height)

# Plot table of p53 results
plotGseaTable(
  pathways = gene_sets[p53_pathways],
  stats = rankings,
  fgseaRes = GSEA_res,
  gseaParam = 1
)

# Save the plot
ggsave(paste0(output_dir, "GSEA p53 tab.png"), width = plot_width, height = plot_height)

# Get adjusted p-value for "FISCHER_DIRECT_P53_TARGETS_META_ANALYSIS"
GSEA_res[GSEA_res$pathway == "FISCHER_DIRECT_P53_TARGETS_META_ANALYSIS", ]$padj

# Plot the pathway "FISCHER_DIRECT_P53_TARGETS_META_ANALYSIS"
plotEnrichment(
  gene_sets[["FISCHER_DIRECT_P53_TARGETS_META_ANALYSIS"]],
  rankings
) +
  labs(
    title = "Curated pathway: Direct p53 targets meta analysis (Fischer et. al.)",
    x = "Rank",
    y = "Enrichment score (ES)"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "GSEA p53.png"), width = plot_width/2, height = plot_height/2)

# Get list of hallmark pathways
H_paths <- GSEA_res %>%
  filter(padj < 0.05) %>%
  .$pathway %>%
  grep(pattern = "HALLMARK_", value = T)

# Plot the hallmark pathways
for (pathway in H_paths) {
  plotEnrichment(
    gene_sets[[pathway]],
    rankings
  ) +
    labs(
      title = str_replace_all(str_replace(str_to_title(pathway), "Hallmark_", ""), "_", " "),
      x = "Rank",
      y = "Enrichment score (ES)"
    ) +
    theme_bw()

  # Save the plot
  ggsave(paste0(output_dir, "GSEA ", pathway, ".png"), width = plot_width/2, height = plot_height/2)
}

# Merge GSEA results with more details from the gene sets data table and save as a .csv
gene_sets_df %>%
  dplyr::select(-c(gene_symbol, ncbi_gene, ensembl_gene, db_gene_symbol, db_ncbi_gene, db_ensembl_gene, source_gene)) %>%
  unique() %>%
  right_join(GSEA_res, by = c("gs_name" = "pathway")) %>%
  lapply(as.character) %>%
write.csv(paste0(output_dir, "TMT_GSEA_results.csv"))
```

## Pathview plots
Use the Identified KEGG pathways from GSEA for pathview analysis
```{r Pathview plots, echo = FALSE, eval = FALSE}
# Get list of significant KEGG pathways
KEGG_pathways <- GSEA_res %>%
  filter(padj < 0.05 & grepl("KEGG", pathway)) %>%
  .$pathway

# Count number of enriched pathways
print(paste("Number of enriched KEGG pathways:", length(KEGG_pathways)))

# Loop through pathway names and extract the hsa codes - scrape from MsigDB webpage
hsa_pathways <- character()
for(i in KEGG_pathways) {
  page <- read_html(paste0("https://www.gsea-msigdb.org/gsea/msigdb/human/geneset/", i))
  cells <- page %>% html_elements("td") %>% html_text(trim = TRUE)
  hsa_code <- cells[grepl("^hsa", cells)]
  hsa_pathways <- c(hsa_pathways, hsa_code)
}

# Change working directory so pathview plots save in the correct location
og_dir <- getwd()
setwd(paste0(output_dir, "pathview_plots/"))

# make plots for all pathways
for (path in hsa_pathways) {
  pathview(
    gene.data = pull(results, log2FC, name = ensembl_gene_id),
    pathway.id = path,
    species = "Homo sapiens",
    gene.idtype = "ensembl",
    same.layer = FALSE,
    kegg.native = TRUE
  )
}

# Return to original working directory
setwd(og_dir)
```

# IPA
## IPA hit summary
Summary of number of hits in each category
```{r}
# Select results to summarise and relevant stat filtering
results_to_analyse <- c("CP", "UR", "CN", "D", "TF", "RE", "N", "ITL")
stats_to_analyse <- c("-log(p-value)", "p-value of overlap", "Network bias-corrected p-value", "p-Value", "p-Value", "Consistency Score", "Score", "-log(p-value)")

# Initialise data frame
IPA_summary <- data.frame()

for (i in 1:length(results_to_analyse)) {
  IPA_summary <- bind_rows(
    IPA_summary,
    data.frame(
      Category = results_to_analyse[i],
      Stat = stats_to_analyse[i],
      "Identification number" = nrow(IPA_results[[results_to_analyse[i]]]),
      "Significant number" = sum(IPA_results[[results_to_analyse[i]]][, stats_to_analyse[i]] < 0.05)
    )
  )
}

# Manually correct for p-values using -log
IPA_summary[IPA_summary$Category == "CP", ]$Significant.number <- sum((10^-IPA_results[["CP"]]$`-log(p-value)`) < 0.05)
IPA_summary[IPA_summary$Category == "ITL", ]$Significant.number <- sum((10^-IPA_results[["ITL"]]$`-log(p-value)`) < 0.05)

# Save results
write.csv(IPA_summary, paste0(output_dir, "IPA_summary.csv"), row.names = FALSE)
```

## IPA CP
IPA Canonical Pathway
Standard GO term enrichment style plots
```{r IPA CN}
# Plot pathways, GO term style
IPA_results$CP %>%
  mutate(
    Count = str_count(Molecules, ",") + 1,
    `p-value` = 10^-`-log(p-value)`,
    `Ingenuity Canonical Pathways` = factor(`Ingenuity Canonical Pathways`,
      levels = `Ingenuity Canonical Pathways`
    )
  ) %>% # Add useful columns
  arrange(`p-value`) %>% # Sort by p-value
  .[1:15, ] %>% # Select the top 15
  ggplot(aes(x = Ratio, y = `Ingenuity Canonical Pathways`, size = Count, col = zScore)) +
  geom_point() +
  scale_color_gradient2(
    midpoint = 0,
    low = "blue",
    mid = "white",
    high = "red",
    space = "Lab"
  ) +
  labs(
    title = "Top 15 IPA canonical pathways"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "TMT IPA CP.png"), width = plot_width, height = plot_height)
```

## IPA UR
IPA Upstream Regulators
Standard GO term enrichment style plots
```{r IPA UR}
# Plot pathways, GO term style
IPA_results$UR %>%
  mutate(
    Count = str_count(`Target molecules in dataset`, ",") + 1,
    `p-value` = `p-value of overlap`,
    `Upstream Regulator` = factor(`Upstream Regulator`,
      levels = `Upstream Regulator`
    )
  ) %>% # Add useful columns
  arrange(`p-value`) %>% # Sort by p-value
  .[1:15, ] %>% # Select the top 15
  ggplot(aes(x = `Predicted Activation State`, y = `Upstream Regulator`, size = Count, col = `Bias-corrected z-score`)) +
  geom_point() +
  scale_color_gradient2(
    midpoint = 0,
    low = "blue",
    mid = "white",
    high = "red",
    space = "Lab"
  ) +
  labs(
    title = "Top 15 IPA upstream regulators"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "TMT IPA UR.png"), width = plot_width, height = plot_height)
```

## IPA CN
IPA causal networks
Standard GO term enrichment style plots
```{r IPA CN}
# Plot pathways, GO term style
IPA_results$CN %>%
  mutate(
    Count = str_count(`Target molecules in dataset`, ",") + 1,
    `p-value` = `Network bias-corrected p-value`,
    `Master Regulator` = as.factor(str_trunc(`Master Regulator`, 30))
  ) %>% # Add useful columns
  arrange(`p-value`) %>% # Sort by p-value
  .[1:15, ] %>% # Select the top 15
  ggplot(aes(x = `Predicted Activation State`, y = `Master Regulator`, size = Count, col = `Activation z-score`)) +
  geom_point() +
  scale_color_gradient2(
    midpoint = 0,
    low = "blue",
    mid = "white",
    high = "red",
    space = "Lab"
  ) +
  labs(
    title = "Top 15 IPA causal networks"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "TMT IPA CN.png"), width = plot_width, height = plot_height)
```

## IPA D
IPA diseases and bio functions
Standard GO term enrichment style plots
```{r IPA D}
# Plot pathways, GO term style
IPA_results$D %>%
  mutate(
    Count = `# Molecules`,
    `p-value` = `p-Value`,
    `Diseases or Functions Annotation` = as.factor(str_trunc(`Diseases or Functions Annotation`, 30))
  ) %>% # Add useful columns
  arrange(`p-value`) %>% # Sort by p-value
  .[1:15, ] %>% # Select the top 15
  ggplot(aes(x = `Predicted Activation State`, y = `Diseases or Functions Annotation`, size = `# Molecules`, col = `Bias-corrected z-score`)) +
  geom_point() +
  scale_color_gradient2(
    midpoint = 0,
    low = "blue",
    mid = "white",
    high = "red",
    space = "Lab"
  ) +
  labs(
    title = "Top 15 IPA diseases and bio functions"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "TMT IPA D.png"), width = plot_width, height = plot_height)
```

## IPA TF
IPA tox functions
Standard GO term enrichment style plots
```{r IPA TF}
# Plot pathways, GO term style
IPA_results$TF %>%
  mutate(
    Count = `# Molecules`,
    `p-value` = `p-Value`,
    `Diseases or Functions Annotation` = as.factor(str_trunc(`Diseases or Functions Annotation`, 30))
  ) %>% # Add useful columns
  arrange(`p-value`) %>% # Sort by p-value
  .[1:15, ] %>% # Select the top 15
  ggplot(aes(x = `Predicted Activation State`, y = `Diseases or Functions Annotation`, size = `# Molecules`, col = `Bias-corrected z-score`)) +
  geom_point() +
  scale_color_gradient2(
    midpoint = 0,
    low = "blue",
    mid = "white",
    high = "red",
    space = "Lab"
  ) +
  labs(
    title = "Top 15 IPA tox functions"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "TMT IPA TF.png"), width = plot_width, height = plot_height)
```

## IPA RE
IPA regulator effects
Standard GO term enrichment style plots
```{r IPA RE}
# Plot pathways, GO term style
IPA_results$RE %>%
  mutate(
    Count = str_count(`Target Molecules in Dataset`, ",") + 1,
    `Regulators` = as.factor(str_trunc(`Regulators`, 30))
  ) %>% # Add useful columns
  ggplot(aes(x = `Consistency Score`, y = `Regulators`, size = Count)) +
  geom_point() +
  labs(
    title = "IPA regulator effects"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "TMT IPA RE.png"), width = plot_width, height = plot_height)
```

## IPA ITL
IPA ingenuity toxicity lists
Standard GO term enrichment style plots
```{r IPA ITL}
# Plot pathways, GO term style
IPA_results$ITL %>%
  mutate(
    `p-value` = 10^-`-log(p-value)`,
    Count = str_count(`...4`, ",") + 1,
    `Ingenuity Toxicity Lists` = as.factor(str_trunc(`Ingenuity Toxicity Lists`, 30))
  ) %>% # Add useful columns
  arrange(`p-value`) %>% # Sort by p-value
  .[1:15, ] %>% # Select the top 15
  ggplot(aes(x = Molecules, y = `Ingenuity Toxicity Lists`, size = Count)) +
  geom_point() +
  labs(
    title = "Top 15 Ingenuity toxicity lists"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "TMT IPA ITL.png"), width = plot_width, height = plot_height)
```

## Network analysis
Use the RCy3 package to connect to Cytoscape and make some networks
```{r Cytoscape plots}
# Check connection to cytoscape
cytoscapePing()

# Run string analysis using command line in Cytoscape - p53 manually added
commandsRun(
  paste0(
    'string protein query query="',
    paste(c(Dif_prots, "P04637"), collapse = ","),
    '" species="Homo sapiens" cutoff=0.9'
  )
)

# Get network name to reselect it later
network_name <- getNetworkName()

# Get clusters
clustermaker2 <- commandsRun(paste0('cluster mcl network="', network_name, '"'))

# Get cluster information
clusters <- getTableColumns(table = "node", columns = c("name", "__mclCluster")) %>%
  group_by(`__mclCluster`) %>%
  summarise("Cluster_size" = n(), .groups = "drop") %>%
  filter(!is.na(`__mclCluster`)) %>%
  arrange(desc(Cluster_size))

# Count numbers of clusters with only two nodes
sum(clusters$Cluster_size == 2)

# Get node table with cluster assignments
node_table <- getTableColumns(table = "node")

# Merge fold-change data with node table
fc_tab <- node_table %>%
  left_join(mutate(results, "query term" = Protein_first), by = "query term") %>%
  group_by(name) %>%
  summarise(log2FC = mean(log2FC), .groups = "drop")

# Change Inf and -Inf values to maximum values
inf_min <- min(fc_tab$log2FC[is.finite(fc_tab$log2FC)])
inf_max <- max(fc_tab$log2FC[is.finite(fc_tab$log2FC)])
fc_tab$log2FC[fc_tab$log2FC == -Inf] <- inf_min
fc_tab$log2FC[fc_tab$log2FC == Inf] <- inf_max

# Add fold-change column
loadTableData(
  fc_tab,
  table = "node",
  data.key.column = "name",
  table.key.column = "name",
  network = network_name
)

# Create a continuous color mapping for log2FC
# Red for upregulated, blue for downregulated
fc_range <- range(fc_tab$log2FC, na.rm = TRUE)

setNodeColorMapping(
  table.column = "log2FC",
  table.column.values = c(min(fc_range), 0, max(fc_range)),
  colors = c("#0000FF", "#FFFFFF", "#FF0000"),
  mapping.type = "continuous",
  style.name = "default",
  network = network_name
)

# Set names to gene symbols
setNodeLabelMapping(
  table.column = "display name",
  style.name = "default",
  network = network_name
)

# Apply colouring style
commandsRun('vizmap apply styles="default"')

# Create subnetworks for top five clusters
for (i in 1:5) {
  cluster_id <- clusters$`__mclCluster`[i]

  # Set the network
  setCurrentNetwork(network_name)

  # Clear selection
  clearSelection(type = "both", network = network_name)

  # Select nodes
  nodes_in_cluster <- node_table$name[node_table$`__mclCluster` == cluster_id]
  selectNodes(nodes_in_cluster, by.col = "name")

  # Create subnetwork
  commandsRun(paste0(
    'network create nodeList="selected" source="', network_name, '" networkName="Cluster_',
    i, '"'
  ))

  layoutNetwork("force-directed")

  # Export subnetwork
  fitContent()

  exportImage(
    filename = file.path(output_dir, paste0("Cluster ", i, " network.png")),
    type = "PNG",
    resolution = 300,
    zoom = 200
  )

  # Wait for processes to complete
  while (!file.exists(paste0(output_dir, "Cluster ", i, " network.png"))) {
    Sys.sleep(5)
  }
}

# Export cluster data and node table to upload to Jennifer for enrichment analysis
save(clusters, file = paste0(output_dir, "clusters.rda"))
save(node_table, file = paste0(output_dir, "node_table.rda"))

# Run enrichment analysis to label clusters
for (i in 1:5) {
  cluster_id <- clusters$`__mclCluster`[i]

  cluster_DEGs <- node_table %>%
    filter(`__mclCluster` == cluster_id) %>%
    merge(
      gene_map,
      by.x = "query term",
      by.y = "uniprotswissprot",
      all.x = TRUE
    ) %>%
    .$ensembl_gene_id %>%
    na.omit() %>%
    unique()

  evoGO_res <- calcGOenrichment(
    goAnnotation = GO_annotation,
    deGenes = cluster_DEGs,
    universe = reference_set,
    domain = "BP",
    minGenes = 5,
    k = 1
  )

  # Print top three most significant pathways
  print(paste("Cluster", i, "top three BPs:"))
  print(evoGO_res %>%
    arrange(evogo.pvalue) %>%
    .$name %>%
    .[1:3])
}

# Export table of nodes for supplementary data
node_table %>%
  filter(!is.na(`__mclCluster`)) %>%
  lapply(as.character) %>%
  write.csv(paste0(output_dir, "TMT_STRING_clusters.csv"))
```
