---
title: "Phosphoproteomics results plotting"
author: "Sam Siljee"
output:
  pdf_document: default
  html_document: default
---

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Packages
library(dplyr) # Data wrangling
library(tidyr) # For `pivot_longer` in labelling efficiency graph, and separate modifications
library(stringr) # For editing strings in annotations files and PTMsigDB
library(forcats) # Reorder factor in kinase inference plot
library(readxl) # Read in IPA results
library(ggplot2) # Plotting
library(ggpubr) # Regression line for kinase z-score
library(ggVennDiagram) # Protein overlap Venn diagram
library(ComplexHeatmap) # Making the heatmaps
library(RCy3) # Make network plots
library(pathview) # Pathview plots
library(evoGO) # Got term enrichment, note package is from github pak::pak("Evotec-Bioinformatics/evoGO")
library(fgsea) # GSEA analysis
library(msigdbr) # GSEA gene sets
library(UniProt.ws) # Download UniProt data
library(benchmarKIN) # Kinase inference analysis, note package is from github remotes::install_github("saezlab/benchmarKIN")
library(org.Hs.eg.db) # Mapping for kinase pathview plot
library(gson) # read.gmt function

# User variables
log_2_FC_cutoff <- log2(1.3)
adj_p_value_cutoff <- 0.05
input_dir <- paste0(getwd(), "/data/Phos/")
output_dir <- paste0(getwd(), "/../produced_plots/MS_results/Phos/")

# Files to load
processed_file_protein <- "PTM_processed.rda"
processed_file_phosphosite <- "PTM_processed_filtered.rda"
results_file_protein <- "MSstatsPTM_results.rda"
results_file_phosphosite <- "MSstatsPTM_results_filtered.rda"
results_file_raw <- "MSstatsPTM_raw_results_filtered.rda"
sites_of_interest_file <- "Phosphosites_of_interest.txt"
IPA_file <- "Phos_summary_processed.xls"
annotations_file <- "annotations.tsv"
FASTA_file <- "Human_swissprot_no_iso_20250915.fasta"
contaminants_FASTA_file <- "0602_Cell_Culture_Contaminants_GFP_to_eGFP_edit.fasta"
UP_mods_file <- "UniProt_mods.tsv"
kinase_file <- "RoKAI_kinase_table.csv"
SIGNOR_file <- "PhosphoSIGNOR_KinaseData.tsv"
TMT_results_file <- "TMT_differential_abundances.csv"
PTMsigDB_file <- "ptm.sig.db.all.uniprot.human.v2.0.0.gmt"
plot_width <- 8
plot_height <- 5.5

# Get the number of available cores, reserve 20% for other tasks
available_cores <- floor(0.8 * parallel::detectCores())
# Set GO term domains to use
GO_domains <- c("BP", "CC", "MF")
```

## Load reference datasets
Load reference data sets
```{r Reference data sets}
# Load custom functions
source(paste0(input_dir, "../../../Functions.R"))

# Get vector of all Uniprot IDs
uniprot_ids <- c(
  read_FASTA_names(paste0(input_dir, FASTA_file)),
  read_FASTA_names(paste0(input_dir, contaminants_FASTA_file))
) %>% unique()

# Contaminant species vector
cont_species <- readLines(paste0(input_dir, contaminants_FASTA_file)) %>%
  grep(pattern = ">", value = TRUE) %>% # Extract only lines with info
  str_match(pattern = "OS=(.*?)\\s*OX=") %>% # Extract just the species
  .[, 2]

# Get list of contaminant IDs, filtering out human proteins
cont_ids <- read_FASTA_names(paste0(input_dir, contaminants_FASTA_file))[cont_species != "Homo sapiens"]

# Get list of known phosphosites from UniProt
if(file.exists(paste0(input_dir, UP_mods_file))) { # Load from local file
  UniProt_mods <- read.delim(paste0(input_dir, UP_mods_file), sep = "\t")
} else { # Load from UniProt web server
  up <- UniProt.ws(taxId = 9606)
  
  # Query UniProt to get modification data
  UniProt_mods <- select(
    up, 
    keys = str_remove(uniprot_ids, "Cont_"),
    columns = c("accession",
                "ft_mod_res",
                "organism_id"),
    keytype = "UniProtKB")
  
  # Process modifications column to get formatted phosphosites
  UniProt_mods <- UniProt_mods %>%
    filter(Organism..ID. == "9606") %>% # Remove non-human results
    mutate(Phosphosite = sapply(Modified.residue, parse_phosphosites)) %>% # Process phosphosite data
    separate_rows(Phosphosite) %>% # Split the phosphosites column into unique rows
    filter(!is.na(Phosphosite)) %>% # Remove entries without phosphosites
    unique() # Remove duplicates
  
  # Write to table
  write.table(UniProt_mods, file = paste0(input_dir, "UniProt_mods.tsv"), sep = "\t", row.names = FALSE)
}

# Get GO term annotations
if(file.exists(paste0(input_dir, "evogo__go20251010__ensembl115__specieshsapiens.rds"))) { # Load downloaded GO annotations
  GO_annotation <- loadGOAnnotation(
    species = "hsapiens",
    goRelease = "2025-10-10",
    ensemblRelease = 115,
    path = input_dir
  )
} else { # Get GO annotations
  GO_annotation <- getGOAnnotation(
    species = "hsapiens",
    goRelease = "2025-10-10",
    ensemblRelease = 115,
    nCores = available_cores,
    save = TRUE,
    path = input_dir
  )
}

# Get gene sets for GSEA, use all pathways
gene_sets_df <- msigdbr(db_species = "HS", species = "Homo sapiens")

# Convert to named list format required by fgsea
gene_sets <- gene_sets_df %>%
  split(x = .$ensembl_gene, f = .$gs_name)

# PTMsigDB dataset
PTMsigDB <- read.gmt(paste0(input_dir, PTMsigDB_file)) %>%
  mutate(gene = str_remove(gene, "-[^-]*$")) %>% # Strip up-down directionality suffix
  mutate(gene = str_replace(gene, ";", "_")) # Format to be consistent with other data

# SIGNOR dataset
SIGNOR <- read.table(paste0(input_dir, SIGNOR_file), sep = "\t", header = TRUE)
```

## Load proteomics data
Load data from my results
```{r load data, include = FALSE}
# Loading the results
load(paste0(input_dir, results_file_protein))
load(paste0(input_dir, results_file_phosphosite))
load(paste0(input_dir, results_file_raw))

# Process results, filter etc - protein level data (unfiltered)
results_protein <- as.data.frame(MSstatsPTM_results) %>%
  filter(Label == "P53 KD vs control") %>% # Only keep one comparison
  filter(!GlobalProtein %in% cont_ids) %>% # Remove contaminants
  mutate(
    Protein = str_remove(Protein, "Cont_"),
    GlobalProtein = str_remove(GlobalProtein, "Cont_"), Dif = NA) %>% # Remove "Cont_" tag from remaining human contaminants, and reset Dif column - not properly calculated in the MSstats workflow
  group_by(GlobalProtein) %>%
  slice_min(adj.pvalue, n = 1) %>% # Summarise multiple phosphosites per protein by taking teh most significant phosphosite
  ungroup()

# Process results, filter etc - phosphosite level data (filtered)
results_phosphosite <- as.data.frame(MSstatsPTM_results_filtered) %>%
  filter(Label == "P53 KD vs control") %>% # Only keep one comparison
  filter(!GlobalProtein %in% cont_ids) %>% # Remove contaminants
  mutate(
    Protein = str_remove(Protein, "Cont_"),
    GlobalProtein = str_remove(GlobalProtein, "Cont_"), Dif = NA) # Remove "Cont_" tag from remaining human contaminants, and reset Dif column - not properly calculated in the MSstats workflow

# Add column for up/downregulated or non-significant
results_protein$Dif <- ifelse(
  results_protein$log2FC > log_2_FC_cutoff & results_protein$adj.pvalue < adj_p_value_cutoff,
  "Upregulated",
  ifelse(
    results_protein$log2FC < -log_2_FC_cutoff & results_protein$adj.pvalue < adj_p_value_cutoff,
    "Downregulated",
    "Not significant"
  )
)
results_phosphosite$Dif <- ifelse(
  results_phosphosite$log2FC > log_2_FC_cutoff & results_phosphosite$adj.pvalue < adj_p_value_cutoff,
  "Upregulated",
  ifelse(
    results_phosphosite$log2FC < -log_2_FC_cutoff & results_phosphosite$adj.pvalue < adj_p_value_cutoff,
    "Downregulated",
    "Not significant"
  )
)

# Add list of entrez IDs to results data
results_protein <- merge(
  mutate(results_protein, Protein_first = sub(";.*", "", GlobalProtein)),
  gene_map,
  by.x = "Protein_first",
  by.y = "uniprotswissprot",
  all.x = TRUE
)

# Load processed protein lists
load(paste0(input_dir, processed_file_protein))
load(paste0(input_dir, processed_file_phosphosite))

# Load annotations for matrixes
annotations <- read.delim(paste0(input_dir, annotations_file), header = TRUE) %>%
  # Add sample column to use as labels
  mutate(Sample = paste(BioReplicate, Condition, sep = "_")) %>%
  group_by(Sample) %>%
  mutate(Sample = paste0(Sample, "_", row_number())) %>%
  ungroup() %>%
  mutate(
    across(
      c(
        Condition,
        BioReplicate,
        Run,
        Sex,
        Smoker,
        Sample
      ),
      as.factor
    )
  )

# Phosphosite level matrix for TMT data
phos_mat <- ({
  df <- MSstatsPTM_processed_filtered$PTM$ProteinLevelData %>%
    filter(BioReplicate != "Norm") %>%
    mutate(
      Sample = paste(BioReplicate, str_split_i(Condition, "_", i = 1), TechRepMixture, sep = "_")
    ) %>%
    dplyr::select(Protein, Sample, Abundance) %>%
    pivot_wider(names_from = Sample, values_from = Abundance)

  df_mat <- as.matrix(df[, -1])

  row.names(df_mat) <- df$Protein

  df_mat
})

# Phosphosite matrix by donor rather than technical replicate
donor_phos_mat <- ({
  df <- MSstatsPTM_processed_filtered$PTM$ProteinLevelData %>%
    filter(BioReplicate != "Norm") %>%
    mutate(
      Sample = paste(BioReplicate, str_split_i(Condition, "_", i = 1), sep = "_")
    ) %>%
    group_by(Sample, Protein) %>%
    summarise(Abundance = mean(Abundance, na.rm = TRUE)) %>%
    dplyr::select(Protein, Sample, Abundance) %>%
    pivot_wider(names_from = Sample, values_from = Abundance)

  df_mat <- as.matrix(df[, -1])

  row.names(df_mat) <- df$Protein

  df_mat
})

# Load results from RokaiXplorer
kinase_res <- read.csv(paste0(input_dir, kinase_file))

# Load TMT fold-changes
TMT_res <- read.csv(paste0(input_dir, TMT_results_file))

# Load phosphosites of interest
sites_of_interest <- readLines(paste0(input_dir, sites_of_interest_file))
```

## Quant numbers
Note that in the text I use numbers including the potential contaminants
```{r}
print("Quantified proteins:")
length(unique(results_protein$GlobalProtein))
print("Quantified phosphosites:")
length(unique(results_phosphosite$Protein))
print("Upregulated phosphosites:")
sum(all_results_phosphosite$Dif == "Upregulated", na.rm = TRUE)
print("Downregulated phosphosites:")
sum(all_results_phosphosite$Dif == "Downregulated", na.rm = TRUE)

# Save table of top 15 most significant phosphosites
merge(
  mutate(
    results_phosphosite,
    Protein_first = sub(";.*", "", GlobalProtein),
    Phosphosite = str_split_i(Protein, "_", i = 2)),
  gene_map,
  by.x = "Protein_first",
  by.y = "uniprotswissprot",
  all.x = TRUE
) %>%
  filter(!is.na(pvalue) & Protein_first != "missing") %>%
  arrange(adj.pvalue) %>%
  .[1:15,] %>%
  dplyr::select(
  "Protein_first",
  "hgnc_symbol",
  "Phosphosite",
  "log2FC",
  "adj.pvalue") %>%
  write.csv(paste0(output_dir, "top_dif_phosphosites.csv"), row.names = FALSE)

# Save table of top 15 phosphosites of interest
merge(
  mutate(
    results_phosphosite,
    Protein_first = sub(";.*", "", GlobalProtein),
    Phosphosite = str_split_i(Protein, "_", i = 2)),
  gene_map,
  by.x = "Protein_first",
  by.y = "uniprotswissprot",
  all.x = TRUE
) %>%
  filter(Protein %in% sites_of_interest) %>%
  dplyr::select(
  "Protein_first",
  "hgnc_symbol",
  "Phosphosite",
  "log2FC",
  "adj.pvalue") %>%
  write.csv(paste0(output_dir, "phosphosites_of_interest.csv"), row.names = FALSE)

# save the formatted protein level differential abundances
write.csv(results_protein, file = paste(output_dir, "Protein_level_differential_abundances.csv", sep = ""))
# save the formatted phosphosite level differential abundances
write.csv(results_phosphosite, file = paste(output_dir, "Phosphosite_level_differential_abundances.csv", sep = ""))
# .csv of protein level abundances
write.csv(MSstatsPTM_processed$PTM$ProteinLevelData, paste0(output_dir, "Protein_level_abundances.csv"))
# .csv of phosphosite level abundances
write.csv(MSstatsPTM_processed_filtered$PTM$ProteinLevelData, paste0(output_dir, "Phosphosite_level_abundances.csv"))
```

## Volcano plot
```{r Volcano plot, include = FALSE}
# Set colours as a named vector
colours <- c("red", "blue", "black")
names(colours) <- c("Upregulated", "Downregulated", "Not significant")

# Make the plot
results_phosphosite %>%
  ggplot(aes(x = log2FC, y = -log10(adj.pvalue), col = Dif)) +
  geom_vline(xintercept = c(-log_2_FC_cutoff, log_2_FC_cutoff), linetype = "dashed", colour = "black") +
  geom_hline(yintercept = -log10(adj_p_value_cutoff), linetype = "dashed", colour = "red") +
  geom_point(alpha = 0.25, show.legend = FALSE) +
  scale_color_manual(values = colours) +
  labs(
    x = "Log 2 fold-change",
    y = "-Log10(adjusted p-value)",
    title = "Volcano plot - Phosphosites"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "Phosphosite volcano plot.png"), width = plot_width, height = plot_height)

# Make the plot
results_protein %>%
  ggplot(aes(x = log2FC, y = -log10(adj.pvalue), col = Dif)) +
  geom_vline(xintercept = c(-log_2_FC_cutoff, log_2_FC_cutoff), linetype = "dashed", colour = "black") +
  geom_hline(yintercept = -log10(adj_p_value_cutoff), linetype = "dashed", colour = "red") +
  geom_point(alpha = 0.25, show.legend = FALSE) +
  scale_color_manual(values = colours) +
  labs(
    x = "Log 2 fold-change",
    y = "-Log10(adjusted p-value)",
    title = "Volcano plot - Phosphoproteins"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "Phosphoprotein volcano plot.png"), width = plot_width, height = plot_height)
```

## Phosphosite ID venn
Query Uniprot to get numbers of know phosphorylation sites. Also use this to make some reference sets.
```{r}
print("Number of UniProt phosphosites:")
print(nrow(UniProt_mods))
print("Number of unique UniProt modified proteins:")
print(length(unique(UniProt_mods$Entry)))

# Make a list of proteins identified in TMT for Venn diagram
TMT_IDs <- unique(MSstatsPTM_processed$PROTEIN$ProteinLevelData$Protein)

# Remove contaminants and take first ID
TMT_IDs <- sub(
  pattern = ";.*",
  replacement = "",
  x = TMT_IDs[!grepl(TMT_IDs, pattern = "Cont_")]
) %>%
  unique()

# Make input for Venn
venn_list <- list(
  UniProt = unique(UniProt_mods$Entry),
  PTMsigDB = unique(str_split_i(PTMsigDB$gene, "_", i = 1)),
  "Phosphoproteomics data" = unique(results_protein$Protein_first),
  "TMT data" = TMT_IDs
)

# make the plot
ggVennDiagram(venn_list) +
  theme_void() +
  ggtitle("Protein dataset overlap") +
  ggplot2::scale_fill_gradient(low = "white", high = "red")

# Save the plot
ggsave(paste0(output_dir, "protein dataset venn diagram.png"), width = plot_width, height = plot_height)
```

## PCA analysis
PCA analysis and plots, along with eigen value plots
```{r PCA}
# Run PCA analysis on phosphosite-level data
# Run PCA analysis
Phos_PCA <- prcomp(t(na.omit(phos_mat)), center = TRUE, scale. = TRUE)

PCA_data <- as.data.frame(Phos_PCA$x) %>%
  mutate(
    Sample = rownames(Phos_PCA$x),
    BioReplicate = str_split_i(Sample, "_", i = 1),
    Condition = str_split_i(Sample, "_", i = 2),
    Sample = paste(BioReplicate, Condition, sep = "_")
  ) %>%
  merge(
    unique(
      dplyr::select(
        mutate(
          filter(annotations, BioReplicate != "Norm"),
          Sample = paste(BioReplicate, Condition, sep = "_")
        ),
        c(Sample, Smoker, Sex)
      )
    ),
    by = "Sample"
  )

# PCA plot PC1 vs PC2
PCA_data %>%
  ggplot(aes(x = PC1, y = PC2, colour = Condition, shape = BioReplicate)) +
  geom_point() +
  scale_shape_manual(values = 1:8) +
  stat_ellipse(aes(group = Condition)) +
  labs(
    x = paste0(
      "PC1 (",
      round(summary(Phos_PCA)$importance[2, 1] * 100, digits = 1),
      "%)"
    ),
    y = paste0(
      "PC2 (",
      round(summary(Phos_PCA)$importance[2, 2] * 100, digits = 1),
      "%)"
    ),
    title = "PCA plot - Phosphosite data",
    color = "Condition",
    shape = "Donor"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "Phos PCA plot.png"), width = plot_width, height = plot_height)

# Eigen value plot = first 15 values
data.frame(
  Proportion = summary(Phos_PCA)$importance[2, 1:15],
  PC = 1:15
) %>%
  ggplot(aes(x = PC, y = Proportion)) +
  geom_col(stat = "identity") +
  labs(
    x = "Principle component",
    y = "Proportion",
    title = "Eigenvalue plot - top 15 principle components"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "Phos eigenvalue plot.png"), width = plot_width, height = plot_height)
```

## Heatmaps
Add in a heatmap of Phos data combined to check how they cluster together. Do the same for PCA plot perhaps?
```{r Heatmaps, include = FALSE}
# List differentially abundant phosphosites for heatmap filtering
Dif_phos <- results_phosphosite %>%
  filter(Dif == "Upregulated" | Dif == "Downregulated") %>%
  .$Protein %>%
  unique() %>%
  na.omit()

# Phos unfiltered heatmap - including NA's
make_custom_heatmap(
  input_matrix = phos_mat,
  include_NAs = TRUE,
  condition_annotations = str_split_i(colnames(phos_mat), "_", 2),
  donor_annotations = str_split_i(colnames(phos_mat), "_", 1),
  na_col = "gray",
  annot_col = custom_colours,
  column_title = "Phosphosite heatmap",
  output_dir = output_dir,
  file_name = "Phos_heatmap_all_missing_values.png"
)

# Phos filtered heatmap - by donor
make_custom_heatmap(
  input_matrix = donor_phos_mat,
  include_NAs = FALSE,
  filter_list = Dif_phos,
  condition_annotations = str_split_i(colnames(donor_phos_mat), "_", 2),
  donor_annotations = str_split_i(colnames(donor_phos_mat), "_", 1),
  annot_col = custom_colours,
  column_title = "Phosphoproteomics donors - differential phosphosites",
  output_dir = output_dir,
  file_name = "Phos_heatmap_filtered_by_donor.png"
)
```

## GO term enrichment
Use evoGO to remove redundant terms, alternatively topGO could be used.
```{r evoGO}
# List of differentially expressed genes for enrichment analysis
DEGs_protein <- results_protein %>%
  filter(Dif == "Upregulated" | Dif == "Downregulated") %>%
  .$ensembl_gene_id %>%
  unique() %>%
  na.omit()

# Reference set - all quantified gene names
reference_set_protein <- unique(results_protein$ensembl_gene_id)

# Initialise list for GO term enrichment results
evoGO_results <- list()

# Perform GO enrichment analysis - loop through domains
for (d in GO_domains) {
  evoGO_results[[d]] <- calcGOenrichment(
    goAnnotation = GO_annotation,
    deGenes = DEGs_protein,
    universe = reference_set_protein,
    domain = d,
    minGenes = 5,
    k = 1,
    nCores = available_cores
  )
}

# Summarise the numbers
evoGO_summary <- data.frame()

for (d in GO_domains) {
  evoGO_summary <- bind_rows(
    evoGO_summary,
    data.frame(
      Domain = d,
      Total_IDs = nrow(evoGO_results[[d]]),
      Sig_IDs = sum(evoGO_results[[d]]$fisher.pvalue < adj_p_value_cutoff),
      Reduntant_IDs = sum(evoGO_results[[d]]$fisher.pvalue < adj_p_value_cutoff) - sum(evoGO_results[[d]]$evogo.pvalue < adj_p_value_cutoff)
    )
  )
}

evoGO_summary

# Make plots for each domain
for (d in c("BP", "MF")) {
  evoGO_results[[d]] %>%
    mutate(
      Ratio = significant / annotated
    ) %>% # Add useful columns
    arrange(evogo.pvalue) %>% # Sort by p-value - evoGO algorythm
    .[1:15, ] %>% # Select the top 15
    mutate(name = as.factor(str_trunc(name, 40))) %>%
    ggplot(aes(x = Ratio, y = name, size = annotated, colour = evogo.pvalue)) +
    geom_point() +
    labs(
      title = paste0("Top 15 enriched GO terms - ", d),
      x = "Ratio",
      y = "GO term",
      colour = "evoGO algorithm p-value",
      size = "Gene count"
    ) +
    theme_bw()

  # Save the plot
  ggsave(paste0(output_dir, d, " GO term plot.png"), width = plot_width, height = plot_height)
}

# Manually export all 12 significant CC terms
evoGO_results[["CC"]] %>%
  mutate(
    Ratio = significant / annotated
  ) %>%
  filter(evogo.pvalue < 0.05) %>%
  mutate(name = as.factor(str_trunc(name, 40))) %>%
  ggplot(aes(x = Ratio, y = name, size = annotated, colour = evogo.pvalue)) +
  geom_point() +
  labs(
    title = paste0("Top 15 enriched GO terms - CC"),
    x = "Ratio",
    y = "GO term",
    colour = "evoGO algorithm p-value",
    size = "Gene count"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "CC GO term plot.png"), width = plot_width, height = plot_height)

# Export table of all GO term results
GO_results_tab <- data.frame()
for(d in GO_domains){
  GO_results_tab <- bind_rows(
    GO_results_tab,
    mutate(evoGO_results[[d]], domain = d)
  )
}
write.csv(GO_results_tab, paste0(output_dir, "Phos_GO_term_results.csv"))

# Check targets for potential discussion
identify_enrichment_targets(
  pathway_name = "epithelial to mesenchymal transition",
  Dif_IDs = DEGs_protein
)
```

## GSEA
Gene set enrichment analysis. Using fgsea.
Based on the method shown in this tutorial:
https://biostatsquid.com/fgsea-tutorial-gsea/
```{r}
# Prepare list of ranked genes. Use sign from foldchange, and magnitude from p-value
rankings_protein <- results_protein %>%
  mutate(ranking = sign(log2FC) * -log10(pvalue)) %>%
  filter(!is.na(ensembl_gene_id) & !is.na(ranking)) %>%
  arrange(desc(ranking)) %>%
  pull(ranking, name = ensembl_gene_id)

# Run GSEA
GSEA_res <- fgsea(
  pathways = gene_sets,
  stats = rankings_protein,
  scoreType = "std",
  minSize = 10,
  maxSize = 500,
  nPermSimple = 20000
)

# Print some numbers
print("Total number of pathways:")
nrow(GSEA_res)

print("Total number of statistically significant pathways")
sum(GSEA_res$padj < 0.05)

print("Total number of statistically significant upregulated pathways")
sum(GSEA_res$padj < 0.05 & GSEA_res$ES > 0)

print("Total number of statistically significant downregulated pathways")
sum(GSEA_res$padj < 0.05 & GSEA_res$ES < 0)

print("Pathways excluded due to multiple hypothesis testing correction:")
sum(GSEA_res$padj > 0.05 & GSEA_res$pval < 0.5)

top_pathways <- GSEA_res[padj < 0.05, pathway]

# Plot table of results
plotGseaTable(
  pathways = gene_sets[top_pathways],
  stats = rankings_protein,
  fgseaRes = GSEA_res,
  gseaParam = 0.5
)

# Save the plot
ggsave(paste0(output_dir, "GSEA tab.png"), width = plot_width, height = plot_height)

# Plot the enriched pathways
for (pathway in top_pathways) {
  plotEnrichment(
    gene_sets[[pathway]],
    rankings_protein
  ) +
    labs(
      title = str_replace_all(str_replace(str_to_upper(pathway), "Hallmark_", ""), "_", " "),
      x = "Rank",
      y = "Enrichment score (ES)"
    ) +
    theme_bw()

  # Save the plot
  ggsave(paste0(output_dir, "GSEA ", pathway, ".png"), width = plot_width, height = plot_height)
}

# Merge GSEA results with more details from the gene sets data table and save as a .csv
gene_sets_df %>%
  dplyr::select(-c(gene_symbol, ncbi_gene, ensembl_gene, db_gene_symbol, db_ncbi_gene, db_ensembl_gene, source_gene)) %>%
  unique() %>%
  right_join(GSEA_res, by = c("gs_name" = "pathway")) %>%
  lapply(as.character) %>%
write.csv(paste0(output_dir, "Phos_GSEA_results.csv"))
```

## Pathview plots
Use the Identified KEGG pathways from GSEA for pathview analysis
```{r Pathview plots, echo = FALSE, eval = FALSE}
# Get list of significant KEGG pathways
KEGG_pathways <- grep("KEGG", top_pathways, value = TRUE)

# Count number of enriched pathways
length(KEGG_pathways)

# Manually look up pathways and get hsa codes
hsa_pathways <- c(
  "hsa03040"
)

# make plots for all pathways
for(path in hsa_pathways) {
  pathview(
    gene.data = pull(results_protein, log2FC, name = ensembl_gene_id),
    pathway.id = path,
    species = "Homo sapiens",
    gene.idtype = "ensembl",
    kegg.native = TRUE
  )
}
```

# PTM-SEA
Run PTM-SEA analysis - based on GSEA, but using PTMsigDB and phosphosite level data instead.
```{r}
# Get more consistent results
set.seed(44)

# Prepare list of ranked genes. Use sign from foldchange, and magnitude from p-value
rankings_phosphosite <- results_phosphosite %>%
  mutate(ranking = sign(log2FC) * -log10(pvalue)) %>%
  filter(!is.na(Protein) & !is.na(ranking)) %>%
  arrange(desc(ranking)) %>%
  pull(ranking, name = Protein)

# Convert to named list format required by fgsea
PTM_gene_sets <- PTMsigDB %>%
  mutate(term = as.character(term)) %>%
  split(x = .$gene, f = .$term)

# Run PTM-SEA
PTM_SEA_res <- fgsea(
  pathways = PTM_gene_sets,
  stats = rankings_phosphosite,
  scoreType = "std",
  minSize = 10,
  maxSize = 500,
  nPermSimple = 20000
)

# Print some numbers
print("Total number of pathways:")
length(PTM_gene_sets)

print("Total number of pathways identified:")
nrow(PTM_SEA_res)

print("Total number of statistically significant pathways")
sum(PTM_SEA_res$padj < 0.05)

print("Total number of statistically significant upregulated pathways")
sum(PTM_SEA_res$padj < 0.05 & PTM_SEA_res$ES > 0)

print("Total number of statistically significant downregulated pathways")
sum(PTM_SEA_res$padj < 0.05 & PTM_SEA_res$ES < 0)

print("Number of sets excluded due to multiple hypopthesis testing correction:")
sum(PTM_SEA_res$padj > 0.05 & PTM_SEA_res$pval < 0.05)

PTM_top_pathways <- PTM_SEA_res[padj < 0.05, pathway]

# Plot table of results
plotGseaTable(
  pathways = PTM_gene_sets[PTM_top_pathways],
  stats = rankings_phosphosite,
  fgseaRes = PTM_SEA_res,
  gseaParam = 0.5
)

# Save the plot
ggsave(paste0(output_dir, "PTM GSEA tab.png"), width = plot_width, height = plot_height)

# Plot the enriched pathways
for (pathway in PTM_top_pathways) {
  plotEnrichment(
    PTM_gene_sets[[pathway]],
    PTM_rankings
  ) +
    labs(
      title = str_replace_all(str_replace(str_to_upper(pathway), "Hallmark_", ""), "_", " "),
      x = "Rank",
      y = "Enrichment score (ES)"
    ) +
    theme_bw()
  
  # Save the plot
  ggsave(paste0(output_dir, "GSEA PTM ", pathway, ".png"), width = plot_width, height = plot_height)
}

# Save as a .csv
PTM_SEA_res %>%
  lapply(as.character) %>%
write.csv(paste0(output_dir, "Phos_PTM-SEA_results.csv"))
```

## benchmarKIN
Infer the presence of kinases using the benchmarKIN package.
I will use the perturbation-based method.
Based on the paper by Müller-Dott et. al. 10.1038/s41467-025-59779-y
Documentation here:
https://benchmarkin.readthedocs.io/en/latest/index.html
```{r Kinase inference benchmarKIN}
# Run pertubation based testing
# Format the curated list from Müller-Dott et. al.
curated_lib <- curated %>%
  mutate(
    target = paste(target, target_protein, position, sep = "|")
  ) %>%
  dplyr::select(source, target, mor) %>%
  dplyr::distinct()

# Format input for RoKAI z-score calculation - combine corrected and uncorrected log2FCs
zs_mat <- MSstatsPTM_raw_results_filtered %>%
  filter(Label == "P53 KD vs control") %>%
  filter(!str_split_i(Protein, "_", i = 1) %in% cont_ids) %>%
  mutate(Raw = log2FC) %>%
  full_join(
    data.frame(
      Protein = results_phosphosite$Protein,
      Corrected = results_phosphosite$log2FC
    ),
    by = "Protein"
  ) %>%
  separate(
    col = Protein,
    into = c("Protein", "Phosphosite"),
    sep = "_"
  ) %>%
  left_join(
    gene_map,
    by = c("Protein" = "uniprotswissprot")
  ) %>%
  filter(!is.na(hgnc_symbol)) %>%
  group_by(hgnc_symbol, Phosphosite) %>%
  summarise(
    Raw = mean(Raw, na.rm = TRUE), 
    Corrected = mean(Corrected, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  filter(hgnc_symbol != "") %>%
  as.data.frame()

rownames(zs_mat) <- paste(paste(zs_mat$hgnc_symbol, zs_mat$Phosphosite, sep = "_"), zs_mat$hgnc_symbol, zs_mat$Phosphosite, sep = "|")

zs_mat <- zs_mat[, c("Raw", "Corrected")]

# Run pertubation-based inference
act_scores <- run_zscore(mat = zs_mat, network = curated_lib)

# Get some numbers
print("Number of kinases included in the network:")
length(unique(curated_lib$source))
print("Number of quantified kinases:")
nrow(act_scores)

# Add kinase column and reorder for plotting
act_scores$Kinase <- factor(row.names(act_scores)) %>%
  fct_reorder(act_scores$Raw, .desc = TRUE)

# Make a plot
act_scores %>%
  arrange(desc(abs(Raw))) %>% # Select the top 30 kinases by absolute z-score
  .[1:30,] %>%
  ggplot(aes(x = Kinase, y = Raw, fill = Raw)) +
  geom_col() +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0
  ) +
  labs(
    title = "Top 30 altered kinases",
    x = "Kinase",
    y = "benchmarKIN z-score",
    fill = "z-score"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# Save the plot
ggsave(paste0(output_dir, "BenchmarKIN kinase plot.png"), width = plot_width, height = plot_height)

# Save all the scores as a .csv file
act_scores %>%
write.csv(paste0(output_dir, "Phos_benchmarKIN_results.csv"), row.names = FALSE)
```

## RoKAI
Kinase inference using the original RoKAI method. Originally implemented in MATLAB, however web version available here:
https://rokai.io/
There is a newer version (the one I used) available here:
https://rokai.io/explorer/
Code to format data for RoKAI input, and make plots from output.
Original paper: 10.1038/s41467-021-21211-6
For the RokaiXplorer analysis, the configuration is available in the inputs folder
```{r Kinase activity inference - RoKAI}
# For the advanced tool:
# Raw phosphosite data
MSstatsPTM_processed_filtered$PTM$ProteinLevelData %>%
  mutate(
    Sample = paste(BioReplicate, Condition, sep = "_"),
    Abundance = 2 ^ Abundance # Reverse log2 transformation
    ) %>%
  separate(
    col = Protein,
    into = c("Protein", "Position"),
    sep = "_"
  ) %>%
  filter(!is.na(Protein) & !is.na(Position)) %>%
  arrange(Sample) %>%
  pivot_wider(
    id_cols = c(Protein, Position),
    values_from = Abundance,
    names_from = Sample,
    values_fn = mean
  ) %>%
  write.csv(paste0(output_dir, "RoKAI phosphorylation data.csv"), quote = FALSE, row.names = FALSE)

# Raw protein data
MSstatsPTM_processed$PROTEIN$ProteinLevelData %>%
  mutate(
    Sample = paste(BioReplicate, Condition, sep = "_"),
    Abundance = 2 ^ Abundance
    ) %>%
  filter(!is.na(Protein)) %>%
  arrange(Sample) %>%
  pivot_wider(
    id_cols = Protein,
    values_from = Abundance,
    names_from = Sample,
    values_fn = mean
  ) %>%
  write.csv(paste0(output_dir, "RoKAI protein expression data.csv"), quote = FALSE, row.names = FALSE)

# Metadata
annotations %>%
  filter(Condition != "Norm") %>%
  mutate(
    RowName = paste(
      BioReplicate,
      Condition,
      ifelse(Sex == "Male", "M", "F"),
      ifelse(Smoker == TRUE, "S", "NS"),
      sep = "_"
    ),
    Group = ifelse(Condition == "pLL3.7", "Control", "Case")
  ) %>%
  dplyr::select(c(RowName, Group, Condition, Smoker, Sex, BioReplicate)) %>%
  unique() %>%
  arrange(RowName) %>%
  t() %>%
  write.table(paste0(output_dir, "RoKAI metadata.csv"), quote = FALSE, sep = ",", row.names = TRUE, col.names = FALSE)

# Kinase results from RokaiXplorer
print("Number of kinases identified:")
nrow(kinase_res)
print("Number of significantly altered kinases:")
sum(kinase_res$FDR < 0.05)

# Order the kinases by z-score
kinase_res$Kinase <- factor(kinase_res$Gene) %>%
  fct_reorder(kinase_res$ZScore, .desc = TRUE)

# Make a plot
kinase_res %>%
  arrange(FDR) %>% # Select the top 30 kinases by FDR
  .[1:30,] %>%
  ggplot(aes(x = Kinase, y = ZScore, ymin = ZScore - StdErr, ymax = ZScore + StdErr, fill = ZScore)) +
  geom_col() +
  geom_errorbar(width = 0.2) +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0
  ) +
  labs(
    title = "Top 30 altered kinases",
    x = "Kinase",
    y = "RokaiXplorer z-score",
    fill = "z-score"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# Save the plot
ggsave(paste0(output_dir, "RokaiXplorer kinase plot.png"), width = plot_width, height = plot_height)

# Enrichment analysis for significantly affected kinases
# make background from all kinases in phosphositeplus and Signor datasets
kinase_ref <- c(
  curated$source,
  str_split_i(SIGNOR$entityA_name, pattern = "_", i = 1)
) %>%
  unique()

# Run evoGO enrichment analysis and print top three names by evoGO p-value
phos_ORA <- calcGOenrichment(
    goAnnotation = GO_annotation,
    deGenes = gene_map$ensembl_gene_id[gene_map$uniprotswissprot %in% kinase_res$UniprotID[kinase_res$FDR < 0.05]],
    universe = gene_map$ensembl_gene_id[gene_map$hgnc_symbol %in% kinase_ref],
    domain = "BP",
    minGenes = 5,
    k = 1
  ) 
phos_ORA %>%
  arrange(evogo.pvalue) %>%
  .$name %>%
    .[1:5]

# Plot top enrichment in inferred kinases
phos_ORA %>%
  mutate(
    Ratio = significant / annotated
  ) %>% # Add useful columns
  arrange(evogo.pvalue) %>% # Sort by p-value - evoGO algorythm
  .[1:15, ] %>% # Select the top 15
  mutate(name = as.factor(str_trunc(name, 40))) %>%
  ggplot(aes(x = Ratio, y = name, size = annotated, colour = evogo.pvalue)) +
  geom_point() +
  labs(
    title = paste0("Top 15 enriched GO terms - biological processes, inferred kinases"),
    x = "Ratio",
    y = "GO term",
    colour = "evoGO algorithm p-value",
    size = "Gene count"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "Kinase GO term BP plot.png"), width = plot_width, height = plot_height)

# pathview plot of MAPK cascase pathway - hsa04010
# Prepare data combining inferred kinase z-scores, phoospho and TMT data
path_dat <- full_join(
  TMT_res,
  kinase_res,
  by = c("Protein_first" = "UniprotID")
) %>%
  mutate(TMT = log2FC) %>%
  full_join(results, by = "Protein_first") %>%
  mutate(Phos = log2FC.y) %>%
  full_join(gene_map, by = c("Protein_first" = "uniprotswissprot")) %>%
  dplyr::select(ensembl_gene_id, ZScore, Phos, TMT) %>%
  group_by(ensembl_gene_id) %>%
  summarise(
    ZScore = mean(ZScore, na.rm = TRUE),
    Phos = mean(Phos, na.rm = TRUE),
    TMT = mean(TMT, na.rm = TRUE)
  )

path_dat_rows <- mapIds(org.Hs.eg.db,
                     keys = path_dat$ensembl_gene_id,
                     column = "ENTREZID",
                     keytype = "ENSEMBL",
                     multiVals = "first"
)

path_dat <- as.matrix(path_dat[,-1])
colnames(path_dat) <- c("z-score", "Phospho", "TMT")
rownames(path_dat) <- path_dat_rows
path_dat[is.nan(path_dat)] <- NA

# Change working directory so pathview plots save in the correct location
og_dir <- getwd()
setwd(paste0(output_dir, "pathview_plots/"))

tmp <- pathview(
  gene.data = path_dat,
  pathway.id = "hsa04010",
  species = "hsa",
  gene.idtype = "entrez",
  multi.state = TRUE,
  kegg.native = TRUE
)

# Return to original working directory
setwd(og_dir)

# Correlation between benchmarKIN and RoKAI z scores
merge(
  kinase_res,
  act_scores,
  by = "Kinase"
) %>%
  ggplot(aes(x = Raw, y = ZScore)) +
  geom_point() +
  geom_smooth(method = "lm") +
  stat_regline_equation(label.x = 1, label.y = 5) +
  theme_bw() +
  labs(
    x = "benchmarKIN z-score",
    y = "RokaiXplorer z-score",
    title = "Kinase z-score correlation"
  ) +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "Phos kinase z-score correlation.png"), width = plot_width, height = plot_height)

# Save all the scores as a .csv file
kinase_res %>%
write.csv(paste0(output_dir, "Phos_RoKAI_results.csv"), row.names = FALSE)
```

## Network analysis
Use the RCy3 package to connect to Cytoscape and make some networks
```{r Cytoscape plots}
# Get list of differential proteins
Dif_prots <- results_protein %>%
  filter(Dif != "Not significant") %>%
  .$Protein_first %>%
  unique

# Check connection to cytoscape
cytoscapePing()

# Run string analysis using command line in Cytoscape - p53 manually added
commandsRun(
  paste0(
    'string protein query query="',
    paste(c(Dif_prots, "P04637"), collapse = ","),
    '" species="Homo sapiens" cutoff=0.9'
  )
)

# Get network name to reselect it later
network_name <- getNetworkName()

# Get clusters
clustermaker2 <- commandsRun(paste0('cluster mcl network="', network_name, '"'))

# Get cluster information
clusters <- getTableColumns(table = "node", columns = c("name", "__mclCluster")) %>%
  group_by(`__mclCluster`) %>%
  summarise("Cluster_size" = n(), .groups = "drop") %>%
  filter(!is.na(`__mclCluster`)) %>%
  arrange(desc(Cluster_size))

# Count numbers of clusters with only two nodes
sum(clusters$Cluster_size == 2)

# Get node table with cluster assignments
node_table <- getTableColumns(table = "node")

# Merge fold-change data with node table
fc_tab <- node_table %>%
  left_join(mutate(results_protein, "query term" = Protein_first), by = "query term") %>%
  group_by(name) %>%
  summarise(log2FC = mean(log2FC), .groups = "drop")

# Change Inf and -Inf values to maximum values
inf_min <- min(fc_tab$log2FC[is.finite(fc_tab$log2FC)])
inf_max <- max(fc_tab$log2FC[is.finite(fc_tab$log2FC)])
fc_tab$log2FC[fc_tab$log2FC == -Inf] <- inf_min
fc_tab$log2FC[fc_tab$log2FC == Inf] <- inf_max

# Add fold-change column
loadTableData(
  fc_tab,
  table = "node",
  data.key.column = "name",
  table.key.column = "name",
  network = network_name
)

# Create a continuous color mapping for log2FC
# Red for upregulated, blue for downregulated
fc_range <- range(fc_tab$log2FC, na.rm = TRUE)

setNodeColorMapping(
  table.column = "log2FC",
  table.column.values = c(min(fc_range), 0, max(fc_range)),
  colors = c("#0000FF", "#FFFFFF", "#FF0000"),
  mapping.type = "continuous",
  style.name = "default",
  network = network_name
)

# Set names to gene symbols
setNodeLabelMapping(
  table.column = "display name",
  style.name = "default",
  network = network_name
)

# Apply colouring style
commandsRun('vizmap apply styles="default"')

# Create subnetworks for top five clusters
for (i in 1:5) {
  cluster_id <- clusters$`__mclCluster`[i]

  # Set the network
  setCurrentNetwork(network_name)

  # Clear selection
  clearSelection(type = "both", network = network_name)

  # Select nodes
  nodes_in_cluster <- node_table$name[node_table$`__mclCluster` == cluster_id]
  selectNodes(nodes_in_cluster, by.col = "name")

  # Create subnetwork
  commandsRun(paste0(
    'network create nodeList="selected" source="', network_name, '" networkName="Cluster_',
    i, '"'
  ))

  layoutNetwork("force-directed")

  # Export subnetwork
  fitContent()

  exportImage(
    filename = file.path(output_dir, paste0("Cluster ", i, " network.png")),
    type = "PNG",
    resolution = 300,
    zoom = 200
  )

  # Wait for processes to complete
  while (!file.exists(paste0(output_dir, "Cluster ", i, " network.png"))) {
    Sys.sleep(5)
  }
}

# Export cluster data and node table to upload to Jennifer for enrichment analysis
save(clusters, file = paste0(output_dir, "clusters.rda"))
save(node_table, file = paste0(output_dir, "node_table.rda"))

# Run enrichment analysis to label clusters
for (i in 1:5) {
  cluster_id <- clusters$`__mclCluster`[i]

  cluster_DEGs <- node_table %>%
    filter(`__mclCluster` == cluster_id) %>%
    merge(
      gene_map,
      by.x = "query term",
      by.y = "uniprotswissprot",
      all.x = TRUE
    ) %>%
    .$ensembl_gene_id %>%
    na.omit() %>%
    unique()

  evoGO_res <- calcGOenrichment(
    goAnnotation = GO_annotation,
    deGenes = cluster_DEGs,
    universe = reference_set_protein,
    domain = "BP",
    minGenes = 5,
    k = 1
  )

  # Print top three most significant pathways
  print(paste("Cluster", i, "top three BPs:"))
  print(evoGO_res %>%
    arrange(evogo.pvalue) %>%
    .$name %>%
    .[1:3])
}

# Export table of nodes for supplementary data
node_table %>%
  filter(!is.na(`__mclCluster`)) %>%
  lapply(as.character) %>%
  write.csv(paste0(output_dir, "Phos_STRING_clusters.csv"))
```
