---
title: "CellProfiler"
author: "Sam Siljee"
date: "2025-05-09"
output: html_document
---

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Packages
library(dplyr) # Data manipulation
library(tidyr) # pivot_longer and pivot_wider functions
library(ggplot2) # Plotting
library(ggh4x) # for independant axes in facet_grid
library(ggpubr) # Add stats to plots

# User variables
background_percentile <- 0.95
scaling_percentiles <- c(0.05, 0.95)
p63_threshold <- 0.5
basal_prop <- 0.36
SCGB1A1_threshold <- 0.5
club_prop <- 0.032
MUC5AC_threshold <- 0.5
goblet_prop <- 0.12
FOXJ1_threshold <- 0.5
ciliated_prop <- 0.33
image_area <- 0.025192038 # Area of image FoV in mm2
plot_width <- 8
plot_height <- 5.5

# Files and directories
input_dir <- paste0(getwd(), "/data/")
output_dir <- paste0(getwd(), "/../produced_plots/CellProfiler/")
annotations_file <- "annotations.tsv"
Ecad_FOXJ1_nuclei_file <- "Ecad_FOXJ1_Nuclei.csv"
p53_FOXJ1_nuclei_file <- "p53_FOXJ1_Nuclei.csv"
p53_p63_nuclei_file <- "p53_p63_Nuclei.csv"
SCGB1A1_p63_nuclei_file <- "SCGB1A1_p63_Nuclei.csv"
SCGB1A1_MUC5AC_nuclei_file <- "SCGB1A1_MUC5AC_Nuclei.csv"
SCGB1A1_p63_cells_file <- "SCGB1A1_p63_Cells.csv"
SCGB1A1_MUC5AC_cells_file <- "SCGB1A1_MUC5AC_Cells.csv"
Ecad_FOXJ1_membranes_file <- "Ecad_FOXJ1_Membranes.csv"
```

## Data loading
Load the data, and process by taking ratio of DRAQ5 expression and subtracting 95th percentile of negative control slides.
```{r Load the data}
# Load custom functions
source(paste0(input_dir, "../../Functions.R"))

# Read in annotations
annotations <- read.table(paste0(input_dir, annotations_file), sep = "\t", header = TRUE, stringsAsFactors = TRUE) %>% mutate(Sample = paste(Donor, Condition, sep = "_"))

# Read in .csv CellProfiler results and merge with annotations - combine datasets for nuclear stains
nuclei_data <- bind_rows(
  merge(read.csv(paste0(input_dir, Ecad_FOXJ1_nuclei_file)), annotations),
  merge(read.csv(paste0(input_dir, p53_FOXJ1_nuclei_file)), annotations),
  merge(read.csv(paste0(input_dir, p53_p63_nuclei_file)), annotations),
  merge(read.csv(paste0(input_dir, SCGB1A1_p63_nuclei_file)), annotations)
) %>% # Calculate ratios relative to DRAQ5
  mutate(
    FOXJ1_ratio = Intensity_MeanIntensity_FOXJ1 / Intensity_MeanIntensity_DRAQ5,
    p53_ratio = Intensity_MeanIntensity_p53 / Intensity_MeanIntensity_DRAQ5,
    p63_ratio = Intensity_MeanIntensity_p63 / Intensity_MeanIntensity_DRAQ5,
    eGFP_ratio = Intensity_MeanIntensity_eGFP / Intensity_MeanIntensity_DRAQ5,
    Slide = paste(Sample, Experiment, Negative_control)
  ) %>% # Select required columns
  dplyr::select(Sample, Donor, Condition, Negative_control, FOXJ1_ratio, p53_ratio, p63_ratio, eGFP_ratio, Intensity_MeanIntensity_FOXJ1, Intensity_MeanIntensity_p53, Intensity_MeanIntensity_p63, Intensity_MeanIntensity_eGFP, Experiment, Slide)

# Read in cells data for SCGB1A1 and MUC5AC
cells_data <- bind_rows(
  merge(read.csv(paste0(input_dir, SCGB1A1_p63_cells_file)), annotations),
  merge(read.csv(paste0(input_dir, SCGB1A1_MUC5AC_cells_file)), annotations)
) %>%
  mutate(
    Slide = paste(Sample, Experiment, Negative_control)
  ) %>% # Select required columns
  dplyr::select(Sample, Donor, Condition, Negative_control, Intensity_MeanIntensity_SCGB1A1, Intensity_MeanIntensity_MUC5AC, Intensity_MeanIntensity_eGFP, Experiment, Slide)

# Read in membranes data for E-cadherin
membranes_data <- merge(read.csv(paste0(input_dir, Ecad_FOXJ1_membranes_file)), annotations) %>%
  mutate(
    Slide = paste(Sample, Experiment, Negative_control)
  ) %>% # Select required columns
  dplyr::select(Sample, Donor, Condition, Negative_control, Intensity_MeanIntensity_Ecadherin, Intensity_MeanIntensity_eGFP, Experiment, Slide)
```

## Data processing
```{r Process the data}
# Process nuclei data
# Calculate negative control 95th percentiles
nuclei_data_neg_summary <- nuclei_data %>%
  filter(Negative_control == TRUE) %>%
  group_by(Sample, Experiment) %>%
  summarise(
    FOXJ1_95_quant = quantile(FOXJ1_ratio, probs = background_percentile, na.rm = TRUE),
    p53_95_quant = quantile(p53_ratio, probs = background_percentile, na.rm = TRUE),
    p63_95_quant = quantile(p63_ratio, probs = background_percentile, na.rm = TRUE),
    eGFP_95_quant = quantile(eGFP_ratio, probs = background_percentile, na.rm = TRUE)
  )

# Manually add negative controls for SCGB1A1/p63 based on p53/p63 (same confocal settings used)
nuclei_data_neg_summary <- nuclei_data_neg_summary %>%
  filter(Experiment == "p53 p63") %>%
  mutate(Experiment = "SCGB1A1 p63") %>%
  bind_rows(
    nuclei_data_neg_summary
  )

# Subtract negative control 95th percentiles
nuclei_data <- merge(nuclei_data, nuclei_data_neg_summary, by = c("Sample", "Experiment")) %>%
  mutate(
    FOXJ1_ratio_corrected = FOXJ1_ratio - FOXJ1_95_quant,
    p53_ratio_corrected = p53_ratio - p53_95_quant,
    p63_ratio_corrected = p63_ratio - p63_95_quant,
    eGFP_ratio_corrected = eGFP_ratio - eGFP_95_quant
  )

# Scale 95th and 5th percentiles from 0 to 1 by slide
nuclei_data <- nuclei_data %>%
  group_by(Slide) %>%
  mutate(
    FOXJ1_scaled = FOXJ1_ratio_corrected - quantile(FOXJ1_ratio_corrected, probs = scaling_percentiles[1], na.rm = TRUE),
    p53_scaled = p53_ratio_corrected - quantile(p53_ratio_corrected, probs = scaling_percentiles[1], na.rm = TRUE),
    p63_scaled = p63_ratio_corrected - quantile(p63_ratio_corrected, probs = scaling_percentiles[1], na.rm = TRUE),
    eGFP_scaled = eGFP_ratio_corrected - quantile(eGFP_ratio_corrected, probs = scaling_percentiles[1], na.rm = TRUE)
  ) %>%
  mutate(
    FOXJ1_scaled = FOXJ1_scaled / quantile(FOXJ1_scaled, probs = scaling_percentiles[2], na.rm = TRUE),
    p53_scaled = p53_scaled / quantile(p53_scaled, probs = scaling_percentiles[2], na.rm = TRUE),
    p63_scaled = p63_scaled / quantile(p63_scaled, probs = scaling_percentiles[2], na.rm = TRUE),
    eGFP_scaled = eGFP_scaled / quantile(eGFP_scaled, probs = scaling_percentiles[2], na.rm = TRUE)
  ) %>%
  ungroup()

# Calculate eGFP expression quartiles by slide
nuclei_data <- nuclei_data %>%
  group_by(Slide) %>%
  mutate(eGFP_quartile = ifelse(
    eGFP_scaled < quantile(eGFP_scaled, probs = 0.25, na.rm = TRUE),
    "Lower",
    ifelse(
      eGFP_scaled > quantile(eGFP_scaled, probs = 0.75, na.rm = TRUE),
      "Upper",
      "Middle"
    )
  ))

# Process cells data
# Calculate 95th percentiles to subtract background
cells_data_neg_summary <- cells_data %>%
  filter(Negative_control == TRUE) %>%
  group_by(Sample) %>%
  summarise(
    SCGB1A1_95_quant = quantile(Intensity_MeanIntensity_SCGB1A1, probs = 0.95, na.rm = TRUE),
    MUC5AC_95_quant = quantile(Intensity_MeanIntensity_MUC5AC, probs = 0.95, na.rm = TRUE),
    eGFP_95_quant = quantile(Intensity_MeanIntensity_eGFP, probs = 0.95, na.rm = TRUE)
  )

# Subtract background - using SCGB1A1/MUC5AC for SCGB1A1/p63; the same confocal settings were used
cells_data <- merge(cells_data, cells_data_neg_summary, by = "Sample") %>%
  mutate(
    SCGB1A1_corrected = Intensity_MeanIntensity_SCGB1A1 - SCGB1A1_95_quant,
    MUC5AC_corrected = Intensity_MeanIntensity_MUC5AC - MUC5AC_95_quant,
    eGFP_corrected = Intensity_MeanIntensity_eGFP - eGFP_95_quant
  )

# Scale 95th and 5th percentiles from 0 to 1 by slide
cells_data <- cells_data %>%
  group_by(Slide) %>%
  mutate(
    SCGB1A1_scaled = SCGB1A1_corrected - quantile(SCGB1A1_corrected, probs = scaling_percentiles[1], na.rm = TRUE),
    MUC5AC_scaled = MUC5AC_corrected - quantile(MUC5AC_corrected, probs = scaling_percentiles[1], na.rm = TRUE),
    eGFP_scaled = eGFP_corrected - quantile(eGFP_corrected, probs = scaling_percentiles[1], na.rm = TRUE)
  ) %>%
  mutate(
    SCGB1A1_scaled = SCGB1A1_scaled / quantile(SCGB1A1_scaled, probs = scaling_percentiles[2], na.rm = TRUE),
    MUC5AC_scaled = MUC5AC_scaled / quantile(MUC5AC_scaled, probs = scaling_percentiles[2], na.rm = TRUE),
    eGFP_scaled = eGFP_scaled / quantile(eGFP_scaled, probs = scaling_percentiles[2], na.rm = TRUE)
  ) %>%
  ungroup()

# Calculate eGFP expression quartiles by slide
cells_data <- cells_data %>%
  group_by(Slide) %>%
  mutate(eGFP_quartile = ifelse(
    eGFP_scaled < quantile(eGFP_scaled, probs = 0.25, na.rm = TRUE),
    "Lower",
    ifelse(
      eGFP_scaled > quantile(eGFP_scaled, probs = 0.75, na.rm = TRUE),
      "Upper",
      "Middle"
    )
  ))

# Process membranes data
# Calculate 95th percentiles to subtract background
membranes_data_neg_summary <- membranes_data %>%
  filter(Negative_control == TRUE) %>%
  group_by(Sample) %>%
  summarise(
    Ecadherin_95_quant = quantile(Intensity_MeanIntensity_Ecadherin, probs = 0.95, na.rm = TRUE),
    eGFP_95_quant = quantile(Intensity_MeanIntensity_eGFP, probs = 0.95, na.rm = TRUE)
  )

# Subtract background - using SCGB1A1/MUC5AC for SCGB1A1/p63; the same confocal settings were used
membranes_data <- merge(membranes_data, membranes_data_neg_summary, by = "Sample") %>%
  mutate(
    Ecadherin_corrected = Intensity_MeanIntensity_Ecadherin - Ecadherin_95_quant,
    eGFP_corrected = Intensity_MeanIntensity_eGFP - eGFP_95_quant
  )

# Scale 95th and 5th percentiles from 0 to 1 by slide
membranes_data <- membranes_data %>%
  group_by(Slide) %>%
  mutate(
    Ecadherin_scaled = Ecadherin_corrected - quantile(Ecadherin_corrected, probs = scaling_percentiles[1], na.rm = TRUE),
    eGFP_scaled = eGFP_corrected - quantile(eGFP_corrected, probs = scaling_percentiles[1], na.rm = TRUE)
  ) %>%
  mutate(
    Ecadherin_scaled = Ecadherin_scaled / quantile(Ecadherin_scaled, probs = scaling_percentiles[2], na.rm = TRUE),
    eGFP_scaled = eGFP_scaled / quantile(eGFP_scaled, probs = scaling_percentiles[2], na.rm = TRUE)
  ) %>%
  ungroup()

# Calculate eGFP expression quartiles by slide
membranes_data <- membranes_data %>%
  group_by(Slide) %>%
  mutate(eGFP_quartile = ifelse(
    eGFP_scaled < quantile(eGFP_scaled, probs = 0.25, na.rm = TRUE),
    "Lower",
    ifelse(
      eGFP_scaled > quantile(eGFP_scaled, probs = 0.75, na.rm = TRUE),
      "Upper",
      "Middle"
    )
  ))
```

## Variation and normalisation
Combine FOXJ1 datasets and plot densities to demonstrate how taking ratio and subtracting background reduces some patient-patient variation
```{r Ratio effects}
nuclei_data %>%
  filter(Negative_control == FALSE & Experiment == "p53 FOXJ1") %>% # Remove negative control results
  pivot_longer(
    cols = c(Intensity_MeanIntensity_FOXJ1, FOXJ1_scaled),
    values_to = "Intensity",
    names_to = "Correction"
  ) %>%
  mutate(Correction = recode(Correction, Intensity_MeanIntensity_FOXJ1 = "Mean intensity", FOXJ1_scaled = "Normalised intensity")) %>%
  ggplot(aes(x = Intensity, fill = Sample)) +
  geom_density(alpha = 0.5) +
  facet_grid2(Correction ~ Condition, scales = "free", independent = "all") +
  labs(
    title = "Nuclear FOXJ1 intensities",
    x = "Nuclear FOXJ1 intensity or ratio",
    y = NULL
  ) +
  theme_bw() +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  ) +
  scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "Normalising effects.png"), width = plot_width, height = plot_height)

# Plot of nuclear DRAQ5 intensities
# Put some data together
bind_rows(
  merge(read.csv(paste0(input_dir, Ecad_FOXJ1_nuclei_file)), annotations),
  merge(read.csv(paste0(input_dir, p53_FOXJ1_nuclei_file)), annotations),
  merge(read.csv(paste0(input_dir, p53_p63_nuclei_file)), annotations),
  merge(read.csv(paste0(input_dir, SCGB1A1_p63_nuclei_file)), annotations),
  merge(read.csv(paste0(input_dir, SCGB1A1_MUC5AC_nuclei_file)), annotations)
) %>% # Select required columns
  dplyr::select(
    Sample, Donor, Condition, Negative_control,
    Intensity_MeanIntensity_DRAQ5, Experiment
  ) %>%
  ggplot(aes(x = Intensity_MeanIntensity_DRAQ5, fill = Sample)) +
  geom_density(alpha = 0.5) +
  facet_grid(Negative_control ~ Experiment) +
  labs(
    title = "Nuclear DRAQ5 intensities",
    x = "Mean nuclear DRAQ5 intensity",
    y = NULL
  ) +
  theme_bw() +
  theme(
    axis.ticks.y = element_blank(),
    axis.text.y = element_blank()
  ) +
  scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "DRAQ5 density.png"), width = plot_width, height = plot_height)
```

## p63 proportions
Plot of combined p63 slides
```{r p63 intensities}
# Make summaries
p63_summary <- nuclei_data %>%
  filter(Negative_control == FALSE) %>%
  group_by(Donor, Condition) %>%
  summarize(
    p63_mean = mean(p63_scaled, na.rm = TRUE),
    p63_prop = mean(p63_scaled > p63_threshold, na.rm = TRUE),
    .groups = "drop"
  )

# Perform paired t-tests with proportion of positive cells
p63_test_prop <- t.test(
  filter(p63_summary, Condition == "Control")$p63_prop,
  filter(p63_summary, Condition == "p53 KD")$p63_prop,
  paired = TRUE
)
interpret_t_test(p63_test_prop, p63_summary, "p63_prop")

# Print range of p63 positive proportion
range(p63_summary$p63_prop)

# Density plot of nuclear p63 intensity
nuclei_data %>%
  filter(Negative_control == FALSE) %>%
  ggplot(aes(x = p63_scaled, fill = Sample)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = p63_threshold, colour = "red") +
  facet_wrap("Condition") +
  labs(
    title = "Nuclear p63 intensity distribution",
    x = "Normalised nuclear p63 intensity",
    y = NULL
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "p63 intensity distribution.png"), width = plot_width, height = plot_height)

# Bar plot of proportion p63 positive nuclei
nuclei_data %>%
  filter(Negative_control == FALSE) %>%
  group_by(Donor, Condition, Sample) %>%
  summarise(
    `Proportion positive` = mean(p63_scaled > p63_threshold, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = Condition, y = `Proportion positive`, fill = Sample)) +
  geom_col() +
  geom_hline(yintercept = basal_prop, col = "red") +
  facet_wrap("Donor") +
  scale_y_continuous(
    labels = function(x) paste0(x * 100, "%")
  ) +
  labs(
    title = "Proportion of p63 positive nuclei",
    subtitle = paste0(
      "p-value: ",
      round(p63_test_prop$p.value, digits = 3)
    ),
    x = NULL,
    y = "p63 positive nuclei"
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "p63 proportion positive nuclei.png"), width = plot_width, height = plot_height)
```

## p53 p63
Interesting observation of mutually exclusive p53/p63 expression - supports idea of p53 driving differentiation.
```{r p53/p63 intensities}
# Scatter plot of p53 vs p63 intensities
nuclei_data %>%
  filter(Negative_control == FALSE) %>%
  ggplot(aes(x = p53_scaled, y = p63_scaled, colour = Sample)) +
  geom_point(alpha = 0.3) +
  facet_grid(Condition ~ Donor) +
  labs(
    title = "Nuclear p53/p63 intensities",
    x = "Normalised nuclear p53 intensity",
    y = "Normalised nuclear p63 intensity"
  ) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "p53 p63 nuclear intensities.png"), width = plot_width, height = plot_height)
```

## SCGB1A1 expression
```{r SCGB1A1}
# Density plot of SCGB1A1 intensities
cells_data %>%
  filter(Negative_control == FALSE) %>%
  ggplot(aes(x = SCGB1A1_scaled, fill = Sample)) +
  geom_density(alpha = 0.5) +
  facet_wrap("Condition") +
  labs(
    title = "Cellular SCGB1A1 intensities",
    x = "Normalised cellular SCGB1A1 intensity",
    y = NULL
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "SCGB1A1 intensity distribution.png"), width = plot_width, height = plot_height)
```

## SCGB1A1 quartile comparison
```{r}
# Summarise the quartiles
SCGB1A1_quartiles <- cells_data %>%
    filter(Negative_control == FALSE & eGFP_quartile != "Middle") %>%
    group_by(Donor, Condition, eGFP_quartile) %>%
    summarise(
        SCGB1A1_scaled = mean(SCGB1A1_scaled, na.rm = TRUE)
    ) %>%
    pivot_wider(names_from = eGFP_quartile, values_from = SCGB1A1_scaled) %>%
    mutate(Ratio = Upper / Lower)

# Run paired t-test
SCGB1A1_ratio_t_result <- t.test(
    filter(SCGB1A1_quartiles, Condition == "Control")$Ratio,
    filter(SCGB1A1_quartiles, Condition == "p53 KD")$Ratio,
    paired = TRUE
)
SCGB1A1_ratio_p_label <- paste0("p = ", signif(SCGB1A1_ratio_t_result$p.value, 3))

# Plot quartile ratios
SCGB1A1_quartiles %>%
    ggplot(aes(x = Condition, y = Ratio, col = Donor, group = Donor)) +
    geom_point(size = 3) +
    geom_line(aes(x = as.numeric(Condition), y = Ratio)) +
    annotate(
        "text",
        x = 1.5,
        y = max(SCGB1A1_quartiles$Ratio) * 1.05,
        label = SCGB1A1_ratio_p_label,
        size = 4
    ) +
    labs(
        title = "SCGB1A1 expression differences",
        x = "Experimental condition",
        y = "Ratio of SCGB1A1 intensity by eGFP quartile"
    ) +
    theme_bw() +
    scale_color_manual(name = "Donor", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "SCGB1A1_quartile_ratios.png"), width = plot_width, height = plot_height)
```

## MUC5AC proportions
```{r MUC5AC intensities}
# Make summaries
MUC5AC_summary <- cells_data %>%
    filter(Negative_control == FALSE) %>%
    group_by(Donor, Condition) %>%
    summarize(
        MUC5AC_mean = mean(MUC5AC_scaled, na.rm = TRUE),
        MUC5AC_prop = mean(MUC5AC_scaled > MUC5AC_threshold, na.rm = TRUE),
        .groups = "drop"
    )

# Perform paired t-tests and print results for corrected ratio values
MUC5AC_test_mean <- t.test(
    filter(MUC5AC_summary, Condition == "p53 KD")$MUC5AC_mean,
    filter(MUC5AC_summary, Condition == "Control")$MUC5AC_mean,
    paired = TRUE
)
interpret_t_test(MUC5AC_test_mean, MUC5AC_summary, "MUC5AC_mean")

# Repeat but with proportion of positive cells rather than corrected ratios
MUC5AC_test_prop <- t.test(
    filter(MUC5AC_summary, Condition == "Control")$MUC5AC_prop,
    filter(MUC5AC_summary, Condition == "p53 KD")$MUC5AC_prop,
    paired = TRUE
)
interpret_t_test(MUC5AC_test_prop, MUC5AC_summary, "MUC5AC_prop")

# Print range of MUC5AC positive proportion
range(MUC5AC_summary$MUC5AC_prop)

# Density plot of cellular MUC5AC intensity
cells_data %>%
    filter(Negative_control == FALSE) %>%
    ggplot(aes(x = MUC5AC_scaled, fill = Sample)) +
    geom_density(alpha = 0.5) +
    geom_vline(xintercept = MUC5AC_threshold, colour = "red") +
    facet_wrap("Condition") +
    labs(
        title = "Cellular MUC5AC intensity distribution",
        x = "Normalised cellular MUC5AC intensity",
        y = NULL
    ) +
    theme_bw() +
    scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "MUC5AC intensity distribution.png"), width = plot_width, height = plot_height)

# Bar plot of proportion MUC5AC positive cells
cells_data %>%
    filter(Negative_control == FALSE) %>%
    group_by(Donor, Condition, Sample) %>%
    summarise(
        `Proportion positive` = mean(MUC5AC_scaled > MUC5AC_threshold, na.rm = TRUE)
    ) %>%
    ggplot(aes(x = Condition, y = `Proportion positive`, fill = Sample)) +
    geom_col() +
    geom_hline(yintercept = goblet_prop, col = "red") +
    facet_wrap("Donor") +
    scale_y_continuous(
        labels = function(x) paste0(x * 100, "%")
    ) +
    labs(
        title = "Proportion of MUC5AC positive cells",
        subtitle = paste0(
            "p-value: ",
            round(MUC5AC_test_prop$p.value, digits = 3)
        ),
        x = NULL,
        y = "MUC5AC positive cells"
    ) +
    theme_bw() +
    scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "MUC5AC proportion positive cells.png"), width = plot_width, height = plot_height)
```

## FOXJ1 proportions
Plot of combined FOXJ1 slides
```{r FOXJ1 intensities}
# Make summaries
FOXJ1_summary <- nuclei_data %>%
    filter(Negative_control == FALSE) %>%
    group_by(Donor, Condition) %>%
    summarize(
        FOXJ1_mean = mean(FOXJ1_scaled, na.rm = TRUE),
        FOXJ1_prop = mean(FOXJ1_scaled > FOXJ1_threshold, na.rm = TRUE),
        .groups = "drop"
    )

# Perform paired t-tests with proportion of positive cells
FOXJ1_test_prop <- t.test(
    filter(FOXJ1_summary, Condition == "Control")$FOXJ1_prop,
    filter(FOXJ1_summary, Condition == "p53 KD")$FOXJ1_prop,
    paired = TRUE
)
interpret_t_test(FOXJ1_test_prop, FOXJ1_summary, "FOXJ1_prop")

# Print range of FOXJ1 positive proportion
range(FOXJ1_summary$FOXJ1_prop)

# Density plot of nuclear FOXJ1 intensity
nuclei_data %>%
    filter(Negative_control == FALSE) %>%
    ggplot(aes(x = FOXJ1_scaled, fill = Sample)) +
    geom_density(alpha = 0.5) +
    geom_vline(xintercept = FOXJ1_threshold, colour = "red") +
    facet_wrap("Condition") +
    labs(
        title = "Nuclear FOXJ1 intensity distribution",
        x = "Normalised nuclear FOXJ1 intensity",
        y = NULL
    ) +
    theme_bw() +
    scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "FOXJ1 intensity distribution.png"), width = plot_width, height = plot_height)

# Bar plot of proportion FOXJ1 positive nuclei
nuclei_data %>%
    filter(Negative_control == FALSE) %>%
    group_by(Donor, Condition, Sample) %>%
    summarise(
        `Proportion positive` = mean(FOXJ1_scaled > FOXJ1_threshold, na.rm = TRUE)
    ) %>%
    ggplot(aes(x = Condition, y = `Proportion positive`, fill = Sample)) +
    geom_col() +
    geom_hline(yintercept = ciliated_prop, col = "red") +
    facet_wrap("Donor") +
    scale_y_continuous(
        labels = function(x) paste0(x * 100, "%")
    ) +
    labs(
        title = "Proportion of FOXJ1 positive nuclei",
        subtitle = paste0(
            "p-value: ",
            round(FOXJ1_test_prop$p.value, digits = 3)
        ),
        x = NULL,
        y = "FOXJ1 positive nuclei"
    ) +
    theme_bw() +
    scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "FOXJ1 proportion positive nuclei.png"), width = plot_width, height = plot_height)
```

## Ecadherin expression
```{r Ecadherin}
# Density plot of Ecadherin intensities
membranes_data %>%
    filter(Negative_control == FALSE) %>%
    ggplot(aes(x = Ecadherin_scaled, fill = Sample)) +
    geom_density(alpha = 0.5) +
    facet_wrap("Condition") +
    labs(
        title = "Membranous E-cadherin intensities",
        x = "Normalised membranous E-cadherin intensity",
        y = NULL
    ) +
    theme_bw() +
    theme(
        axis.text.y = element_blank(),
        axis.ticks.y = element_blank()
    ) +
    scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "Ecadherin intensity distribution.png"), width = plot_width, height = plot_height)
```

## Ecadherin quartile comparison
```{r}
# Summarise the quartiles
Ecadherin_quartiles <- membranes_data %>%
    filter(Negative_control == FALSE & eGFP_quartile != "Middle") %>%
    group_by(Donor, Condition, eGFP_quartile) %>%
    summarise(
        Ecadherin_scaled = mean(Ecadherin_scaled, na.rm = TRUE)
    ) %>%
    pivot_wider(names_from = eGFP_quartile, values_from = Ecadherin_scaled) %>%
    mutate(Ratio = Upper / Lower)

# Run paired t-test
Ecadherin_ratio_t_result <- t.test(
    filter(Ecadherin_quartiles, Condition == "Control")$Ratio,
    filter(Ecadherin_quartiles, Condition == "p53 KD")$Ratio,
    paired = TRUE
)
Ecadherin_ratio_p_label <- paste0("p = ", signif(Ecadherin_ratio_t_result$p.value, 3))

# Plot quartile ratios
Ecadherin_quartiles %>%
    ggplot(aes(x = Condition, y = Ratio, col = Donor, group = Donor)) +
    geom_point(size = 3) +
    geom_line(aes(x = as.numeric(Condition), y = Ratio)) +
    annotate(
        "text",
        x = 1.5,
        y = max(Ecadherin_quartiles$Ratio) * 1.05,
        label = Ecadherin_ratio_p_label,
        size = 4
    ) +
    labs(
        title = "E-cadherin expression differences",
        x = "Experimental condition",
        y = "Ratio of E-cadherin intensity by eGFP quartile"
    ) +
    theme_bw() +
    scale_color_manual(name = "Donor", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "Ecadherin_quartile_ratios.png"), width = plot_width, height = plot_height)
```

## Epithelial density
Overall density as measured by nuclear counts per field of view
```{r epithelial density}
# Make summaries
density_data <- nuclei_data %>%
  group_by(Sample, Donor, Condition, Experiment) %>%
  summarise(nuclei_count = n()) %>%
  mutate(nuclei_count = nuclei_count/image_area)

density_summary <- density_data %>%
  group_by(Donor, Condition) %>%
  summarize(nuclei_count_mean = mean(nuclei_count), .groups = "drop")

# Perform paired t-tests and print results
density_test <- t.test(
  filter(density_summary, Condition == "Control")$nuclei_count_mean,
  filter(density_summary, Condition == "p53 KD")$nuclei_count_mean,
  paired = TRUE
)
interpret_t_test(density_test, density_summary, "nuclei_count_mean")

# Make a box-plot
density_data %>%
  ggplot(aes(x = Donor, y = nuclei_count, fill = Sample)) +
  geom_boxplot() +
  labs(
    title = "Cell density",
    subtitle = paste0("p-value = ", round(density_test$p.value, digits = 3)),
    x = "Donor",
    y = expression("Nuclei count per mm"^2)
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "Epithelial_density_boxplot.png"), width = plot_width, height = plot_height)
```

# OLD CODE

## Cell count totals
Add up the proportions from the other cell type counts.
```{r}
# Merge the datasets together
overall_summary <- merge(
  FOXJ1_summary,
  p63_summary,
  by = c("Donor", "Condition")
) %>%
  merge(
    MUC5AC_summary,
    by = c("Donor", "Condition")
  ) %>%
  mutate(
    total_prop = (FOXJ1_prop + p63_prop + MUC5AC_prop)
  )

# Run some stats
total_test_prop <- t.test(
  filter(overall_summary, Condition == "Control")$total_prop,
  filter(overall_summary, Condition == "p53 KD")$total_prop,
  paired = TRUE
)
interpret_t_test(total_test_prop, overall_summary, "total_prop")
```

## p53
Plot of combined p53 slides
```{r p53 intensities}
# Run some stats
# Make summaries
p53_summary <- nuclei_data %>%
  filter(Negative_control == FALSE) %>%
  group_by(Donor, Condition) %>%
  summarize(
    p53_mean = mean(p53_scaled, na.rm = TRUE),
    p53_prop = mean(p53_scaled > p53_threshold, na.rm = TRUE),
    .groups = "drop"
  )

# Perform paired t-tests and print results for corrected ratio values
p53_test_mean <- t.test(
  filter(p53_summary, Condition == "Control")$p53_mean,
  filter(p53_summary, Condition == "p53 KD")$p53_mean,
  paired = TRUE
)
interpret_t_test(p53_test_mean, p53_summary, "p53_mean")

# Repeat but with proportion of positive cells rather than corrected ratios
p53_test_prop <- t.test(
  filter(p53_summary, Condition == "Control")$p53_prop,
  filter(p53_summary, Condition == "p53 KD")$p53_prop,
  paired = TRUE
)
interpret_t_test(p53_test_prop, p53_summary, "p53_prop")

# Density plot of nuclear FOXJ1 intensity
nuclei_data %>%
  filter(Negative_control == FALSE) %>% # Remove negative control results
  ggplot(aes(x = p53_scaled, fill = Sample)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = p53_threshold, colour = "red") +
  facet_wrap("Condition") +
  labs(
    title = "Nuclear p53 intensities",
    subtitle = paste0(
      "Mean difference: ",
      round(p53_test_mean$estimate, digits = 3),
      ", p-value: ",
      round(p53_test_mean$p.value, digits = 3)
    ),
    x = "Nuclear p53 intensity relative to DRAQ5",
    y = NULL
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "p53 relative nuclear intensities.png"), width = plot_width, height = plot_height)

# Bar plot of proportion FOXJ1 positive nuclei
nuclei_data %>%
  group_by(Donor, Condition, Sample) %>%
  summarise(
    `Proportion positive` = mean(p53_ratio_corrected > p53_threshold, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = Condition, y = `Proportion positive`, fill = Sample)) +
  geom_col() +
  facet_wrap("Donor") +
  labs(
    title = "Proportion of p53 positive nuclei",
    subtitle = paste0(
      "Mean difference: ",
      round(p53_test_prop$estimate, digits = 3),
      ", p-value: ",
      round(p53_test_prop$p.value, digits = 3)
    ),
    x = NULL,
    y = "p53 positive nuclei"
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "p53 proportion positive nuclei.png"), width = plot_width, height = plot_height)
```

## eGFP
Plot of combined eGFP slides
```{r eGFP intensities}
# Run some stats
# Make summaries
eGFP_summary <- nuclei_data %>%
  group_by(Donor, Condition) %>%
  summarize(
    eGFP_mean = mean(eGFP_ratio_corrected, na.rm = TRUE),
    eGFP_prop = mean(eGFP_ratio_corrected > eGFP_threshold, na.rm = TRUE),
    .groups = "drop"
  )

# Perform paired t-tests and print results for corrected ratio values
eGFP_test_mean <- t.test(
  filter(eGFP_summary, Condition == "Control")$eGFP_mean,
  filter(eGFP_summary, Condition == "p53 KD")$eGFP_mean,
  paired = TRUE
)
interpret_t_test(eGFP_test_mean, eGFP_summary, "eGFP_mean")

# Repeat but with proportion of positive cells rather than corrected ratios
eGFP_test_prop <- t.test(
  filter(eGFP_summary, Condition == "Control")$eGFP_prop,
  filter(eGFP_summary, Condition == "p53 KD")$eGFP_prop,
  paired = TRUE
)
interpret_t_test(eGFP_test_prop, eGFP_summary, "eGFP_prop")

# Density plot of nuclear FOXJ1 intensity
nuclei_data %>%
  filter(Negative_control == FALSE) %>% # Remove negative control results
  ggplot(aes(x = eGFP_ratio_corrected, fill = Sample)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = eGFP_threshold, colour = "red") +
  facet_wrap("Condition") +
  labs(
    title = "Nuclear eGFP intensities",
    subtitle = paste0(
      "Mean difference: ",
      round(eGFP_test_prop$estimate, digits = 3),
      ", p-value: ",
      round(eGFP_test_prop$p.value, digits = 3)
    ),
    x = "Nuclear eGFP intensity relative to DRAQ5",
    y = NULL
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "eGFP relative nuclear intensities.png"), width = plot_width, height = plot_height)

# Bar plot of proportion FOXJ1 positive nuclei
nuclei_data %>%
  group_by(Donor, Condition, Sample) %>%
  summarise(
    `Proportion positive` = mean(eGFP_ratio_corrected > eGFP_threshold, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = Condition, y = `Proportion positive`, fill = Sample)) +
  geom_col() +
  facet_wrap("Donor") +
  labs(
    title = "Proportion of eGFP positive nuclei",
    subtitle = paste0(
      "Mean difference: ",
      round(eGFP_test_prop$estimate, digits = 3),
      ", p-value: ",
      round(eGFP_test_prop$p.value, digits = 3)
    ),
    x = NULL,
    y = "eGFP positive nuclei"
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "eGFP proportion positive nuclei.png"), width = plot_width, height = plot_height)
```

## p53 eGFP
Scatter plot of p53 vs eGFP expression
```{r p53 eGFP}
# Scatter plot of nuclear p53 vs nuclear eGFP intensities.
nuclei_data %>%
  filter(Negative_control == FALSE) %>% # Remove negative control results
  ggplot(aes(x = p53_ratio_corrected, y = eGFP_ratio_corrected, colour = Sample)) +
  geom_point(alpha = 0.3) +
  facet_grid(Condition ~ Donor) +
  labs(
    title = "Mean nuclear p53/eGFP intensities",
    x = "Corrected nuclear p53 ratio relative to DRAQ5",
    y = "Corrected nuclear eGFP ratio relative to DRAQ5"
  ) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "p53 eGFP nuclear intensities.png"), width = plot_width, height = plot_height)
```

## SCGB1A1 p63
SCGB1A1 also seems to be somewhat mutually exclusive, demonstrating distiniction of club cell population from basal cell population. The loading of this data may need to be edited to compare nuclear p63 with cellular SCGB1A1 of cytoplasmic SCGB1A1.
```{r SCGB1A1/p63 intensities}
# Scatter plot of nuclear p63 vs nuclear SCGB1A1 intensities.
bind_cols(
  filter(nuclei_data, Experiment == "SCGB1A1 p63"),
  SCGB1A1_corrected = filter(cells_data, Experiment == "SCGB1A1 p63")$SCGB1A1_corrected
) %>%
  ggplot(aes(x = p63_ratio_corrected, y = SCGB1A1_corrected, colour = Sample)) +
  geom_point(alpha = 0.3) +
  facet_grid(Condition ~ Donor) +
  labs(
    title = "Mean nuclear SCGB1A1/p63 intensities",
    x = "Nuclear p63 intensity relative to DRAQ5",
    y = "Nuclear SCGB1A1 intensity relative to DRAQ5"
  ) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "SCGB1A1 p63 nuclear intensities.png"), width = plot_width, height = plot_height)
```

## SCGB1A1 MUC5AC
```{r SCGB1A1 MUC5AC}
# Plot of cellular SCGB1A1 and MUC5AC intensity
cells_data %>% # normalise against DRAQ5 intensity
  ggplot(aes(x = SCGB1A1_corrected, y = MUC5AC_corrected, colour = Sample)) +
  geom_point(alpha = 0.3) +
  facet_grid(Condition ~ Donor) +
  labs(
    title = "Mean cellular SCGB1A1/MUC5AC intensities",
    x = "Mean cellular SCGB1A1 intensity",
    y = "Mean cellular MUC5AC intensity"
  ) +
  theme_bw() +
  scale_colour_manual(name = "Sample", values = custom_colours) +
  theme(
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "SCGB1A1 MUC5AC cellular intensities.png"), width = plot_width, height = plot_height)
```

# Quartile comparisons
## p63 quartile comparison
```{r}
# Summarise the quartiles
p63_quartiles <- nuclei_data %>%
  filter(Negative_control == FALSE & eGFP_quartile != "Middle") %>%
  group_by(Donor, Condition, eGFP_quartile) %>%
  summarise(
    p63_scaled = mean(p63_scaled, na.rm = TRUE)
  ) %>%
  pivot_wider(names_from = eGFP_quartile, values_from = p63_scaled) %>%
  mutate(Ratio = Upper / Lower)

# Run paired t-test
FOXJ1_ratio_t_result <- t.test(
  filter(FOXJ1_quartiles, Condition == "Control")$Ratio,
  filter(FOXJ1_quartiles, Condition == "p53 KD")$Ratio,
  paired = TRUE
)
FOXJ1_ratio_p_label <- paste0("p = ", signif(FOXJ1_ratio_t_result$p.value, 3))

# Plot quartile ratios
FOXJ1_quartiles %>%
  ggplot(aes(x = Condition, y = Ratio, col = Donor, group = Donor)) +
  geom_point(size = 3) +
  geom_line(aes(x = as.numeric(Condition), y = Ratio)) +
  annotate(
    "text",
    x = 1.5,
    y = max(FOXJ1_quartiles$Ratio) * 1.05,
    label = FOXJ1_ratio_p_label,
    size = 4
  ) +
  labs(
    title = "FOXJ1 expression differences",
    x = "Experimental condition",
    y = "Ratio of FOXJ1 intensity by eGFP quartile"
  ) +
  theme_bw() +
  scale_color_manual(name = "Donor", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "FOXJ1_quartile_ratios.png"), width = plot_width, height = plot_height)
```
