---
title: "TEER visualisation"
author: "Sam Siljee"
date: '2022-06-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, include = FALSE}
library("dplyr")
library("ggplot2")
library("tidyr")
library("rstatix") # To run ANOVA and t-test
library("ggpubr") # Easy QQ plot

# Files and directories
input_dir <- paste0(getwd(), "/data/")
output_dir <- paste0(getwd(), "/../produced_plots/TEER/")
final_data_file <- "final_TEER_data.tsv"
annotations_file <- "annotations.tsv"
colours_file <- "colours.csv"
phenotype_summary_path <- paste0(input_dir, "../../Misc/data/phenotype_summary.csv")
```


## Load the data
```{r Load the data}
# Read in annotations
annotations <- read.table(paste0(input_dir, annotations_file), sep = "\t", header = TRUE, stringsAsFactors = TRUE) %>% mutate(Sample = paste(Donor, Condition, sep = "_"))

# Custom colour scheme for plotting
custom_colours <- read.csv(paste0(input_dir, "../../", colours_file))$Colour
names(custom_colours) <- read.csv(paste0(input_dir, "../../", colours_file))$Name

# Reading in the TEER readings
TEER_data <- read.table(paste0(input_dir, final_data_file), sep = "\t", header = TRUE, stringsAsFactors = TRUE)

# Merge with annotations
TEER_data <- merge(
  TEER_data,
  annotations
)

# Summarise the replicates
dat_summarised <- TEER_data %>%
  group_by(Donor, Condition, DayPostAirlift) %>%
  summarize(SD = sd(TEER), TEER = mean(TEER)) %>%
  as.data.frame()
```
# Add to phenotype data
Summarise and add to the phenotype data
```{r}
# Summarise, merge with existing data, and write to .csv for the phenotype summary
TEER_data %>%
  pivot_wider(names_from = DayPostAirlift, values_from = TEER, names_glue = "{.value}_{DayPostAirlift}") %>%
  select(- c(Donor, Well, Condition, Sample)) %>%
  merge(
    select(
      read.csv(phenotype_summary_path),
      -c(
        TEER_7,
        TEER_14,
        TEER_21,
        TEER_28
      )
    ),
    by = "Index"
  ) %>%
  write.csv(phenotype_summary_path, row.names = FALSE)
```

# Summarise
Make some summaries for use in thesis
```{r summaries}
# Mean values by condition and by day
paste("p53 KD mean:", round(mean(TEER_data$TEER[TEER_data$Condition == "pLVUHshp53"])))
paste("Control mean:", round(mean(TEER_data$TEER[TEER_data$Condition == "pLL3.7"])))
paste("Day 7 mean:", round(mean(TEER_data$TEER[TEER_data$DayPostAirlift == 7])))
paste("Day 14 mean:", round(mean(TEER_data$TEER[TEER_data$DayPostAirlift == 14])))
paste("Day 21 mean:", round(mean(TEER_data$TEER[TEER_data$DayPostAirlift == 21])))
paste("Day 28 mean:", round(mean(TEER_data$TEER[TEER_data$DayPostAirlift == 28])))

# Get measure of inter-patient heterogeneity
SD_measures <- c()
for(d in unique(TEER_data$DayPostAirlift)) {
  for(c in c("pLVUHshp53", "pLL3.7")) {
    SD_measures <- paste0(
      SD_measures,
      paste0(
        ", day ",
        d,
        ifelse(c == "pLL3.7", " Control SD ", " p53 KD SD "),
        round(sd(TEER_data$TEER[TEER_data$DayPostAirlift == d & TEER_data$Condition == c]))
      )
    )
  }
}
SD_measures
```

# Stats
In the publication Chiu et. al. the Two-tailed unpaired Student’s t-test method is used for TEER data.
Chiu, M. C., Li, C., Liu, X., Yu, Y., Huang, J., Wan, Z., Xiao, D., Chu, H., Cai, J.-P., Zhou, B., Sit, K.-Y., Au, W.-K., Wong, K. K.-Y., Li, G., Chan, J. F.-W., To, K. K.-W., Chen, Z., Jiang, S., Clevers, H., … Zhou, J. (2022). A bipotential organoid model of respiratory epithelium recapitulates high infectivity of SARS-CoV-2 Omicron variant. Cell Discovery, 8(1), 57. https://doi.org/10.1038/s41421-022-00422-1
However, ANOVA also seems to be commonly used.
Discussed with Clint, suggested repeated measures ANOVA as the most robust method, with paired t-test for individual time-points.
Using a mixed three-way anova with one within subjects factor (wells) and two between subjects factors (condition and timepoint) is a way to account for replicates.
Methods followed:
https://www.youtube.com/watch?v=wL0MVlusawY
https://www.datanovia.com/en/lessons/repeated-measures-anova-in-r/
https://www.datanovia.com/en/lessons/mixed-anova-in-r/

## Outliers
The identify outliers function from the `rstatix` package has identified two outliers; these are 
21-161 pLL3.7,	day: 7, well: 2 (TEER 829)
21-073 pLVUHshp53, day: 14, well: 6 (TEER 528)
None of these are extreme outliers.

```{r Outliers}
# First check for outliers
TEER_data %>%
  group_by(Condition, DayPostAirlift) %>%
  identify_outliers(TEER)
```

## Normality
The Shapiro-Wilks test was used to check normal distribution of the TEER values.
Originally one set of TEER values was identified with p > 0.05, namely pLL3.7 controls at day 7.
I believe this is due to the outlyign data point, as the QQ plot appears to fall largely within the predicted values - save for the highest point. With the addition of more data however this is no longer significant.
Normality can also be checked by QQ plot

```{r Normality}
# Use Shapiro-Wilks test to check normal distribution of TEER
TEER_data %>%
  group_by(Condition, DayPostAirlift) %>%
  shapiro_test(TEER)

# QQ plot to check normality
TEER_data %>%
  ggqqplot("TEER", ggtheme = theme_bw()) + facet_grid(DayPostAirlift ~ Condition)
```

## Run ANOVA
Run the two-way repeated measures ANOVA.
I'm not yet sure how to handle replicate measures in the two-way ANOVA analysis, so for now I'm summarising the replicates using `mean`.
I will look into using a different ANOVA method to deal with the replicates.
```{r ANOVA}
# Base R ANOVA test
# anova(lm(TEER ~ Condition * DayPostAirlift, data = TEER_data))

# Run the ANOVA
anova_res <- anova_test(
  data = dat_summarised,
  dv = TEER,
  wid = Donor,
  within = c(Condition, DayPostAirlift)
)
anova_res
```

## Sphericity
After the ANOVA has been run, sphericity can be checked.
```{r sphericity}
```

## Post-hoc tests
After significant two-way interactions have been identified, simple main effect and simple pairwise comparisons can be run
```{r post-hoc by condition}
# Run simple main effect one-way ANOVA
one_way_cond_res <- dat_summarised %>%
  group_by(DayPostAirlift) %>%
  anova_test(dv = TEER, wid = Donor, within = Condition) %>%
  get_anova_table() %>%
  adjust_pvalue(method = "bonferroni")
one_way_cond_res

# Run pairwise comparisons
pairwise_cond_res <- TEER_data %>%
  group_by(DayPostAirlift) %>%
  pairwise_t_test(
    TEER ~ Condition,
    paired = TRUE,
    p.adjust.method = "bonferroni"
  )
pairwise_cond_res
```

```{r post-hoc by airlift day}
# Run simple main effect one-way ANOVA
one_way_day_res <- dat_summarised %>%
  group_by(Condition) %>%
  anova_test(dv = TEER, wid = Donor, within = DayPostAirlift) %>%
  get_anova_table() %>%
  adjust_pvalue(method = "bonferroni")
one_way_day_res

# Run pairwise comparisons
pairwise_day_res <- TEER_data %>%
  group_by(Condition) %>%
  pairwise_t_test(
    TEER ~ DayPostAirlift,
    paired = TRUE,
    p.adjust.method = "bonferroni"
  )
pairwise_day_res
```

```{r combined plot}
# Creating a combined plot
dat_summarised %>%
  ggplot(aes(
    x = DayPostAirlift,
    y = TEER,
    ymin = TEER - SD,
    ymax = TEER + SD,
    colour = Donor,
    shape = Condition
  )) +
  geom_point(size = 3) +
  geom_errorbar(alpha = 0.5, width = 0.2) +
  geom_line() +
  scale_x_continuous(breaks = seq(0, 28, by = 7), limits = c(0, 28)) + # Adjust X scale
  ylim(0, 800) +
  labs(
    x = "Day post airlift",
    y = expression("TEER (" ~ Omega * cm^2 ~ ")"),
    title = "TEER values by donor - triplicate membranes"
  ) +
  theme_bw() +
  scale_colour_manual(name = "Donor", values = custom_colours)

# And saving
ggsave("../produced_plots/timeseries/Combined plot.png", width = 8, height = 5.5)
```

# Bar graphs
```{r bar graphs}
# Summarise down to donors
TEER_data %>%
  group_by(Condition, DayPostAirlift) %>%
  summarise(SD = sd(TEER), TEER = mean(TEER)) %>%
  ggplot(aes(
    x = Condition,
    y = TEER,
    ymin = TEER - SD,
    ymax = TEER + SD,
    fill = Condition
  )) +
  geom_col() +
  geom_errorbar(alpha = 0.5, width = 0.2) +
  facet_wrap("DayPostAirlift") +
  labs(x = "Day post airlift", y = expression("TEER (" ~ Omega * cm^2 ~ ")")) +
  ggtitle("TEER values by day post airlift") +
  theme_bw()

# And saving
ggsave("../produced_plots/timeseries/Facet bar plot.png", width = 8, height = 5.5)
```

# Box plot
```{r box plot}
TEER_data %>%
  ggplot(aes(
    x = Condition,
    y = TEER,
    fill = Condition
  )) +
  geom_boxplot() +
  facet_wrap("DayPostAirlift") +
  labs(x = "Day post airlift", y = expression("TEER (" ~ Omega * cm^2 ~ ")")) +
  ggtitle("TEER values by day post airlift") +
  theme_bw()

# And saving
ggsave("../produced_plots/timeseries/Facet boxplot.png", width = 8, height = 5.5)
```

# Paired line chart
Personally I like this one best.
```{r paired line plot}
# Custom labels for facet wrap
cust_labels <- character()
for(day in 7*1:4){
  cust_labels <- c(
    cust_labels,
    paste0("Day ", day, ", p.adj = ", filter(pairwise_cond_res, DayPostAirlift == day)$p.adj)
  )
}
names(cust_labels) <- 7*1:4

# Create chart
dat_summarised %>%
  ggplot(aes(x = Condition, y = TEER, ymax = TEER + SD, ymin = TEER - SD, col = Donor)) +
  geom_boxplot(aes(x = Condition, y = TEER, col = Condition)) +
  geom_point(size = 3) +
  geom_errorbar(alpha = 0.5, width = 0.07) +
  geom_line(aes(x = as.numeric(Condition), y = TEER)) +
  facet_wrap(
    "DayPostAirlift",
    labeller = as_labeller(cust_labels),
    nrow = 1) +
  labs(
    x = "Lentivirus vector",
    y = expression("TEER (" ~ Omega * cm^2 ~ ")"),
    title = "TEER values by day post airlift") +
  theme_bw() +
  theme(strip.background = element_blank()) +
  scale_colour_manual(name = "Donor", values = custom_colours)
  
# Save the plot
ggsave(paste0(output_dir, "Paired line plot.png"), width = 8, height = 5.5)
```

# Filtered plot

```{r filtered plot}
# Creating a combined plot
singlePlot <- dat %>%
  filter(Donor %in% tail(unique(dat$Donor), 8)) %>%
  separate_wider_delim(Donor, delim = " ", names = c("Donor", "Condition")) %>%
  ggplot(
    aes(
      x = DayPostAirlift,
      y = TEER,
      colour = Condition,
      shape = Well
    )
  ) +
  geom_point(
    alpha = 0.5,
    show.legend = TRUE
  ) +
  geom_line(
    alpha = 0.5,
    show.legend = FALSE
  ) +
  facet_grid(rows = vars(Donor)) +
  labs(x = "Day post airlift", y = expression("TEER (" ~ Omega * cm^2 ~ ")")) +
  ggtitle("TEER values - triplicate membranes") +
  theme_bw()

singlePlot

# And saving
ggsave(paste0(output_dir, "filtered plot.png"), width = 8, height = 5.5)
```

# Facet wrap of timepoints
For the individual time points, use a facet wrap to present each time-point in bar-graph form (for thesis, not so much for publication)
