---
title: "CBF_results"
author: "Sam Siljee"
date: "2025-06-12"
output: html_document
---

Script to further process and plot results of the CBF analysis.

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Packages
library(dplyr) # Data manipulation
library(ggplot2) # Plotting
library(png) # writePNG function for detected motion images

# Files and directories
input_dir <- paste0(getwd(), "/data/")
output_dir <- paste0(getwd(), "/../produced_plots/CBF/")
annotations_file <- "annotations.tsv"
colours_file <- "colours.csv"
phenotype_summary_path <- paste0(input_dir, "../../Misc/data/phenotype_summary.csv")

# User variables
lower_threshold <- 3.1
upper_threshold <- 19.7
use_cluster_mask <- TRUE # If false, uses detected_motion instead
```

## Load the data
Loading the detected motion masks and .rda output of the CBF analysis
```{r Load the data}
# Load custom functions and colours
source(paste0(input_dir, "../../Functions.R"))

# Read in annotations
annotations <- read.table(paste0(input_dir, annotations_file), sep = "\t", header = TRUE, stringsAsFactors = TRUE) %>% mutate(Sample = paste(Donor, Condition, sep = "_"))

# Load detected motion
detected_motion <- list()
for (i in annotations$Index) {
  # Read in data
  ifelse(
    use_cluster_mask,
    # Use cluster mask from CBF workflow
    dat <- read.table(paste0(input_dir, i, "_cluster_mask.tsv"), header = FALSE, sep = "\t") %>% as.matrix(),
    # Else use detected_motion from multi-DDM (MATLAB) workflow, and convert to logical
    dat <- ifelse(read.csv(paste0(input_dir, i, "_detected_motion.csv"), header = FALSE) %>% as.matrix() == 1, TRUE, FALSE)
  )
  # Add to list
  detected_motion[[i]] <- dat
}

# Load FFT frequency data, converted, unconverted, and phase
load(paste0(input_dir, "FFT_list_unconverted.rda"))
load(paste0(input_dir, "FFT_list_converted.rda"))
load(paste0(input_dir, "phase_metrics.rda"))
```

## Process data
Mask the frequency data matrix by the motion detected matrix, then extract frequency data to a dataframe.
Summarise the proportion of motile detected by sample.
```{r Process frequency data}
# Initialise dataframe
frequency_data <- data.frame()

# Loop through all fields of view
for (i in annotations$Index) {
  # extract and filter frequency data
  ifelse(
    sum(detected_motion[[i]]) == 0, # Check for mask masking everything out
    frequencies <- 0,
    frequencies <- FFT_list_converted[[i]][detected_motion[[i]]]
  )

  # Add to data frame
  frequency_data <- bind_rows(
    frequency_data,
    data.frame(
      Frequency = frequencies,
      Index = i
    )
  )
}

# Merge with annotations
frequency_data <- merge(frequency_data, annotations)

# Summarise proportions of motion detected
# Initialise blank data frame
motion_detected <- data.frame()

for (i in annotations$Index) {
  # Calculate mean motion detected and add to dataframe
  motion_detected <- bind_rows(
    motion_detected,
    data.frame(
      Index = i,
      Proportion_motile = mean(detected_motion[[i]]),
      Proportion_motile_sd = sd(detected_motion[[i]])
    )
  )
}

# Merge with annotations
motion_detected <- merge(
  motion_detected,
  annotations
)

# Write frequency data to phenotype summary
# Summarise CBF results
CBF_freq_summary <- frequency_data %>%
  group_by(Index) %>%
  summarize(
    CBF_freq_mean = mean(Frequency),
    CBF_freq_sd = sd(Frequency),
    .groups = "drop"
  )

# Merge with existing data, and write to .csv
merge(
  CBF_freq_summary,
  motion_detected,
  by = "Index"
) %>%
  dplyr::select(-c(Donor, Condition, Well, Sample)) %>%
  merge(
    dplyr::select(
      read.csv(phenotype_summary_path),
      -c(
        CBF_freq_mean,
        CBF_freq_sd,
        Proportion_motile,
        Proportion_motile_sd
      )
    ),
    by = "Index"
  ) %>%
  write.csv(phenotype_summary_path, row.names = FALSE)
```

## Stats
Run basic paired t-test
```{r stats}
# Run t-test to compare frequency
freq_summary <- frequency_data %>%
  group_by(Donor, Condition) %>%
  summarize(freq_mean = mean(Frequency), .groups = "drop")

# Perform paired t-test
freq_t_test <- t.test(
   filter(freq_summary, Condition == "pLVUHshp53")$freq_mean,
   filter(freq_summary, Condition == "pLL3.7")$freq_mean
   , paired = TRUE)

# Display results
interpret_t_test(freq_t_test, freq_summary, "freq_mean", c("pLVUHshp53", "pLL3.7"))

# Run t-test to compare proportion detected motion
prop_motile_summary <- motion_detected %>%
  group_by(Donor, Condition) %>%
  summarize(prop_mean = mean(Proportion_motile), .groups = "drop")

# Perform paired t-test
prop_motile_t_test <- t.test(
  filter(prop_motile_summary, Condition == "pLVUHshp53")$prop_mean,
  filter(prop_motile_summary, Condition == "pLL3.7")$prop_mean,
  paired = TRUE)

# Display results
interpret_t_test(prop_motile_t_test, prop_motile_summary, "prop_mean", c("pLVUHshp53", "pLL3.7"))

# Run SD to compare inter-patient heterogeneity
freq_summary %>%
  group_by(Condition) %>%
  summarise(SD = sd(freq_mean))

prop_motile_summary %>%
  group_by(Condition) %>%
  summarise(SD = sd(prop_mean))
```

## Frequency boxplot
Boxplots of pixel dominant frequencies, filtered by cluster mask. Also some stats.
```{r Frequency boxplot}
# Box plot of frequencies by sample
frequency_data %>%
  filter(Frequency > lower_threshold & Frequency < upper_threshold) %>%
  ggplot(aes(x = Donor, y = Frequency, fill = Sample)) +
  geom_boxplot() + # Use arg "outlier.shape = NA" to remove outlier point plotting
  labs(
    #title = "Filtered pixel frequencies (cilia cluster mask)",
    subtitle = paste0(
      "p-value: ",
      formatC(freq_t_test$p.value, format = "e", digits = 2)
    ),
    x = "Donor",
    y = "Frequency (Hz)"
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "Frequency_boxplot.png"), width = 8, height = 5.5)
```

## Paired line plots
Paired line plots - styled after the plot used to display Frequency
```{r Paired line plot}
# Summarise the datapoints
dat_summarised <- frequency_data %>%
  group_by(Donor, Condition) %>%
  summarize(SD = sd(Frequency), Frequency = mean(Frequency)) %>%
  as.data.frame()

# Custom subtitle
subtitle <- paste0("p.value = ", round(freq_t_test$p.value, digits = 3))

# Create chart
dat_summarised %>%
  ggplot(aes(x = Condition, y = Frequency, ymax = Frequency + SD, ymin = Frequency - SD, col = Donor)) +
  geom_boxplot(aes(x = Condition, y = Frequency, col = Condition)) +
  geom_point(size = 3) +
  geom_errorbar(alpha = 0.5, width = 0.07) +
  geom_line(aes(x = as.numeric(Condition), y = Frequency)) +
  labs(
    title = "Dominant FFT frequency values",
    subtitle = subtitle,
    x = "Condition",
    y = "Dominant FFT frequency (Hz)"
  ) +
  theme_bw() +
  scale_colour_manual(name = "Donor", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "CBF paired line plot.png"), width = 8, height = 5.5)
```

## Proportion of motion in field of view
Plot showing proption of field of view positive with motion detected.
```{r Detected motion}
# Make box plot of proportions motile
motion_detected %>%
  ggplot(aes(x = Donor, y = Proportion_motile, fill = Sample)) +
  geom_boxplot() +
  labs(
    x = "Donor",
    y = "Proportion with motion detected",
    title = "Proportion of field of view with motion detected"
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "Proportion_motile_sample_boxplot.png"), width = 8, height = 5.5)
```

## Phase metric plots
Plot FFT phase metrics
```{r}
# Initialise data.frame for phase metrics
phase_data <- data_frame()

# Convert list to data frame
for (i in annotations$Index) {
  phase_data <- bind_rows(
    phase_data,
    c(
      Index = i,
      phase_metrics[[i]]
    )
  )
}

# Merge with annotations and convert columns to numeric
phase_data <- merge(
  phase_data,
  annotations
) %>%
  mutate(
    CircularVariance = as.numeric(CircularVariance),
    Entropy = as.numeric(Entropy),
    MLAD = as.numeric(MLAD),
    DiscontinuityFraction = as.numeric(DiscontinuityFraction)
  )

# Circular variance plot
phase_data %>%
  ggplot(aes(x = Donor, y = CircularVariance, fill = Sample)) +
  geom_boxplot() +
  labs(
    x = "Donor",
    y = "Circular variance",
    title = "Circular variance by sample"
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "Circular_variance_boxplot.png"), width = 8, height = 5.5)

# Entropy plot
phase_data %>%
  ggplot(aes(x = Donor, y = Entropy, fill = Sample)) +
  geom_boxplot() +
  labs(
    x = "Donor",
    y = "Entropy",
    title = "Entropy by sample"
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "Entropy_boxplot.png"), width = 8, height = 5.5)

# MLAD plot
phase_data %>%
  ggplot(aes(x = Donor, y = MLAD, fill = Sample)) +
  geom_boxplot() +
  labs(
    x = "Donor",
    y = "MLAD",
    title = "MLAD by sample"
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "MLAD_boxplot.png"), width = 8, height = 5.5)

# Discunituity fraction plot
phase_data %>%
  ggplot(aes(x = Donor, y = DiscontinuityFraction, fill = Sample)) +
  geom_boxplot() +
  labs(
    x = "Donor",
    y = "Discontinuity fraction",
    title = "Discontinuity fraction by sample"
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "Discontinuity_fraction_boxplot.png"), width = 8, height = 5.5)
```

## Phase metric stats
Run some stats on the phase metrics
```{r Phase metrics stats}
# Run some stats
# Circular variance
CircularVariance_summary <- phase_data %>%
  group_by(Donor, Condition) %>%
  summarize(CircularVariance_mean = mean(CircularVariance), .groups = "drop")

# Perform paired t-test and print results
CircularVariance_test <- t.test(
  filter(CircularVariance_summary, Condition == "pLVUHshp53")$CircularVariance_mean,
  filter(CircularVariance_summary, Condition == "pLL3.7")$CircularVariance_mean,
  paired = TRUE)

# Display results
interpret_t_test(CircularVariance_test, CircularVariance_summary, "CircularVariance_mean", c("pLVUHshp53", "pLL3.7"))

# Discontinuity Fraction variance
DiscontinuityFraction_summary <- phase_data %>%
  group_by(Donor, Condition) %>%
  summarize(DiscontinuityFraction_mean = mean(DiscontinuityFraction), .groups = "drop")

# Perform paired t-test and print results
DiscontinuityFraction_test <- t.test(
  filter(DiscontinuityFraction_summary, Condition == "pLVUHshp53")$DiscontinuityFraction_mean,
  filter(DiscontinuityFraction_summary, Condition == "pLL3.7")$DiscontinuityFraction_mean,
  paired = TRUE)

# Display results
interpret_t_test(DiscontinuityFraction_test, DiscontinuityFraction_summary, "DiscontinuityFraction_mean", c("pLVUHshp53", "pLL3.7"))

# Entropy variance
Entropy_summary <- phase_data %>%
  group_by(Donor, Condition) %>%
  summarize(Entropy_mean = mean(Entropy), .groups = "drop")

# Perform paired t-test and print results
Entropy_test <- t.test(
  filter(Entropy_summary, Condition == "pLVUHshp53")$Entropy_mean,
  filter(Entropy_summary, Condition == "pLL3.7")$Entropy_mean,
  paired = TRUE)

# Display results
interpret_t_test(Entropy_test, Entropy_summary, "Entropy_mean", c("pLVUHshp53", "pLL3.7"))

# MLAD variance
MLAD_summary <- phase_data %>%
  group_by(Donor, Condition) %>%
  summarize(MLAD_mean = mean(MLAD), .groups = "drop")

# Perform paired t-test and print results
MLAD_test <- t.test(
  filter(MLAD_summary, Condition == "pLVUHshp53")$MLAD_mean,
  filter(MLAD_summary, Condition == "pLL3.7")$MLAD_mean,
  paired = TRUE)

# Display results
interpret_t_test(MLAD_test, MLAD_summary, "MLAD_mean", c("pLVUHshp53", "pLL3.7"))
```
