---
title: "CBF_results"
author: "Sam Siljee"
date: "2025-07-18"
output: html_document
---

Script to plot results of apical fluid analysis, based on paper by Harvey et. al.
Harvey, P. R., Tarran, R., Garoff, S., & Myerburg, M. M. (2011). Measurement of the Airway Surface Liquid Volume with Simple Light Refraction Microscopy. American Journal of Respiratory Cell and Molecular Biology, 45(3), 592. https://doi.org/10.1165/RCMB.2010-0484OC

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Packages
library(dplyr) # Data manipulation
library(ggplot2) # Plotting
library(tidyr) # pivot_wider function
library(stringr) # String manipulation to get Index names

# Files and directories
input_dir <- paste0(getwd(), "/data/")
output_dir <- paste0(getwd(), "/../produced_plots/Apical_fluid/")
pre_results_file <- "4x_pre_results.rda"
post_results_file <- "4x_post_results.rda"
colours_file <- "colours.csv"
phenotype_summary_path <- paste0(input_dir, "../../Misc/data/phenotype_summary.csv")
```

## Load the data
Loading the ouput of the multi-DDM analysis run in MATLAB. Results from the .mat object exported as .csv files.
```{r Load the data}
# Load custom functions and colours
source(paste0(input_dir, "../../Functions.R"))

# Read in the results
load(paste0(input_dir, pre_results_file))
pre_results <- results
load(paste0(input_dir, post_results_file))
post_results <- results

# Make annotations from results names
# Create annotations from file names
annotations <- data.frame(ID = names(results)) %>%
  mutate(
    Donor = substr(ID, start = 1, stop = 6),
    Condition = ifelse(grepl("_[123]_", ID), "Control", "p53 KD"),
    Sample = paste(Donor, Condition, sep = "_"),
    Index = str_remove(str_remove(ID, "_90"), "_270")
  )
```

## Process the data
Make summaries of the data and add to phenotype summary
```{r}
# Format data from area_results
area_data <- data.frame()
for (i in annotations$ID) {
  area_data <- bind_rows(
    area_data,
    data.frame(
      ID = i,
      sum_area = pre_results[[i]]["sum_area"],
      sigmoid_area = pre_results[[i]]["sigmoid_area"],
      sigmoid_RMSE = sqrt(mean(residuals(pre_results[[i]]$sigmoid_model)^2)),
      Wash = "Pre"
    ),
    data.frame(
      ID = i,
      sum_area = post_results[[i]]["sum_area"],
      sigmoid_area = post_results[[i]]["sigmoid_area"],
      sigmoid_RMSE = sqrt(mean(residuals(post_results[[i]]$sigmoid_model)^2)),
      Wash = "Post"
    )
  )
}

# Merge with annotations
area_data <- merge(
  area_data,
  annotations,
  by = "ID",
  all.x = TRUE
)

# Summarise, merge with existing data, and write to .csv for the phenotype summary
area_data %>%
  group_by(Index, Wash) %>%
  summarise(
    sum_asl_mean = mean(sum_area),
    sigmoid_asl_mean = mean(sigmoid_area),
    sum_asl_sd = sd(sum_area),
    sigmoid_asl_sd = sd(sigmoid_area),
    sigmoid_asl_rmse = mean(sigmoid_RMSE)
  ) %>%
  pivot_wider(names_from = Wash, values_from = c(3:7), names_glue = "{Wash}_{.value}") %>%
  mutate(Index = str_replace(Index, "-", "_")) %>%
  merge(
    dplyr::select(
      read.csv(phenotype_summary_path),
      -c(
        Post_sum_asl_mean,
        Pre_sum_asl_mean,
        Post_sigmoid_asl_mean,
        Pre_sigmoid_asl_mean,
        Post_sum_asl_sd,
        Pre_sum_asl_sd,
        Post_sigmoid_asl_sd,
        Pre_sigmoid_asl_sd,
        Post_sigmoid_asl_rmse,
        Pre_sigmoid_asl_rmse
      )
    ),
    by = "Index"
  ) %>%
  write.csv(phenotype_summary_path, row.names = FALSE)
```

## Area boxplot - sum area
Boxplots of areas calculated by summing inverse intensity areas
```{r Area boxplot}
# Run t-test to compare
pre_sum_area_summary <- area_data %>%
  filter(Wash == "Pre") %>%
  group_by(Donor, Condition) %>%
  summarize(area_mean = mean(sum_area), .groups = "drop")

# Perform paired t-test
pre_sum_area_test <- t.test(
   filter(pre_sum_area_summary, Condition == "p53 KD")$area_mean,
   filter(pre_sum_area_summary, Condition == "Control")$area_mean
   , paired = TRUE)

# Display results
interpret_t_test(pre_sum_area_test, pre_sum_area_summary, "area_mean")

# Box plot of frequencies by sample
area_data %>%
  filter(Wash == "Pre") %>%
  ggplot(aes(x = Donor, y = sum_area, fill = Sample)) +
  geom_boxplot() +
  labs(
    title = "Apical surface fluid (sum intensity values)",
    x = "Donor",
    y = "Sum intensity values"
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "Sum_area_boxplot.png"), width = 8, height = 5.5)
```

## Area boxplot - sigmoid integration
Boxplots of areas calculated by integrating sigmoid models
```{r Area boxplot}
# Run t-test to compare
pre_sigmoid_area_summary <- area_data %>%
  filter(Wash == "Pre") %>%
  group_by(Donor, Condition) %>%
  summarize(area_mean = mean(sigmoid_area), .groups = "drop")

# Perform paired t-test
pre_sigmoid_area_test <- t.test(
   filter(pre_sigmoid_area_summary, Condition == "p53 KD")$area_mean,
   filter(pre_sigmoid_area_summary, Condition == "Control")$area_mean
   , paired = TRUE)

# Display results
interpret_t_test(pre_sigmoid_area_test, pre_sigmoid_area_summary, "area_mean")

# Box plot of frequencies by sample
area_data %>%
  filter(Wash == "Pre") %>%
  ggplot(aes(x = Donor, y = sigmoid_area, fill = Sample)) +
  geom_boxplot() +
  labs(
    title = "Apical surface fluid (sigmoid integration values)",
    x = "Donor",
    y = "Sigmoid integration values"
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "Sigmoid_area_boxplot.png"), width = 8, height = 5.5)
```

## Sigmoid RMSE boxplots
```{r}
# Run stats on how well the sigmoid curves fit
# Run t-test to compare
pre_sigmoid_RMSE_summary <- area_data %>%
  filter(Wash == "Pre") %>%
  group_by(Donor, Condition) %>%
  summarize(RMSE_mean = mean(sigmoid_RMSE), .groups = "drop")

# Perform paired t-test
pre_sigmoid_RMSE_test <- t.test(
   filter(pre_sigmoid_RMSE_summary, Condition == "p53 KD")$RMSE_mean,
   filter(pre_sigmoid_RMSE_summary, Condition == "Control")$RMSE_mean
   , paired = TRUE)

# Display results
interpret_t_test(pre_sigmoid_RMSE_test, pre_sigmoid_RMSE_summary, "RMSE_mean")

# Box plot of frequencies by sample
area_data %>%
  filter(Wash == "Pre") %>%
  ggplot(aes(x = Donor, y = sigmoid_RMSE, fill = Sample)) +
  geom_boxplot() +
  labs(
    title = "Sigmoid curve RMSE values",
    x = "Donor",
    y = "Sigmoid RMSE values"
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "Sigmoid_RMSE_boxplot.png"), width = 8, height = 5.5)
```
