# Script to try out cilia beat frequency analysis in R
# Sam Siljee
# 19 February 2025

---
title: "CBF analysis"
author: "Sam Siljee"
date: '2022-06-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this script I'm using the variable `d` to represent the donor/well, `x` for the pixel x coordinate, `y` for the pixel y coordinate, and `z` for the frame number.

```{r packages, include = FALSE}
library(dplyr) # Pipe and data manipulation
library(ggplot2) # plotting
library(signal) # Windowing for FFT
library(tiff) # Read in .tiff files
library(png) # Write png images
library(stringr) # Strip ".MOV" from file names
library(tidyr) # For pivot_wider function in FFT data summary
library(EBImage) # For label function to identify clusters
```

```{r Set some constants}
# Directories
input_dir <- "Input/"
output_dir <- "Output/"

# Options for running the script
load_from_rda <- TRUE
write_tables <- FALSE
run_signal_power <- FALSE
run_variance <- FALSE
run_sd <- FALSE
run_phase_variance <- FALSE
run_shannon_phase_entropy <- FALSE
run_MLAD <- FALSE
run_cluster_phase <- FALSE

# Set video details
height <- 360
width <- 640
frame_rate <- 480
num_frames <- 1024

# Set some thresholds
frequency_threshold <- 3.1
cluster_size_threshold <- 3
FFT_ratio_threshold <- 0.014
phase_bins <- 10

# LUT file
annotations_file <- "annotations.tsv"
```

```{r Setup}
# Vector for donor IDs - read from available .MOV files in input folder
donor_IDs <- list.files(input_dir) %>%
  grep(pattern = ".MOV$", value = TRUE) %>%
  str_remove(".MOV")

# Read in annotations
annotations <- read.delim(paste0(input_dir, annotations_file))

# Load functions
source("CBF_functions.R")
```

## Run cilia analysis
Changed to run the whole analysis on a donor by donor basis to speed up processing. only saving the 3D arrays from the video input rather than saving data from every intermittent step.
```{r Run cilia beat frequency analysis - whole process by donor}
# Initialise list to save unconverted FFT data for all donors
FFT_list_unconverted <- list()

# Initialise list to save converted FFT data for all donors
FFT_list_converted <- list()

# Initialise list to save phase metrics
phase_metrics <- list()

# Repeat process for each donor
for (d in donor_IDs) {
  # Print message to keep track of progress
  print(
    paste0(
      "Starting on donor ",
      d,
      "; ",
      which(donor_IDs %in% d),
      " of ",
      length(donor_IDs)
    )
  )

  ## Extract frames / load data ----
  if (load_from_rda) {
    # Load from prepared data
    print("Loading .rda")
    load(paste0(output_dir, "prepared_data/", d, "_3D_array.rda"))
  } else {
    # Read in from video data
    print("Extracting frames")

    # Use ffmpeg to extract all frames as single-channel greyscale TIFF (lossless)
    cmd <- paste0(
      "ffmpeg -i ",
      input_dir,
      d,
      ".MOV -pix_fmt gray -compression_algo lzw -vsync 0 ",
      output_dir,
      "temp_frames/frame_%d.tiff"
    )
    system(cmd)

    # Initialize an empty 3D array
    video_array <- array(NA, dim = c(height, width, num_frames))

    # Read in .png frames and select one channel (gray scale)
    for (z in 1:num_frames) {
      # Read in the frame
      video_array[, , z] <- readTIFF(paste0(output_dir, "temp_frames/frame_", z, ".tiff"), as.is = TRUE)
    }

    # Clear free space
    gc()

    # Remove temp frames from system
    cmd <- paste0(
      "rm ",
      output_dir,
      "temp_frames/frame_*.tiff"
    )
    system(cmd)

    # Save the data as a .rda file - for faster loading in future
    save(video_array, file = paste0(output_dir, "prepared_data/", d, "_3D_array.rda"))
  } # Close data loading else statement

  # Compensate for fluctuating background brightness
  frame_means <- numeric()
  for (z in 1:num_frames) {
    frame_means <- c(frame_means, mean(video_array[, , z]))
  }

  # Centre compensation around zero
  frame_means <- frame_means - mean(frame_means)

  # Subtract compensation from video array
  for (z in 1:num_frames) {
    video_array[, , z] <- video_array[, , z] - frame_means[z]
  }

  ## Signal power ----
  if (run_signal_power) {
    print("Making signal power images")

    # Initialise blank matrix for signal power image
    signal_power_image <- matrix(data = 0, ncol = width, nrow = height)

    # Loop through frames to get differences
    for (z in 2:num_frames) {
      # Get absolute difference with previous frame
      difference <- abs(video_array[, , z] - video_array[, , (z - 1)])

      # Add to signal power image
      signal_power_image <- signal_power_image + difference
    }

    # Save as a PNG image - grayscale
    writePNG(normalise_image(signal_power_image), paste0(output_dir, "signal_power_images/signal_power_image_", d, ".png"))

    # Save as .tsv file
    if (write_tables) {
      write.table(
        signal_power_image,
        paste0(output_dir, "tsv_results/", d, "_signal_power_image.tsv"),
        row.names = FALSE,
        col.names = FALSE,
        sep = "\t"
      )
    }

    # Clear free space
    gc()
  }

  ## Variance ----
  if (run_variance) {
    print("Making variance images")

    # Initialise blank matrix
    variance_matrix <- matrix(ncol = width, nrow = height)

    # Get variance by pixel
    for (y in 1:width) { # loop through columns
      for (x in 1:height) { # Loop through rows
        # Calculate variances and add to list in the required position
        variance_matrix[x, y] <- var(video_array[x, y, ])
      }
    }

    # Save as a PNG image - grayscale
    writePNG(normalise_image(variance_matrix), paste0(output_dir, "variance_images/variance_image_", d, ".png"))

    # Save as .tsv file
    if (write_tables) {
      write.table(
        variance_matrix,
        paste0(output_dir, "tsv_results/", d, "_variance_image.tsv"),
        row.names = FALSE,
        col.names = FALSE,
        sep = "\t"
      )
    }
  }

  ## SD ----
  if (run_sd) {
    print("Making SD images")

    # Initialise blank matrix
    sd_matrix <- matrix(ncol = width, nrow = height)

    # Get variance by pixel
    for (y in 1:width) { # loop through columns
      for (x in 1:height) { # Loop through rows
        # Calculate variances and add to list in the required position
        sd_matrix[x, y] <- sd(video_array[x, y, ])
      }
    }

    # Save as a PNG image - grayscale
    writePNG(normalise_image(sd_matrix), paste0(output_dir, "sd_images/sd_image_", d, ".png"))

    # Save as .tsv file
    if (write_tables) {
      write.table(
        sd_matrix,
        paste0(output_dir, "tsv_results/", d, "_sd_image.tsv"),
        row.names = FALSE,
        col.names = FALSE,
        sep = "\t"
      )
    }
  }

  ## FFT ----
  print("Making FFT images")

  # Create frequency vector
  freq <- seq(0, frame_rate / 2, length.out = num_frames / 2 + 1)

  # Calculate offset to exclude results below the threshold
  threshold_offset <- sum(freq < frequency_threshold)

  # Initialise matrices to contain dominant frequencies
  FFT_matrix <- matrix(ncol = width, nrow = height)
  FFT_matrix_unconverted <- FFT_matrix
  phase_matrix <- FFT_matrix
  FFT_ratio_matrix <- FFT_matrix

  # Calculate FFT and extract dominant frequencies
  for (y in 1:width) { # loop through columns
    for (x in 1:height) { # Loop through rows
      # Get vector of pixel values
      pixel_values <- video_array[x, y, ]

      # Run FFT
      fft_result <- fft(pixel_values)

      # Get magnitudes - using only first half of FFT results
      magnitude <- abs(fft_result)[1:(num_frames / 2 + 1)]

      # Set all magnitudes below frequency threshold to -1 so they won't be selected
      magnitude[1:threshold_offset] <- -1

      # Find the frequency with maximum amplitude
      max_index <- which.max(magnitude)

      # Add to list in the required position - converted to real frequency and unconverted
      FFT_matrix[x, y] <- freq[max_index]
      FFT_matrix_unconverted[x, y] <- max_index

      # Add to phase matrix
      phase_matrix[x, y] <- Arg(fft_result[max_index])

      # Add to ratio matrix (for masking)
      FFT_ratio_matrix[x, y] <- max(magnitude) / sum(magnitude)
    }
  }

  # Add to list of all donors
  FFT_list_unconverted[[d]] <- FFT_matrix_unconverted
  FFT_list_converted[[d]] <- FFT_matrix

  # Save as a PNG image - grayscale
  writePNG(normalise_image(FFT_matrix), paste0(output_dir, "fft_images/FFT_freq_image_", d, ".png"))
  writePNG(normalise_image(phase_matrix), paste0(output_dir, "fft_images/FFT_phase_image_", d, ".png"))

  # Save as .tsv file
  if (write_tables) {
    write.table(
      FFT_matrix,
      paste0(output_dir, "tsv_results/", d, "_fft_image.tsv"),
      row.names = FALSE,
      col.names = FALSE,
      sep = "\t"
    )
    write.table(
      phase_matrix,
      paste0(output_dir, "tsv_results/", d, "_fft_phase_image.tsv"),
      row.names = FALSE,
      col.names = FALSE,
      sep = "\t"
    )
  }

  # Save as a PNG image - grayscale
  writePNG(normalise_image(FFT_matrix_unconverted), paste0(output_dir, "fft_images/FFT_unconverted_freq_image_", d, ".png"))

  # Save as .tsv file
  if (write_tables) {
    write.table(
      FFT_matrix_unconverted,
      paste0(output_dir, "tsv_results/", d, "_fft_unconverted_image.tsv"),
      row.names = FALSE,
      col.names = FALSE,
      sep = "\t"
    )
  }

  # Clear free space
  gc()

  ## Ratio masking ----
  print("FFT ratio images/masking")
  # Save as a PNG image - grayscale
  writePNG(normalise_image(FFT_ratio_matrix), paste0(output_dir, "ratio_images/FFT_ratio_image_", d, ".png"))

  # Save as .tsv file
  if (write_tables) {
    write.table(
      FFT_ratio_matrix,
      paste0(output_dir, "tsv_results/", d, "_fft_ratio_image.tsv"),
      row.names = FALSE,
      col.names = FALSE,
      sep = "\t"
    )
  }

  # Binarise using the threshold as a mask
  # Fill the cluster mask
  ratio_mask <- FFT_ratio_matrix > FFT_ratio_threshold

  # Save as .tsv file
  write.table(
    ratio_mask,
    paste0(output_dir, "tsv_results/", d, "_ratio_mask.tsv"),
    row.names = FALSE,
    col.names = FALSE,
    sep = "\t"
  )

  # Save as a PNG image - grayscale
  writePNG(normalise_image(ratio_mask), paste0(output_dir, "ratio_images/Ratio_mask_", d, ".png"))

  ## Identify clusters ----
  print("Identifying clusters")
  # Use custom function to identify cluster sizes
  cluster_size_matrix <- identify_clusters(FFT_matrix_unconverted)

  # Save as a PNG image - grayscale
  writePNG(normalise_image(cluster_size_matrix), paste0(output_dir, "cluster_images/Cluster_size_image_", d, ".png"))

  # Create binary mask
  # Fill the cluster mask
  cluster_mask <- fill_clusters(cluster_size_matrix) > cluster_size_threshold

  # Save as .tsv file
  write.table(
    cluster_mask,
    paste0(output_dir, "tsv_results/", d, "_cluster_mask.tsv"),
    row.names = FALSE,
    col.names = FALSE,
    sep = "\t"
  )

  # Save as a PNG image - grayscale
  writePNG(normalise_image(cluster_mask), paste0(output_dir, "cluster_images/Cluster_mask_", d, ".png"))

  ## Phase clusters ----
  print("Identifying phase clusters")

  # Save phase image
  writePNG(normalise_image(phase_matrix), paste0(output_dir, "phase_images/Phase_image_", d, ".png"))

  # Use custom function to identify cluster sizes
  phase_cluster_size_matrix <- identify_clusters(
    array(
      as.numeric(cut(phase_matrix, breaks = seq(-pi, pi, length.out = phase_bins))),
      dim = dim(phase_matrix)
    )
  )

  # Save as a PNG image - grayscale
  writePNG(normalise_image(phase_cluster_size_matrix), paste0(output_dir, "phase_images/Phase_cluster_size_image_", d, ".png"))

  # Create binary mask
  # Fill the cluster mask
  phase_cluster_mask <- fill_clusters(phase_cluster_size_matrix) > cluster_size_threshold

  # Save as .tsv file
  write.table(
    cluster_mask,
    paste0(output_dir, "tsv_results/", d, "_phase_cluster_mask.tsv"),
    row.names = FALSE,
    col.names = FALSE,
    sep = "\t"
  )

  # Save as a PNG image - grayscale
  writePNG(normalise_image(phase_cluster_mask), paste0(output_dir, "phase_images/Phase_cluster_mask_", d, ".png"))

  ## Histograms ----
  print("Making histograms")

  # Histogram of converted frequencies - filtered by cluster mask
  png(paste0(output_dir, "histograms/", d, " frequency histogram.png"), width = 8 * 100, height = 5.5 * 100, res = 100)
  print(ggplot(data.frame(Frequency = as.numeric(FFT_matrix[cluster_mask])), aes(x = Frequency)) +
    geom_histogram(binwidth = 1) +
    labs(x = "Frequency (Hz)", y = "Count", title = paste("Histogram of dominant frequency:", d)) +
    theme_bw())
  dev.off()

  # Histogram of unconverted (indexes)
  png(paste0(output_dir, "histograms/", d, " frequency index histogram.png"), width = 8 * 100, height = 5.5 * 100, res = 100)
  print(ggplot(data.frame(Index = as.numeric(FFT_matrix_unconverted[cluster_mask])), aes(x = Index)) +
    geom_histogram(binwidth = 1) +
    labs(x = "Dominant frequency index (binwidth = 1)", y = "Count", title = paste("Histogram of dominant frequency indexes:", d)) +
    theme_bw())
  dev.off()

  ## Phase metrics ----
  phase_metrics[[d]] <- analyze_phase_cluster(phaseImage = phase_matrix, clusterMask = cluster_mask, threshold = pi / 4)
} # Close loop over donors

# Save the FFT data as .rda files
save(FFT_list_unconverted, file = paste0(output_dir, "FFT_list_unconverted.rda"))
save(FFT_list_converted, file = paste0(output_dir, "FFT_list_converted.rda"))
save(phase_metrics, file = paste0(output_dir, "phase_metrics.rda"))
```

## Check bitdepth of signal
Code to check bit depth of signal
```{r Signal bitdepth}
# Initialise vector of mean dit depth values
bit_depth <- numeric()

image_mean_bit_depth <- numeric()

# Repeat process for each donor
for (d in donor_IDs) {
  # Print message to keep track of progress
  print(
    paste0(
      "Starting on donor ",
      d,
      "; ",
      which(donor_IDs %in% d),
      " of ",
      length(donor_IDs)
    )
  )

  ## Extract frames / load data ----
  if (load_from_rda) {
    # Load from prepared data
    print("Loading .rda")
    load(paste0(output_dir, "prepared_data/", d, "_3D_array.rda"))
  } else {
    # Read in from video data
    print("Extracting frames")

    # Use ffmpeg to extract all frames as single-channel greyscale TIFF (lossless)
    cmd <- paste0(
      "ffmpeg -i ",
      input_dir,
      d,
      ".MOV -pix_fmt gray -compression_algo lzw -vsync 0 ",
      output_dir,
      "temp_frames/frame_%d.tiff"
    )
    system(cmd)

    # Initialize an empty 3D array
    video_array <- array(NA, dim = c(height, width, num_frames))

    # Read in .png frames and select one channel (gray scale)
    for (z in 1:num_frames) {
      # Read in the frame
      video_array[, , z] <- readTIFF(paste0(output_dir, "temp_frames/frame_", z, ".tiff"), as.is = TRUE)
    }

    # Clear free space
    gc()

    # Remove temp frames from system
    cmd <- paste0(
      "rm ",
      output_dir,
      "temp_frames/frame_*.tiff"
    )
    system(cmd)

    # Save the data as a .rda file - for faster loading in future
    save(video_array, file = paste0(output_dir, "prepared_data/", d, "_3D_array.rda"))
  } # Close data loading else statement

  # Initialise bitdepth vetor for this image
  image_bit_depth <- numeric()

  ## Bit depth ----
  for (x in 1:height) {
    for (y in 1:width) {
      image_bit_depth <- c(
        image_bit_depth,
        length(unique(video_array[x, y, ]))
      )
    }
  }

  # Add mean image bit depth
  image_mean_bit_depth <- c(image_mean_bit_depth, mean(image_bit_depth))
  bit_depth <- c(bit_depth, image_bit_depth)
} # Close loop over donors

summary(image_mean_bit_depth)
summary(bit_depth)

# Add in metadata and summarise
bit_depth_summary <- merge(
  data.frame(
    ID = donor_IDs,
    bit_depth = image_mean_bit_depth
  ),
  annotations,
  by = "ID"
) %>%
  group_by(Donor, Condition) %>%
  summarize(bit_depth_mean = mean(bit_depth), .groups = "drop")

# Perform paired t-test
bit_depth_test <- t.test(
   dplyr::filter(bit_depth_summary, Condition == "p53 KD")$bit_depth_mean,
   dplyr::filter(bit_depth_summary, Condition == "Control")$bit_depth_mean
   , paired = TRUE)

# Display results
interpret_t_test(bit_depth_test, bit_depth_summary, "bit_depth_mean")
```
