---
title: "MS LFQ QC (Proteome Discoverer)"
author: "Sam Siljee"
output:
  pdf_document: default
  html_document: default
---

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Packages
library(dplyr) # Pipe and data manipulation
library(tidyr) # For pivot_wider call in PSM counts
library(stringr) # Edit sample names in missed cleavages plot
library(ggplot2) # Graphing
library(ggridges) # ridgeline plots
library(vroom) # fast data reading
library(ggVennDiagram) # Make Venn diagrams
library(ComplexHeatmap) # Make heatmaps
library(rawrr) # Extract data for chromatograms
library(MSstats) # To make profile plots for proteins of interest
library(rstatix) # Run ANOVA tests
library(ggpubr) # Add significance to plots

# Set output directory
output_dir <- paste0(getwd(), "/../produced_plots/MS_QC/LFQ/")
input_dir <- paste0(getwd(), "/data/LFQ/")
plot_width <- 8
plot_height <- 5.5
annotations_file <- "run_annotations.tsv"
final_proteins_file <- "LFQ_TMT_common_differential_proteins.txt"
sample_prep_file <- "sample_prep.csv"
Gene_symbol_map <- "gene_symbol_map.tsv"

# Options with running the script
run_chromatograms <- FALSE
load_from_rda <- TRUE
```

## Load data
```{r load final LFQ data, include = FALSE}
# Load custom functions
source(paste0(input_dir, "../../../Functions.R"))

# Load data from .rda files, or from .txt files - depending on options
if (load_from_rda) {
  # Load pre-prepared data
  rda_files <- c(
    "Run_annotations.rda",
    "PSMs.rda",
    "ProteinGroups.rda",
    "Proteins.rda",
    "PeptideGroups.rda",
    "MSMSSpectrumInfo.rda",
    "LFQ_processed.rda",
    "LFQ_processed_no_normalisation.rda"
  )
  for (i in rda_files) {
    load(paste0(input_dir, i))
  }
  
  # Load chromatogram data if required
  if(run_chromatograms) {
    load(paste0(input_dir, "chrom_data.rda"))
  }
  
} else { # Else load data from .txt files
  # Load PD LFQ output from txt files
  files_list <- list.files(input_dir, full.names = TRUE)
  
  # Get lists of PSM and protein groups files separately
  PSM_files <- grep(files_list, pattern = "_PSMs.txt$", value = TRUE)
  PG_files <- grep(files_list, pattern = "_ProteinGroups.txt$", value = TRUE)
  Proteins_files <- grep(files_list, pattern = "_Proteins.txt$", value = TRUE)
  PeptideGroups_files <- grep(files_list, pattern = "_PeptideGroups.txt$", value = TRUE)
  MSMSSpectrumInfo_files <- grep(files_list, pattern = "_MSMSSpectrumInfo.txt$", value = TRUE)
  
  # Load run summary to use as annontations file, add "Path" column to merge with other datasets
  Run_annotations <- read.delim(paste0(input_dir, annotations_file), header = TRUE) %>%
    mutate(
      # Add column to match up to protein and peptide groups files
      PG_path = paste0(
        input_dir,
        "/",
        gsub(".raw", "", Run),
        "_ProteinGroups.txt"
      ),
      Protein_path = paste0(
        input_dir,
        "/",
        gsub(".raw", "", Run),
        "_Proteins.txt"
      ),
      PeptideGroups_path = paste0(
        input_dir,
        "/",
        gsub(".raw", "", Run),
        "_PeptideGroups.txt"
      ),
      # Add columns useful for annotation
      Donor = gsub(" pLL3.7", "", gsub(" pLVUHshp53", "", Sample)),
      Condition = ifelse(grepl(" pLVUHshp53", Sample), "p53 KD", "Control"),
    ) %>%
    # Convert some columns to factors for faster computation
    mutate(
      across(
        c(
          Batch,
          Key,
          Sample,
          Run,
          Donor,
          Condition
        ),
        as.factor
      )
    )
  # Save as r data file for faster loading in future
  save(Run_annotations, file = paste0(input_dir, "Run_annotations.rda"))
  
  # Load PSM data
  PSMs <- data.frame()
  for (i in PSM_files) {
    PSMs <- bind_rows(PSMs, vroom(i))
  }
  # Combine with annotations
  PSMs <- merge(
    PSMs,
    mutate(Run_annotations, "Spectrum File" = paste0(Run, ".raw")),
    by = "Spectrum File",
    all.x = TRUE
  )
  # Save as r data file for faster loading in future
  save(PSMs, file = paste0(input_dir, "PSMs.rda"))
  
  # Load protein group data
  ProteinGroups <- data.frame()
  for (i in PG_files) {
    ProteinGroups <- bind_rows(ProteinGroups, mutate(vroom(i), Path = i))
  }
  # Combine with annotations
  ProteinGroups <- merge(
    ProteinGroups,
    Run_annotations,
    by.x = "Path",
    by.y = "PG_path",
    all.x = TRUE
  )
  # Save as r data file for faster loading in future
  save(ProteinGroups, file = paste0(input_dir, "ProteinGroups.rda"))
  
  # Load protein group data
  Proteins <- data.frame()
  for (i in Proteins_files) {
    Proteins <- bind_rows(Proteins, mutate(vroom(i), Path = i))
  }
  # Combine with annotations
  Proteins <- merge(
    Proteins,
    Run_annotations,
    by.x = "Path",
    by.y = "Protein_path",
    all.x = TRUE
  )
  # Save as r data file for faster loading in future
  save(Proteins, file = paste0(input_dir, "Proteins.rda"))
  
  # Load pepetide group data
  PeptideGroups <- data.frame()
  for (i in PeptideGroups_files) {
    PeptideGroups <- bind_rows(PeptideGroups, mutate(vroom(i), Path = i))
  }
  # Combine with annotations
  PeptideGroups <- merge(
    PeptideGroups,
    Run_annotations,
    by.x = "Path",
    by.y = "PeptideGroups_path",
    all.x = TRUE
  )
  # Save as r data file for faster loading in future
  save(PeptideGroups, file = paste0(input_dir, "PeptideGroups.rda"))
  
  # Load spectrum information data
  MSMSSpectrumInfo <- data.frame()
  for (i in MSMSSpectrumInfo_files) {
    MSMSSpectrumInfo <- bind_rows(MSMSSpectrumInfo, mutate(vroom(i), Path = i))
  }
  # Combine with annotations
  MSMSSpectrumInfo <- merge(
    MSMSSpectrumInfo,
    mutate(Run_annotations, "Spectrum File" = paste0(Run, ".raw")),
    by = "Spectrum File",
    all.x = TRUE
  )
  # Save as r data file for faster loading in future
  save(MSMSSpectrumInfo, file = paste0(input_dir, "MSMSSpectrumInfo.rda"))
  
  # Load data if running chromatograms
  if (run_chromatograms) {
    # Load chromatogram data
    raw_files <- list.files(input_dir, full.names = TRUE) %>%
      grep(pattern = ".raw$", value = TRUE)
    
    # Initialise data.frame
    chrom_data <- data.frame()
    
    # Read in raw files and extract TIC values to data.frame
    for (i in raw_files) {
      # Open file
      tic <- rawrr::readChromatogram(rawfile = i, type = "tic")
      bpc <- rawrr::readChromatogram(rawfile = i, type = "bpc")
      
      chrom_data <- rbind(
        chrom_data,
        data.frame(
          TIC = tic$intensities,
          BP = bpc$intensities,
          RT = as.numeric(tic$times),
          raw_path = i
        )
      )
    }
    
    # Merge with run overview data for annotations
    chrom_data <- merge(
      chrom_data,
      mutate(
        Run_annotations,
        # Add column to match up to protein groups files
        raw_path = paste0(input_dir, "/", Run, ".raw")
      ),
      by = "raw_path",
      all.x = TRUE
    )
    
    # Save as R data for faster loading in future
    save(chrom_data, file = paste0(input_dir, "chrom_data.rda"))
  }
}

# Read in list of final proteins of interest
final_proteins <- readLines(paste0(input_dir, final_proteins_file))

# Load the gene symbols map
gene_symbols_map <- read.delim(paste0(input_dir, Gene_symbol_map), header = TRUE, sep = "\t")

# Load sample pre data
sample_prep_data <- read.delim(paste0(input_dir, sample_prep_file), header = TRUE, sep = ",") %>%
  merge(
    unique(dplyr::select(Run_annotations, Key, Donor, Condition)),
    by.x = "Sample",
    by.y = "Key",
    all.x = TRUE
  ) %>%
  mutate(Sample = paste(Donor, Condition, sep = "_"))

# Create run summaries
# Count number of proteins per run
PG_summary <- ProteinGroups %>%
  group_by(Run) %>%
  summarise("Protein groups" = length(unique(`Protein Groups Protein Group ID`)))

# Count number of PSMs per run
PSM_summary <- PSMs %>%
  group_by(Run) %>%
  summarise("PSMs" = n())

# Count number of spectra per run
MSMSSpectrumInfo_summary <- MSMSSpectrumInfo %>%
  group_by(Run) %>%
  summarise("Spectra" = n())

# Count number of proteins per run
Proteins_summary <- Proteins %>%
  group_by(Run) %>%
  summarise("Proteins" = n())

# Combine with Run_annotations and select desired columns
Run_summary <- merge(
  PG_summary,
  PSM_summary
) %>%
  merge(MSMSSpectrumInfo_summary) %>%
  merge(Proteins_summary) %>%
  merge(Run_annotations) %>%
  mutate("Identification rate" = PSMs / Spectra) %>%
  dplyr::select(Run, Sample, Donor, Condition, Key, Batch, Spectra, PSMs, `Identification rate`, `Protein groups`, Proteins)

# Save as .csv file
write.csv(Run_summary, paste0(output_dir, "Run_summary.csv"))

# Summarise some results
range(Run_summary$Proteins)
mean(Run_summary$Proteins)
```

## Sample preparation
Plot of final final sample prep yields. Data taken from "Colourimetric calculations 20241111.xlsx"
```{r Sample prep}
# Run some stats
# Perform paired t-tests and print results for corrected ratio values
yield_test <- t.test(
  filter(sample_prep_data, Condition == "Control")$Yield,
  filter(sample_prep_data, Condition == "p53 KD")$Yield,
  paired = TRUE
)
interpret_t_test(yield_test, sample_prep_data, "Yield")

# Plot with error bars
sample_prep_data %>%
  ggplot(aes(
    x = Sample,
    y = Yield,
    fill = Sample
  )) +
  geom_col() +
  geom_errorbar(aes(
    ymin = Yield - Yield.SD,
    ymax = Yield + Yield.SD),
    width = 0.2) +
  labs(
    x = "Sample",
    y = "Yield +/- standard deviation",
    title = "Yield by sample"
  ) +
  scale_y_continuous(
    labels = scales::percent,
    limits = c(0, 1)
  ) +
  annotate("text", x = 3, y = 1, label = paste0("paired t-test, p = ", round(yield_test$p.value, 3))) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 90,
      vjust = 0
    )
  ) +
  scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot standard size
ggsave(paste0(output_dir, "Sample prep yield.png"), width = plot_width, height = plot_height)
```

## Missed Cleavages
Plot of missed cleavages, from peptide groups level data
```{r missed cleavages, echo = FALSE}
# Make data summary
missed_cleavages_summary <- PeptideGroups %>%
  group_by(Donor, Condition) %>%
  count(`# Missed Cleavages` = `# Missed Cleavages`) %>%
  mutate(Percent = prop.table(n))

# Perform paired t-tests and print results for corrected ratio values
missed_cleavage_test <- t.test(
  filter(missed_cleavages_summary, Condition == "Control" & `# Missed Cleavages` == 0)$Percent,
  filter(missed_cleavages_summary, Condition == "p53 KD" & `# Missed Cleavages` == 0)$Percent,
  paired = TRUE
)
interpret_t_test(missed_cleavage_test, filter(missed_cleavages_summary, `# Missed Cleavages` == 0), "Percent")

# Run ANOVA for proportion fully tryptic by sample
missed_cleavage_anova <- PeptideGroups %>%
  group_by(Run, Sample) %>%
  count(`# Missed Cleavages` = `# Missed Cleavages`) %>%
  mutate(Percent = prop.table(n)) %>%
  filter(`# Missed Cleavages` == 0) %>%
  ungroup() %>%
  anova_test(Percent ~ Sample)

# Print results
missed_cleavage_anova$p

# Make plot
PeptideGroups %>%
  group_by(Run, Sample) %>%
  count(`# Missed Cleavages` = `# Missed Cleavages`) %>%
  mutate(Percent = prop.table(n), Sample = str_replace(Sample, " ", "_")) %>%
  ggplot(aes(x = Sample, y = Percent, fill = Sample)) +
  stat_summary(fun.data = mean_sd, geom = "bar") +
  stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2) +
  stat_compare_means(
    method = "anova",
    aes(label = paste0("Anova, p = ", signif(after_stat(p), 3)))) +
  facet_wrap("`# Missed Cleavages`") +
  scale_y_continuous(
    limits = c(0, 1),
    labels = function(x) paste0(x * 100, "%")
  ) +
  scale_x_discrete(
    position = "bottom",
    sec.axis = dup_axis(name = "Number of missed cleavages", labels = NULL)
  ) +
  labs(
    title = "Missed cleavages",
    x = "Sample",
    y = "Proportion"
  ) +
  theme_bw() +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank()) +
  scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "LFQ missed cleavages.png"), width = plot_width, height = plot_height)
```

## Chromatograms
```{r Chromatograms, include = FALSE}
if(run_chromatograms) {
  # Make the plot TIC chromatograms
  chrom_data %>%
    ggplot(aes(x = RT, y = TIC, col = Sample)) +
    geom_line() +
    scale_x_discrete(
      position = "bottom",
      sec.axis = dup_axis(name = "Batch", labels = NULL)
    ) +
    scale_y_discrete(
      position = "left",
      sec.axis = dup_axis(name = "Sample", labels = NULL)
    ) +
    labs(
      title = "Total ion current chromatograms for LFQ runs",
      x = "Retention time",
      y = "Total ion current intensity"
    ) +
    facet_grid(Sample ~ Batch, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "none") +
    scale_colour_manual(name = "Sample", values = custom_colours)
  
  # Save the plot standard size
  ggsave(paste0(output_dir, "TIC LFQ chromatograms.png"), width = plot_width, height = plot_height * 2)
  
  # Make the plot BP chromatograms
  chrom_data %>%
    ggplot(aes(x = RT, y = BP, col = Sample)) +
    geom_line() +
    scale_x_discrete(
      position = "bottom",
      sec.axis = dup_axis(name = "Batch", labels = NULL)
    ) +
    scale_y_discrete(
      position = "left",
      sec.axis = dup_axis(name = "Sample", labels = NULL)
    ) +
    labs(
      title = "Base peak chromatograms for LFQ runs",
      x = "Retention time",
      y = "Base peak intensity"
    ) +
    facet_grid(Sample ~ Batch, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "none") +
    scale_colour_manual(name = "Sample", values = custom_colours)
  
  # Save the plot standard size
  ggsave(paste0(output_dir, "BP LFQ chromatograms.png"), width = plot_width, height = plot_height * 2)
}
```

## PSMs By RT
Plot of peptide-spectrum matches by retention time. Coloured by run or condition
```{r PSMs by retention time plot, echo = FALSE}
# Make the plot by sample with facet grid by batch/sample
PSMs %>%
  ggplot(aes(x = `RT [min]`, color = Sample)) +
  geom_density() +
  scale_y_continuous(
    name = "Peptide-spectrum matches",
    breaks = NULL,
    position = "left",
    sec.axis = dup_axis(name = "Sample", labels = NULL)
  ) +
  scale_x_discrete(
    position = "bottom",
    sec.axis = dup_axis(name = "Batch", labels = NULL)
  ) +
  facet_grid(Sample ~ Batch, scales = "free_y") +
  labs(
    title = "PSMs by retention time for LFQ runs",
    x = "Retention time",
    y = "PSM identifications"
  ) +
  theme_bw() +
  theme(
    legend.position = "none") +
  scale_colour_manual(name = "Sample", values = custom_colours)

# Save the plot standard size
ggsave(paste0(output_dir, "LFQ PSMs by RT.png"), width = plot_width, height = plot_height * 2)
```

## Peak widths
My approach here is to get some extracted ion chromatograms and then measure peak width using full width half height.
LFQ_peak_width_selection.R script used to determine the peptide to use.
Use the peptide "[R].VATVSLPR.[S]" (from pig trypsin), RT ~40 minutes, m/z ~421.7584
```{r}
# List raw files
raw_files <- list.files(input_dir, full.names = TRUE) %>%
  grep(pattern = ".raw$", value = TRUE)

# Initialise vector for peak widths
peak_widths <- numeric(length = length(raw_files))
names(peak_widths) <- raw_files

# Loop through_files
for(n in 1:length(raw_files)) {
  print(paste0("Starting on run ", n, " of ", length(raw_files)))
  i <- raw_files[n]
  peak_width <- calculate_peak_width(
    raw_file = i,
    peak_mass = 421.7584,
    tol = 5,
    min_RT = 39.1 - 5,
    max_RT = 39.1 + 5
  )
  
  # Assign to vector
  peak_widths[i] <- peak_width * 60
}

# Merge with annotations
peak_widths <- merge(
  mutate(
    Run_annotations,
    index = paste0(input_dir, "/", Run, ".raw")
  ),
  data.frame(
    Peak_width = peak_widths,
    index = names(peak_widths)
  ),
  by = "index"
)

# Run some stats
# Run ANOVA for peak width by sample
peak_widths %>%
  anova_test(Peak_width ~ Sample) %>%
  .$p

# Run ANOVA for peak width by condition
peak_widths %>%
  anova_test(Peak_width ~ Condition) %>%
  .$p

# Run ANOVA for peak width by batch
peak_widths %>%
  anova_test(Peak_width ~ Batch) %>%
  .$p

# Get mean peak width
mean(peak_widths$Peak_width)

# Example extracted ion chromatogram
XIC <- rawrr::readChromatogram(rawfile = raw_files[1], mass = 421.7584, tol = 5, type = "xic")

# Extract to a data.frame and make plot
data.frame(
  RT = as.numeric(XIC[[1]]$times),
  Intensity = XIC[[1]]$intensities
) %>%
  ggplot(aes(x = RT, y = Intensity)) +
  geom_line() +
  labs(
    title = "Extracted-ion chromatogram m/z: 421.7584",
    x = "Retention time"
  ) +
  theme_bw() +
  theme(legend.position = "none")

# Save the plot standard size
ggsave(paste0(output_dir, "LFQ XIC.png"), width = plot_width, height = plot_height)

# Box plot of frequencies by sample
peak_widths %>%
  ggplot(aes(x = Batch, y = Peak_width, fill = Batch)) +
  geom_boxplot() +
  stat_compare_means(
    method = "anova",
    aes(label = paste0("Anova, p = ", signif(after_stat(p), 3)))) +
  labs(
    title = "Full width at half maximum (FWHM) peak widths for peptide  \"VATVSLPR\" in LFQ runs by batch",
    x = "Batch",
    y = "Peak width (seconds)"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "LFQ_peak_width_boxplot.png"), width = plot_width, height = plot_height)
```

## RT analysis
Decided to run analysis as follows:
1. Group PSMs by annotated sequence - get global mean RTs
2. Group PSMs by run and annotated sequence, get run specific RTs
3. Merge
4. Subtract global means from run specific means and look at distribution
```{r RT analysis}
# Get global RT means
glob_RT_means <- PSMs %>%
  group_by(`Annotated Sequence`) %>%
  summarise(glob_RT = mean(`RT [min]`, na.rm = TRUE), .groups = "drop")

# Get run specific means and mean differences
run_RT_means <- PSMs %>%
  group_by(Run, `Annotated Sequence`) %>%
  summarise(run_RT = mean(`RT [min]`, na.rm = TRUE), .groups = "drop") %>%
  merge(glob_RT_means, by = "Annotated Sequence", all.x = TRUE) %>%
  mutate(RT_dif = run_RT - glob_RT) %>%
  merge(Run_annotations, all.x = TRUE)

# Run ANOVA for proportion for RT differences by run
RT_anova <- run_RT_means %>%
  anova_test(RT_dif ~ Run)

# Print results
RT_anova$p

# Print mean absolute retention time differences
mean(abs(run_RT_means$RT_dif))

# Print proportion of absolute delta RT less than 5 minutes
mean(abs(run_RT_means$RT_dif) < 5)
mean(abs(run_RT_means$RT_dif) < 0.5)

# Ridgeline plot of peak width by batch
run_RT_means %>%
  filter(RT_dif < 2 & RT_dif > -2) %>%
  ggplot(aes(x = RT_dif, y = Batch, fill = Batch)) +
  geom_density_ridges() +
  facet_wrap("Sample") +
  labs(
    title = "Peptide retention time differences by run",
    x = "Peptide retention time difference relative to overall mean (minutes)",
    y = "Density"
  ) +
  scale_y_discrete(limits = rev) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

# Save the plot
ggsave(paste0(output_dir, "LFQ RT difference ridgeline plot.png"), width = plot_width, height = plot_height * 2)
```

## Protein identifications by Sample
```{r Protein identifications by key}
# Summarise counts from PSM data
ID_counts <- PSMs %>%
  dplyr::select(c(Run, `Master Protein Accessions`)) %>%
  unique() %>%
  group_by(Run) %>%
  summarise(Protein_count = n()) %>%
  merge(Run_annotations, by = "Run")

# Run stats and averages
ID_test <- ID_counts %>%
  anova_test(Protein_count ~ Batch)
ID_test$p
round(mean(ID_counts$Protein_count))

# Make a plot
ID_counts %>%
  ggplot(aes(x = Sample, y = Protein_count, col = Batch)) +
  geom_point() +
  annotate("text", x = 3, y = 5000, label = paste0("ANOVA, p = ", round(ID_test$p, 3))) +
  expand_limits(y = 0) +
  ggtitle("Identifications by run") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0, hjust = 1)
  )

# Save the plot
ggsave(paste0(output_dir, "LFQ IDs by run.png"), width = plot_width, height = plot_height)
```

## Mass Error
Plot of mass error by run
```{r precursor mass error, echo = FALSE}
# Run ANOVA for proportion for mass error by batch
# Summarise data - arbitrary significance from large sample size
mass_er_summary <- PSMs %>%
  group_by(Run, Batch) %>%
  summarise(
    mz = mean(`Deltam/z [Da]`),
    ppm = mean(`DeltaM [ppm]`),
    .groups = "drop"
  )

# Run ANOVA for m/z
mz_anova <- mass_er_summary %>%
  anova_test(mz ~ Batch)
mz_anova$p

# Run ANOVA for ppm
ppm_anova <- mass_er_summary %>%
  anova_test(ppm ~ Batch)
ppm_anova$p

# Summaries
mean(PSMs$`Deltam/z [Da]`)
sd(PSMs$`Deltam/z [Da]`)
mean(PSMs$`DeltaM [ppm]`)
sd(PSMs$`DeltaM [ppm]`)

# Mean absolute errors
mean(abs(PSMs$`Deltam/z [Da]`))
mean(abs(PSMs$`DeltaM [ppm]`))
mean(abs(PSMs$`DeltaM [ppm]`) < 1)

# Density plot of precursor mass error - PSM level
PSMs %>%
  ggplot(aes(x = `Deltam/z [Da]`, colour = Batch)) +
  geom_density(alpha = 0.2) +
  labs(
    title = "Mass error - PSM level by run, LFQ data",
    x = expression(~Delta ~ "m/z (Da)"),
    y = "Density") +
  facet_wrap("Sample", scales = "free_y") +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# Save the plot
ggsave(paste0(output_dir, "LFQ PSM mass error by run.png"), width = plot_width, height = plot_height)

# Density plot of precursor mass error - PSM level - by Batch - ppm
PSMs %>%
  ggplot(aes(x = `DeltaM [ppm]`, colour = Batch)) +
  geom_density(alpha = 0.2) +
  labs(
    title = "Mass error - PSM level by run, LFQ data",
    x = expression(~Delta ~ "m/z (ppm)"),
    y = "Density") +
  facet_wrap("Sample", scales = "free_y") +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

# Save the plot
ggsave(paste0(output_dir, "LFQ PSM mass error by run ppm.png"), width = plot_width, height = plot_height)
```

## Venn diagrams - reproducibility
```{r}
# Loop through all samples
for(i in Run_annotations$Sample){
  # Filter data and make a list of unique proteins in each replicate
  venn_list <- PSMs %>%
    filter(Sample == i) %>%
    dplyr::select(c(Batch, `Master Protein Accessions`)) %>%
    unique() %>%
    split(.$Batch) %>%
    lapply(function(df) {
      df %>%
        dplyr::select(`Master Protein Accessions`) %>%
        pull()
    })
  
  # make the plot
  ggVennDiagram(venn_list) +
    theme_void() +
    labs(
      title = paste0(i, " technical replicate overlap")
      ) +
    ggplot2::scale_fill_gradient(low = "white", high = "red")
  
  # Save the plot
  ggsave(paste0(output_dir, i, " venn diagram.png"), width = plot_width, height = plot_height)
}
```

## Missing values
Look at imputation numbers from MSstats - present as heatmap
```{r missing values heatmap}
# Overview of imputation numbers
MSstats_processed$ProteinLevelData %>%
  mutate(imputation = NumImputedFeature/(NumImputedFeature + NumMeasuredFeature)) %>%
  .$imputation %>%
  mean

# Create the heatmap
# Create the presence/absence matrix, summarise to sample-level using identification in at least one run.
heatmap_data <- MSstats_processed$ProteinLevelData %>%
  mutate(Run = originalRUN) %>%
  merge(
    mutate(Run_annotations, Run = paste0(Run, "raw")),
    by = "Run"
  ) %>%
  dplyr::select(Sample, Protein) %>%
  unique() %>%
  mutate(Present = TRUE) %>%
  complete(Sample, Protein, fill = list(Present = FALSE))

# Convert to wide format for clustering
wide_matrix <- heatmap_data %>%
  pivot_wider(names_from = Sample, values_from = Present, values_fill = FALSE) %>%
  tibble::column_to_rownames("Protein") %>%
  as.matrix()

# Perform hierarchical clustering on proteins (rows)
protein_dist <- dist(wide_matrix * 1, method = "binary")  # binary distance for TRUE/FALSE
protein_hclust <- hclust(protein_dist)
protein_order <- rownames(wide_matrix)[protein_hclust$order]

# calculate proportion missing values
signif(mean(!wide_matrix) * 100, digits = 3)

# Plot with ordered proteins
heatmap_data %>%
  mutate(Protein = factor(Protein, levels = protein_order)) %>%
  ggplot(aes(x = Sample, y = Protein, fill = Present)) +
  geom_tile() +
  scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white")) +
  theme_bw() +
  labs(
    x = "Sample",
    y = "Protein",
    fill = "Identified",
    title = "Protein identification missing values") +
  theme(
    axis.text.x = element_text(angle = 90),
    axis.text.y = element_blank(),
    panel.grid = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank()
  )

# Save the plot
ggsave(paste0(output_dir, "LFQ missing values heatmap.png"), width = plot_width, height = plot_height)
```

## Normalisation
```{r Normalisation}
# Protein level normalisation plots
# Prepare the data
norm_data <- rbind(
  mutate(
    MSstats_processed$ProteinLevelData,
    Normalisation = "Normalised"
  ),
  mutate(
    MSstats_processed_no_normalisation$ProteinLevelData,
    Normalisation = "Original"
  )
) %>%
  dplyr::select(c(originalRUN, LogIntensities, Normalisation))

# Make the plot
norm_data %>%
  ggplot(aes(x = as.numeric(originalRUN), y = LogIntensities, fill = originalRUN)) +
  geom_boxplot() +
  stat_compare_means(
    method = "anova",
    label.x = 6,
    aes(label = paste0("Anova, p = ", signif(after_stat(p), 3)))) +
  facet_wrap("Normalisation") +
  scale_x_continuous(breaks = 1:64) +
  labs(
    x = "Run",
    y = "Protein-level abundance",
    title = "Protein level LFQ protein abundances by run",
    fill = "Donor"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(
      angle = 90,
      vjust = 0
    ),
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "LFQ protein level run intensities.png"), width = plot_width, height = plot_height)
```

## Precursor charge
```{r Precursor charge}
PSMs %>%
  mutate(Charge = as.factor(Charge)) %>%
  group_by(Run, Charge) %>%
  summarise(count = n()) %>%
  group_by(Charge) %>%
  summarise(Proportion = sum(count), SD = sd(count)) %>%
  mutate(Proportion = Proportion / nrow(PSMs), SD = SD / nrow(PSMs)) %>%
  ggplot(aes(x = Charge, y = Proportion, ymin = Proportion - SD, ymax = Proportion + SD, fill = Charge)) +
  geom_col() +
  geom_errorbar(width = 0.2) +
  scale_y_continuous(
    limits = c(0, 0.7),
    labels = function(x) paste0(x * 100, "%")
  ) +
  labs(
    x = "Precursor charge",
    y = "Proportion of PSMs",
    title = "LFQ precursor charge distribution"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "LFQ precursor charge.png"), width = plot_width, height = plot_height)
```

## Peptide length
```{r}
# Summarise data
peptide_lengths <- PSMs %>%
  mutate(
    Length = as.factor(nchar(gsub("^\\[.*?\\].|.\\[.*?\\]$", "", `Annotated Sequence`)))) %>%
  group_by(Run, Length) %>%
  summarise(count = n()) %>%
  group_by(Length) %>%
  summarise(Proportion = sum(count), SD = sd(count)) %>%
  mutate(Proportion = Proportion / nrow(PSMs), SD = SD / nrow(PSMs))

# Present most common length
peptide_lengths[which.max(peptide_lengths$Proportion),]
sum(as.numeric(peptide_lengths$Length) * peptide_lengths$Proportion, na.rm = TRUE)

# Make a plot
peptide_lengths %>%
  ggplot(aes(x = Length, y = Proportion, ymin = Proportion - SD, ymax = Proportion + SD, fill = Length)) +
  geom_col() +
  geom_errorbar(width = 0.2) +
  scale_y_continuous(
    labels = function(x) paste0(x * 100, "%")
  ) +
  labs(
    x = "Peptide length",
    y = "Proportion of PSMs",
    title = "Peptide length - LFQ runs"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "LFQ peptide lengths.png"), width = plot_width, height = plot_height)
```

## Unique peptides
```{r}
# Get some stats
mean(Proteins$`# Unique Peptides` == 0) * 100
mean(Proteins$`# Unique Peptides` == 1) * 100

# Summarise the data
unique_peptides <- Proteins %>%
  group_by(Run, `# Unique Peptides`) %>%
  mutate(`# Unique Peptides` = as.factor(
    ifelse(as.numeric(`# Unique Peptides`) >= 20, "20+", as.character(`# Unique Peptides`))
  )) %>%
  summarise(count = n()) %>%
  group_by(`# Unique Peptides`) %>%
  summarise(Proportion = sum(count), SD = sd(count)) %>%
  mutate(Proportion = Proportion / nrow(Proteins), SD = SD / nrow(Proteins))

# Make a plot
unique_peptides %>%
  ggplot(aes(x = `# Unique Peptides`, y = Proportion, ymin = Proportion - SD, ymax = Proportion + SD, fill = `# Unique Peptides`)) +
  geom_col() +
  geom_errorbar(width = 0.2) +
  scale_y_continuous(
    labels = function(x) paste0(x * 100, "%")
  ) +
  labs(
    x = "Number of unique peptides per protein",
    y = "Proportion of identified proteins",
    title = "Protein unique peptide counts - LFQ runs"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "LFQ unique peptides.png"), width = plot_width, height = plot_height)
```

## Protein coverage
```{r}
mean(Proteins$`Coverage [%]`)
mean(Proteins$`Coverage [%]` > 3.74)

# Density plot of protein coverage
Proteins %>%
  ggplot(aes(x = `Coverage [%]`)) +
  geom_density() +
  labs(
    x = "Protein sequence coverage (%)",
    y = "Density",
    title = "Protein sequence coverage - LFQ runs"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

# Save the plot
ggsave(paste0(output_dir, "LFQ protein coverage.png"), width = plot_width, height = plot_height)
```

## Sequest score
```{r}
mean(Proteins$`Score Sequest HT: Sequest HT`)

# Density plot of protein coverage
Proteins %>%
  ggplot(aes(x = `Score Sequest HT: Sequest HT`)) +
  geom_density() +
  scale_x_continuous(trans = "log2") +
  labs(
    x = "Sequest HT score (log 2 scale)",
    y = "Density",
    title = "Sequest HT scores - LFQ runs"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

# Save the plot
ggsave(paste0(output_dir, "LFQ SEQUEST scores.png"), width = plot_width, height = plot_height)
```

## UNUSED CODES
Codes for older plots, no longer required.

## PTM distribution
Code to test out the distribution of PTMs found in the PSMs. Used to justify which to include in the final search.
```{r PTM distribution}
# Load list of searched modifications
modifications_list <- readLines("data/modifications_list.txt")
# Load the data - select modifications
modifications_data <- vroom("data/PTM_optimistation_PSMs.txt") %>% .$Modifications
# Get total numbers for ratio calculations
number_of_PSMs <- length(modifications_data)
# Calculate the numbers of times that modifications occur
# Do the counting
counts <- numeric()
for (i in modifications_list) {
  count <- sum(grepl(i, modifications_data))
  counts <- c(counts, count)
}
# Make a data frame and plot
data.frame(
  Modification = modifications_list,
  Proportion = counts / number_of_PSMs
) %>%
  ggplot(aes(x = Modification, y = Proportion, fill = Modification)) +
  geom_col() +
  ggtitle("Rates of modification of PSMs by modification") +
  theme_bw() +
  theme(legend.position = "none")

# Save the plot
ggsave(paste0(output_dir, "modifications.png"), width = plot_width, height = plot_height)
```

## Identification Efficiency
```{r identification efficiency, echo = FALSE}
# Count PSMs
TotalPSMs <- PSMs %>%
  group_by(`Spectrum File`) %>%
  summarise(PSMs = n())

# Count spectra
TotalSpectra <- MSMSSpectrumInfo %>%
  group_by(`Spectrum File`) %>%
  summarise(Spectra = n())

# Combine and calculate
EfficiencyTab <- merge(TotalPSMs, TotalSpectra, by = "Spectrum File") %>%
  mutate("Identification rate" = paste(round(PSMs / Spectra * 100, 1), "%"))

# Display as table
knitr::kable(EfficiencyTab)
```
