---
title: "MS TMT QC (Proteome Discoverer)"
author: "Sam Siljee"
output:
  pdf_document: default
  html_document: default
---
## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Packages
library(dplyr) # Data wrangling
library(tidyr) # For `pivot_longer` in labelling efficiency graph, and replace_na for Modifications
library(stringr) # For annotating labelling efficiency and channel abundances
library(ggplot2) # Plotting
library(ggridges) # For ridgeline plots
library(vroom) # Fast reading of data
library(ggVennDiagram) # New package to make venn diagrams
library(rawrr) # For loading raw data for chromatograms
library(Peptides) # for calculating GRAVY scores
library(ComplexUpset) # Used to make Upset plots
# library(gganimate) # For animated chromatograms
library(MSstatsTMT) # For profile plots
library(rstatix) # to run ANOVA test
library(ggpubr) # Add stats to plots

# Set some variables
load_from_rda <- TRUE
run_chromatograms <- FALSE
plot_width <- 8
plot_height <- 5.5
input_dir <- paste0(getwd(), "/data/TMT/")
efficiency_dir <- paste0(getwd(), "/data/TMT_labelling_efficiency/")
overlabelling_dir <- paste0(getwd(), "/data/TMT_overlabelling/")
mixing_dir <- paste0(getwd(), "/data/TMT_mixing_ratio/")
output_dir <- paste0(getwd(), "/../produced_plots/MS_QC/TMT/")
run_annotations_file <- "run_annotations.tsv"
channel_annotations_file <- "channel_annotations.tsv"
final_proteins_file <- "LFQ_TMT_common_differential_proteins.txt"
Gene_symbol_map <- "gene_symbol_map_manually_added.tsv"
channels_vector <- c(
  "126",
  paste0(rep(127:134, each = 2), c("N", "C")),
  "135N"
)
```

## Load the data
```{r load data, include = FALSE}
# Load custom functions
source(paste0(input_dir, "../../../Functions.R"))

# Load the data
if (load_from_rda) {
  # load files from .rda files for faster loading
  rda_files <- c(
    "Run_annotations.rda",
    "Channel_annotations.rda",
    "MSMSSpectrumInfo.rda",
    "ProteinGroups.rda",
    "PSMs.rda",
    "QuanSpectra.rda",
    "PeptideGroups.rda",
    "Proteins.rda",
    "TMT_processed.rda",
    "TMT_processed_no_normalisation.rda",
    "Efficiency_PSMs.rda",
    "Overlabelling_PSMs.rda",
    "Mixing_PSMs.rda"
  )
  for (i in rda_files) {
    load(paste0(input_dir, i))
  }
  
  # Load chromatogram data if required
  if(run_chromatograms) {
    load(paste0(input_dir, "chrom_data.rda"))
  }
  
} else {
  # Load files from .txt data to refresh data
  files_list <- list.files(input_dir, full.names = TRUE)

  # Load run summary to use as annontations file, add "Path" column to merge with other datasets
  Run_annotations <- read.delim(paste0(input_dir, run_annotations_file), header = TRUE) %>%
    # Convert some columns to factors for faster computation
    mutate(
      across(
        c(
          Batch,
          Run,
          Fraction
        ),
        as.factor
      )
    )
  # Save for faster loading in future
  save(Run_annotations, file = paste0(input_dir, "Run_annotations.rda"))
  
  # Load channel annotations
  Channel_annotations <- read.delim(paste0(input_dir, channel_annotations_file), header = TRUE) %>%
    # Convert some columns to factors for faster computation
    mutate(
      across(
        c(
          Mixture,
          Channel,
          Condition,
          Donor
        ),
        as.factor
      )
    )
  # Save for faster loading in future
  save(Channel_annotations, file = paste0(input_dir, "Channel_annotations.rda"))

  # Load different types of data - and save as R data files for faster loading in future
  MSMSSpectrumInfo <- import_txt_data("MSMSSpectrumInfo", input_dir, files_list, Run_annotations)
  save(MSMSSpectrumInfo, file = paste0(input_dir, "MSMSSpectrumInfo.rda"))

  ProteinGroups <- import_txt_data("ProteinGroups", input_dir, files_list, Run_annotations)
  save(ProteinGroups, file = paste0(input_dir, "ProteinGroups.rda"))

  PSMs <- import_txt_data("PSMs", input_dir, files_list, Run_annotations)
  save(PSMs, file = paste0(input_dir, "PSMs.rda"))

  QuanSpectra <- import_txt_data("QuanSpectra", input_dir, files_list, Run_annotations)
  save(QuanSpectra, file = paste0(input_dir, "QuanSpectra.rda"))

  PeptideGroups <- import_txt_data("PeptideGroups", input_dir, files_list, Run_annotations)
  save(PeptideGroups, file = paste0(input_dir, "PeptideGroups.rda"))

  Proteins <- import_txt_data("Proteins", input_dir, files_list, Run_annotations)
  save(Proteins, file = paste0(input_dir, "Proteins.rda"))
  
  Efficiency_PSMs <- import_txt_data("PSMs", efficiency_dir, list.files(efficiency_dir, full.names = TRUE), Run_annotations)
  save(Efficiency_PSMs, file = paste0(input_dir, "Efficiency_PSMs.rda"))
  
  Overlabelling_PSMs <- import_txt_data("PSMs", overlabelling_dir, list.files(overlabelling_dir, full.names = TRUE), Run_annotations)
  save(Overlabelling_PSMs, file = paste0(input_dir, "Overlabelling_PSMs.rda"))
  
  Mixing_PSMs <- import_txt_data("PSMs", mixing_dir, list.files(mixing_dir, full.names = TRUE), Run_annotations)
  save(Mixing_PSMs, file = paste0(input_dir, "Mixing_PSMs.rda"))
  
  # Load data if running chromatograms
  if (run_chromatograms) {
    # Load chromatogram data
    raw_files <- list.files(input_dir, full.names = TRUE) %>%
      grep(pattern = ".raw$", value = TRUE)
    
    # Initialise data.frame
    chrom_data <- data.frame()
    
    # Read in raw files and extract TIC values to data.frame
    for (i in raw_files) {
      # Open file
      tic <- rawrr::readChromatogram(rawfile = i, type = "tic")
      bpc <- rawrr::readChromatogram(rawfile = i, type = "bpc")
      
      chrom_data <- rbind(
        chrom_data,
        data.frame(
          TIC = tic$intensities,
          BP = bpc$intensities,
          RT = as.numeric(tic$times),
          raw_path = i
        )
      )
    }
    
    # Merge with run overview data for annotations
    chrom_data <- merge(
      chrom_data,
      mutate(
        Run_annotations,
        # Add column to match up to protein groups files
        raw_path = paste0(input_dir, "/", Run, ".raw")
      ),
      by = "raw_path",
      all.x = TRUE
    )
    
    # Save as R data for faster laoding in future
    save(chrom_data, file = paste0(input_dir, "chrom_data.rda"))
  }
}

# Read in list of final proteins of interest
final_proteins <- readLines(paste0(input_dir, final_proteins_file))

# Load the gene symbols map
gene_symbols_map <- read.delim(paste0(input_dir, Gene_symbol_map), header = TRUE, sep = "\t")

# Calculate GRAVY score for fractionation testing
PSMs <- PSMs %>%
  mutate(
    # Create a trimmed sequence to count lysines - ignoring lysines before and after
    `Trimmed sequence` = gsub("^\\[.*?\\].|.\\[.*?\\]$", "", `Annotated Sequence`),
    "GRAVY score" = hydrophobicity(`Trimmed sequence`, scale = "KyteDoolittle")
  )

# Add in columns for over labelling and labelling efficiency
Efficiency_PSMs <- Efficiency_PSMs %>%
  mutate(
    # Replace NAs in the modifications column for peptides with no modifications
    Modifications = replace_na(Modifications, "None"),
    # Create a trimmed sequence to count lysines - ignoring lysines before and after
    `Trimmed sequence` = gsub("^\\[.*?\\].|.\\[.*?\\]$", "", `Annotated Sequence`),
    # Count some residues for relative modification rates
    `Lysine count` = str_count(tolower(`Trimmed sequence`), "k"),
    # Count numbers of tags on specific residues
    `Lysine tag count` = str_count(Modifications, "K\\d{1,2}\\(TMTpro"),
    # Add some columns for fully tagged or partially tagged
    `N-Term tagged` = grepl("N-Term(TMTpro)", Modifications, fixed = TRUE),
    `Lysine tagged` = ifelse(
      `Lysine count` != 0, # Only count sequences containing lysine
      ifelse(
        `Lysine tag count` == `Lysine count`,
        TRUE, # All lysines are tagged
        FALSE # Not all lysines are tagged
      ),
      NA
    ), # Return NA if no lysine in the sequence
    `Completely tagged` = `N-Term tagged` & (`Lysine tagged` | is.na(`Lysine tagged`))
  )

# Add in columns for over labelling and labelling efficiency
Overlabelling_PSMs <- Overlabelling_PSMs %>%
  mutate(
    # Replace NAs in the modifications column for peptides with no modifications
    Modifications = replace_na(Modifications, "None"),
    # Create a trimmed sequence to count lysines - ignoring lysines before and after
    `Trimmed sequence` = gsub("^\\[.*?\\].|.\\[.*?\\]$", "", `Annotated Sequence`),
    # Count some residues for relative modification rates
    `Histidine count` = str_count(tolower(`Trimmed sequence`), "h"),
    `Serine count` = str_count(tolower(`Trimmed sequence`), "s"),
    `Threonine count` = str_count(tolower(`Trimmed sequence`), "t"),
    `Tyrosine count` = str_count(tolower(`Trimmed sequence`), "y"),
    # Count numbers of tags on specific residues
    `Histidine tag count` = str_count(Modifications, "H\\d{1,2}\\(TMTpro"),
    `Serine tag count` = str_count(Modifications, "S\\d{1,2}\\(TMTpro"),
    `Threonine tag count` = str_count(Modifications, "T\\d{1,2}\\(TMTpro"),
    `Tyrosine tag count` = str_count(Modifications, "Y\\d{1,2}\\(TMTpro")
  )

# Add in details to mixing ratio PSMs
Mixing_PSMs <- Mixing_PSMs %>%
  mutate(Mix = ifelse(`Spectrum File` == "TMT_mixing_ratio_new_mix.raw", "Corrected", "Uncorrected"))
```

## Labelling efficiency
```{r labelling efficiency, echo = FALSE}
# Check overall complete tagging rate
print(paste0("Overall completely tagged: ", round(
  mean(Efficiency_PSMs$`Completely tagged`)
  * 100, 2), "%"))
# Check partially tagged peptides
print(paste0("Overall at least one tag: ", round(
mean(Efficiency_PSMs$`N-Term tagged` | Efficiency_PSMs$`Lysine tagged`, na.rm = TRUE)
  * 100, 2), "%"))
# Check N-terminus tagging rate
print(paste0("Overall N-terminus tagged: ", round(
mean(Efficiency_PSMs$`N-Term tagged`)
  * 100, 2), "%"))
# Check Lysine tagging rate
print(paste0("Overall N-lysine tagged: ", round(
mean(Efficiency_PSMs$`Lysine tagged`, na.rm = TRUE)
  * 100, 2), "%"))
# Tagging percentage by fraction
Efficiency_PSMs %>%
  group_by(Fraction) %>%
  summarise("Percent fully tagged" = round(mean(`Completely tagged`) * 100, 4))
# Fraction 1 fully tagged
print(paste0("Fraction 1 completely tagged: ", round(
  mean(filter(Efficiency_PSMs, Fraction == 1)$`Completely tagged`)
  * 100, 2), "%"))
# Fraction 7 fully tagged
print(paste0("Fraction 7 completely tagged: ", round(
  mean(filter(Efficiency_PSMs, Fraction == 7)$`Completely tagged`)
  * 100, 2), "%"))

# Use reporter ion abundance missing values as a proxy for over or under tagging by sample
labelling_summary <- Efficiency_PSMs %>%
  pivot_longer(
    cols = starts_with("Abundance: "),
    values_to = "Abundance",
    names_to = "Channel",
    names_prefix = "Abundance: "
  ) %>%
  filter(`Completely tagged` == TRUE) %>%
  group_by(Channel, Run) %>%
  summarise(prop_na = mean(is.na(Abundance))) %>%
  merge(Channel_annotations, by = "Channel") %>%
  arrange(Donor)

# Perform paired t-tests and print results
test_summary <- labelling_summary %>%
  filter(Condition != "Norm") %>%
  group_by(Condition, Donor) %>%
  summarise(prop_na = mean(prop_na))
labelling_test <- t.test(
  filter(test_summary, Condition == "pLL3.7")$prop_na,
  filter(test_summary, Condition == "pLVUHshp53")$prop_na,
  paired = TRUE
)
interpret_t_test(labelling_test, test_summary, "prop_na", conditions = c("pLVUHshp53", "pLL3.7"))

# Calculate mean missing values by condition
mean(filter(test_summary, Condition == "pLVUHshp53")$prop_na)
mean(filter(test_summary, Condition == "pLL3.7")$prop_na)

# Run ANOVA for proportion fully tagged by sample
labelling_anova <- labelling_summary %>%
  mutate(Sample = paste(Donor, Condition)) %>%
  anova_test(prop_na ~ Sample)

# Print results
labelling_anova$p

# Create plot
Efficiency_PSMs %>%
  pivot_longer(
    cols = c("Completely tagged", "N-Term tagged", "Lysine tagged"),
    names_to = "Measure",
    values_to = "Tagged"
  ) %>%
  group_by(Fraction, Measure) %>%
  summarise(
    n = n(),
    mean = mean(Tagged, na.rm = TRUE),
    sd = sqrt(mean * (1 - mean) / n)
  ) %>%
  ggplot(aes(x = Fraction, y = mean, ymin = mean - sd, ymax = mean + sd, fill = Fraction)) +
  geom_bar(position = position_dodge(width = 0.9), stat = "identity") +
  geom_errorbar(width = 0.2, position = position_dodge(width = 0.9)) +
  geom_hline(yintercept = 0.98, col = "blue", alpha = 0.5) +
  facet_wrap("Measure") +
  scale_y_continuous(
    limits = c(0, 1),
    labels = function(x) paste0(x * 100, "%")
  ) +
  labs(
    x = "Fraction",
    y = "Percent tagged",
    title = "TMT labelling efficiency") +
  theme_bw() +
  theme(legend.position = "none")

# Save the plot
ggsave(paste0(output_dir, "Labelling efficiency.png"), width = plot_width, height = plot_height)

# Plot of reporter ion missing values by sample
labelling_summary %>%
  mutate(Sample = paste(Donor, Condition, sep = "_")) %>%
  group_by(Sample, Run) %>%
  summarise(
    prop_na = mean(prop_na)
  ) %>%
  ggplot(aes(x = Sample, y = prop_na, fill = Sample)) +
  stat_summary(fun.data = mean_sd, geom = "bar") +
  stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2) +
  stat_compare_means(
    method = "anova",
    aes(label = paste0("Anova, p = ", signif(after_stat(p), 3)))) +
  scale_y_continuous(
    labels = function(x) paste0(x * 100, "%")
  ) +
  labs(
    x = "Sample",
    y = "Proportion of reporter ion missing values",
    title = "Reporter ion missing values by sample") +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(
      angle = 90,
      vjust = 0
    )
  ) +
  scale_fill_manual(name = "Sample", values = custom_colours)

# Save the plot
ggsave(paste0(output_dir, "Reporter ion missing values by sample.png"), width = plot_width, height = plot_height)
```

## Over labelling
Note that this uses data from a different search, where TMT-tagged HSTY was included as a dynamic modification
```{r TMT over labelling}
# Make data summary
HSTY_summary <- Overlabelling_PSMs %>%
  group_by(Fraction, `Spectrum File`) %>%
  summarise(
    "Histidine" = sum(`Histidine tag count`) / sum(`Histidine count`),
    "Serine" = sum(`Serine tag count`) / sum(`Serine count`),
    "Threonine" = sum(`Threonine tag count`) / sum(`Threonine count`),
    "Tyrosine" = sum(`Tyrosine tag count`) / sum(`Tyrosine count`)
  ) %>%
  pivot_longer(
    cols = c(
      "Histidine", 
      "Serine", 
      "Threonine", 
      "Tyrosine"),
    names_to = "Modification",
    values_to = "Rate"
  )

# Get a summary of rates
HSTY_summary %>%
  group_by(Modification) %>%
  summarise(Percent = round(mean(Rate)*100, digits = 2))

# Get overall overlabeled peptides - at least one inappropriate tag
mean(
  Overlabelling_PSMs$`Histidine tag count` > 0 |
    Overlabelling_PSMs$`Serine tag count` > 0 |
    Overlabelling_PSMs$`Threonine tag count` > 0 |
    Overlabelling_PSMs$`Tyrosine tag count` > 0
)

# Make a plot
HSTY_summary %>%
  ggplot(aes(x = Fraction, y = Rate, fill = Fraction)) +
  stat_summary(fun.data = mean_sd, geom = "bar") +
  stat_summary(fun.data = mean_sd, geom = "errorbar", width = 0.2) +
  facet_wrap("Modification") +
  scale_y_continuous(
    labels = function(x) paste0(x * 100, "%")
  ) +
  labs(x = "Fraction", y = "Proportion tagged") +
  ggtitle("TMTpro over labelling") +
  theme_bw() +
  theme(legend.position = "none")

# Save the plot
ggsave(paste0(output_dir, "Over labelling.png"), width = plot_width, height = plot_height)
```

## Mixing ratios
```{r mixing ratio, include = FALSE}
# Mixing ratio before and after summary
Mixing_abundances <- Mixing_PSMs %>%
  dplyr::select(c(Mix, paste("Abundance:", channels_vector))) %>%
  pivot_longer(cols = !c(Mix), names_to = "Channel", values_to = "Abundance") %>%
  group_by(Mix, Channel) %>%
  summarise("Sum abundances" = sum(Abundance, na.rm = TRUE)) %>%
  as.data.frame() %>%
  mutate(Channel = str_remove(Channel, "Abundance: "))

# Calculate coefficient of variation by mix
Mixing_abundances %>%
  group_by(Mix) %>%
  summarise(CV = round(sd(`Sum abundances`)/mean(`Sum abundances`)*100, digits = 2))

# Create the plot
Mixing_abundances %>%
  ggplot(aes(x = Channel, y = `Sum abundances`, fill = Channel)) +
  geom_col() +
  labs(
    title = "Sum of PSM channel abundances",
    x = "Channel",
    y = "Sum reporter ion abundance"
  ) +
  facet_wrap("Mix") +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )

# Save the plot standard size
ggsave(paste0(output_dir, "Mixing ratio reporter ion abundances.png"), width = plot_width, height = plot_height)

# Pivot data to long format and prepare for plotting
Channel_abundances <- PSMs %>%
  dplyr::select(c(Fraction, Batch, paste("Abundance:", channels_vector))) %>%
  pivot_longer(cols = !c(Fraction, Batch), names_to = "Channel", values_to = "Abundance") %>%
  group_by(Fraction, Batch, Channel) %>%
  summarise("Sum abundances" = sum(Abundance, na.rm = TRUE)) %>%
  as.data.frame() %>%
  mutate(Channel = str_remove(Channel, "Abundance: "))

# Create the plot
Channel_abundances %>%
  ggplot(aes(x = Channel, y = `Sum abundances`, fill = Channel)) +
  geom_col() +
  labs(
    title = "Sum of PSM channel abundances",
    x = "Channel",
    y = "Sum reporter ion abundance"
  ) +
  facet_grid(Fraction ~ Batch) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)
  )

# Save the plot standard size
ggsave(paste0(output_dir, "PSM reporter ion abundances.png"), width = plot_width, height = plot_height)
```

## Fractionation
Assess efficacy of fractionation using GRAVY score.
```{r Hydrophobicity}
# Run one-way ANOVA to check significance of fractionation
fractionation_anova <- PSMs %>%
  group_by(Fraction, Run) %>%
  summarise(`GRAVY score` = mean(`GRAVY score`), .groups = "drop") %>%
  anova_test(`GRAVY score` ~ Fraction)

# Print results
fractionation_anova$p

# Density plot of GRAVY scores
PSMs %>%
  ggplot(aes(x = `GRAVY score`, colour = Fraction)) +
  geom_density() +
  annotate("text", x = -2.5, y = 0.7, label = paste0("ANOVA, p = ", formatC(fractionation_anova$p, format = "e", digits = 2))) +
  labs(
    x = "GRAVY index (Kyte-Doolittle scoring)",
    y = NULL,
    title = "Peptide hydrophobicity by fraction"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

# Save the plot
ggsave(paste0(output_dir, "PSM hydrophobicity.png"), width = plot_width, height = plot_height)

# Run one-way ANOVA to check significance of fractionation
fractionation_RT_anova <- PSMs %>%
  group_by(Fraction, Run) %>%
  summarise(`RT [min]` = mean(`RT [min]`), .groups = "drop") %>%
  anova_test(`RT [min]` ~ Fraction)

# Print results
fractionation_RT_anova$p

# Make the plot
PSMs %>%
  ggplot(aes(x = `RT [min]`, color = Fraction)) +
  geom_density(bw = 2.5) +
  annotate("text", x = 40, y = 0.014, label = paste0("ANOVA, p = ", formatC(fractionation_RT_anova$p, format = "e", digits = 2))) +
  scale_y_continuous(
    name = "Peptide-spectrum matches",
    breaks = NULL
  ) +
  labs(x = "Retention time (minutes)") +
  ggtitle("PSMs by retention time") +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

# Save the plot
ggsave(paste0(output_dir, "PSMs over RT by fraction.png"), width = plot_width, height = plot_height)
```

## UpSet plots by fraction
Plots to check overlap of peptide identifications by fraction
```{r}
# prepare the data for UpSet plots. Rows represent peptide groups, columns represent runs. 1 and 0 are used to represent whether the protein was detected in that sample or not.

# Pivot wider to create the matrix
UpSet_matrix_peptides <- PeptideGroups %>%
  distinct(`Annotated Sequence`, Run) %>%
  mutate(present = 1) %>%
  pivot_wider(
    id_cols = `Annotated Sequence`,
    names_from = Run,
    values_from = present,
    values_fill = 0
  )

# Set up blank matrix for accumulated farction data
fraction_names <- paste("Fraction", unique(Run_annotations$Fraction))
fraction_matrix <- matrix(NA, nrow = nrow(UpSet_matrix_peptides), ncol = length(fraction_names) + 1) %>%
    as.data.frame()
colnames(fraction_matrix) <- c("Annotated Sequence", fraction_names)

# Fill in protein names
fraction_matrix$`Annotated Sequence` <- UpSet_matrix_peptides$`Annotated Sequence`

# Loop to summarise fractions
for(i in fraction_names) {
    # Get list of runs for the fraction
    runs <- filter(Run_annotations, Fraction == str_remove(i, "Fraction "))$Run %>% as.character
    # Check if identified in at least one run and add to data
    fraction_matrix[,i] <- as.numeric(rowSums(UpSet_matrix_peptides[,runs]) > 0)
}

# Count overlap fractions
Fractions_identified <- rowSums(fraction_matrix[,fraction_names])
sum(Fractions_identified == 8) / length(Fractions_identified)
sum(Fractions_identified == 1) / length(Fractions_identified)

# make the plot
upset(
    fraction_matrix,
    fraction_names,
    name = "Fraction intersections",
    width_ratio = 0.1,
    height_ratio = 0.3,
    n_intersections = 15,
    # base_annotations = list(
    #     "Peptide group identifications" = intersection_size(text_mapping = aes(label = !!upset_text_percentage()))
    # ),
    min_degree = 1
) +
  labs(
    title = "Peptide group identification numbers by fraction"
  )

# Save the plot
ggsave(paste0(output_dir, "Fraction overlap UpSet plot.png"), width = plot_width * 2, height = plot_height * 2)
```

## Chromatograms
```{r Load mzML files and create chromatograms}
if(run_chromatograms) {
  # Make the plot TIC chromatograms
  chrom_data %>%
    ggplot(aes(x = RT, y = TIC, col = Fraction)) +
    geom_line() +
    scale_x_discrete(
    position = "bottom",
    sec.axis = dup_axis(name = "Batch", labels = NULL)
  ) +
    scale_y_discrete(
    position = "left",
    sec.axis = dup_axis(name = "Fraction", labels = NULL)
  ) +
    labs(
      title = "Total ion current chromatograms for TMT runs",
      x = "Retention time",
      y = "Total ion current intensity"
    ) +
    facet_grid(Fraction ~ Batch, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "none")
  
  # Save the plot standard size
  ggsave(paste0(output_dir, "TIC TMT chromatograms.png"), width = plot_width, height = plot_height)
  
  # Make the plot BP chromatograms
  chrom_data %>%
    ggplot(aes(x = RT, y = BP, col = Fraction)) +
    geom_line() +
    scale_x_discrete(
    position = "bottom",
    sec.axis = dup_axis(name = "Batch", labels = NULL)
  ) +
    scale_y_discrete(
    position = "left",
    sec.axis = dup_axis(name = "Fraction", labels = NULL)
  ) +
    labs(
      title = "Base peak chromatograms for TMT runs",
      x = "Retention time",
      y = "Base peak intensity"
    ) +
    facet_grid(Fraction ~ Batch, scales = "free_y") +
    theme_bw() +
    theme(legend.position = "none")
  
  # Save the plot standard size
  ggsave(paste0(output_dir, "BP TMT chromatograms.png"), width = plot_width, height = plot_height)
}
```

## PSMs By RT
Plot of peptide-spectrum matches by retention time. Coloured by run or condition
```{r PSMs by retention time plot, echo = FALSE}
# Make the plot by sample with facet grid by batch/sample
PSMs %>%
  ggplot(aes(x = `RT [min]`, color = Fraction)) +
  geom_density() +
  scale_y_continuous(
    name = "Peptide-spectrum matches",
    breaks = NULL,
    position = "left",
    sec.axis = dup_axis(name = "Fraction", labels = NULL)
  ) +
  scale_x_discrete(
    position = "bottom",
    sec.axis = dup_axis(name = "Batch", labels = NULL)
  ) +
  facet_grid(Fraction ~ Batch, scales = "free_y") +
  labs(
    title = "PSMs by retention time for TMT runs",
    x = "Retention time",
    y = "PSM identifications"
  ) +
  theme_bw() +
  theme(
    legend.position = "none")

# Save the plot standard size
ggsave(paste0(output_dir, "TMT PSMs by RT.png"), width = plot_width, height = plot_height)
```

## Peak widths
My approach here is to get some extracted ion chromatograms and then measure peak width using full width half height.
LFQ_peak_width_selection.R script used to determine the peptide to use.
Use the peptide "[RK].sEVTDLRR.[TA]", RT: ~ 41.5, m/z: ~427.2478
```{r}
# List raw files
raw_files <- list.files(input_dir, full.names = TRUE) %>%
  grep(pattern = ".raw$", value = TRUE)

# Initialise vector for peak widths
peak_widths <- numeric(length = length(raw_files))
names(peak_widths) <- raw_files

# Loop through fractions
for(fraction in unique(PSMs$Fraction)) {
  # Set list of raw files for fractions
  raw_files <- paste0(
    input_dir,
    Run_annotations$Run[Run_annotations$Fraction == fraction],
    ".raw"
  )
  
  # Assign variables by fraction
  peak_mass <- selected_peptides[[fraction]]$top_mass
  top_RT <- selected_peptides[[fraction]]$top_RT
  
  # Loop through_files
  for(i in raw_files) {
    peak_width <- calculate_peak_width(
      raw_file = i,
      peak_mass = peak_mass,
      tol = 5,
      min_RT = top_RT - 5,
      max_RT = top_RT + 5
    )
    
    # Assign to vector
    peak_widths[i] <- peak_width * 60
  }
  
}

# Merge with annotations
peak_widths <- merge(
  mutate(
    Run_annotations,
    index = paste0(input_dir, "/", Run, ".raw")
  ),
  data.frame(
    Peak_width = peak_widths,
    index = names(peak_widths)
  ),
  by = "index"
)

# Run some stats
# Run ANOVA for peak width by sample
peak_widths %>%
  anova_test(Peak_width ~ Sample) %>%
  .$p

# Run ANOVA for peak width by condition
peak_widths %>%
  anova_test(Peak_width ~ Condition) %>%
  .$p

# Run ANOVA for peak width by batch
peak_widths %>%
  anova_test(Peak_width ~ Batch) %>%
  .$p


# Example extracted ion chromatogram
XIC <- rawrr::readChromatogram(rawfile = raw_files[1], mass = 427.2478, tol = 5, type = "xic")

# Extract to a data.frame and make plot
data.frame(
  RT = as.numeric(XIC[[1]]$times),
  Intensity = XIC[[1]]$intensities
) %>%
  ggplot(aes(x = RT, y = Intensity)) +
  geom_line() +
  labs(
    title = "Extracted-ion chromatogram m/z: 427.2478",
    x = "Retention time"
  ) +
  theme_bw() +
  theme(legend.position = "none")

# Save the plot standard size
ggsave(paste0(output_dir, "TMT XIC.png"), width = plot_width, height = plot_height)

# Box plot of peak width by batch
peak_widths %>%
  ggplot(aes(x = Batch, y = Peak_width, fill = Batch)) +
  geom_boxplot() +
  labs(
    title = "Full width at half maximum (FWHM) peak width for label free quantification (LFQ) runs (VATVSLPR)",
    x = "Batch",
    y = "Peak width (seconds)"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "TMT_peak_width_boxplot.png"), width = plot_width, height = plot_height)
```

## RT analysis
Decided to run analysis as follows:
1. Group PSMs by annotated sequence - get global mean RTs
2. Group PSMs by run and annotated sequence, get run specific RTs
3. Merge
4. Subtract global means from run specific means and look at distribution
```{r RT analysis}
# Get global RT means
glob_RT_means <- PSMs %>%
  group_by(`Annotated Sequence`) %>%
  summarise(glob_RT = mean(`RT [min]`, na.rm = TRUE), .groups = "drop")

# Get run specific means and mean differences
run_RT_means <- PSMs %>%
  group_by(Run, `Annotated Sequence`) %>%
  summarise(run_RT = mean(`RT [min]`, na.rm = TRUE), .groups = "drop") %>%
  merge(glob_RT_means, by = "Annotated Sequence", all.x = TRUE) %>%
  mutate(RT_dif = run_RT - glob_RT) %>%
  merge(Run_annotations, all.x = TRUE)

# Run ANOVA for proportion for RT differences by run
RT_anova <- run_RT_means %>%
  anova_test(RT_dif ~ Run)

# Print results
RT_anova$p

# Print mean absolute retention time differences
mean(abs(run_RT_means$RT_dif))

# Print proportion of absolute delta RT less than 5 minutes
mean(abs(run_RT_means$RT_dif) < 5)
mean(abs(run_RT_means$RT_dif) < 0.5)

# Ridgeline plot of RT differences by run
run_RT_means %>%
  filter(RT_dif < 2 & RT_dif > -2) %>%
  ggplot(aes(x = RT_dif, y = Batch, fill = Batch)) +
  geom_density_ridges() +
  facet_wrap("Fraction") +
  scale_y_discrete(limits = rev) +
  scale_x_continuous(
    position = "bottom",
    sec.axis = dup_axis(name = "Fraction", labels = NULL)
  ) +
  labs(
    title = "Peptide retention time differences by run",
    x = "Peptide retention time difference relative to overall mean (minutes)",
    y = "Density"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x.top = element_blank()
  )

# Save the plot
ggsave(paste0(output_dir, "TMT RT difference ridgeline plot.png"), width = plot_width, height = plot_height * 2)
```

## Protein identifications by Sample
```{r Protein identifications by key}
# Summarise counts from PSM data
ID_counts <- PSMs %>%
  dplyr::select(c(Run, `Master Protein Accessions`)) %>%
  unique() %>%
  group_by(Run) %>%
  summarise(Protein_count = n()) %>%
  merge(Run_annotations, by = "Run")

# Run stats and averages
ID_test <- ID_counts %>%
  anova_test(Protein_count ~ Batch)
ID_test$p
round(mean(ID_counts$Protein_count))

# Make a plot
ID_counts %>%
  ggplot(aes(x = Fraction, y = Protein_count, col = Batch)) +
  geom_point() +
  annotate("text", x = 2, y = 4000, label = paste0("ANOVA, p = ", round(ID_test$p, 3))) +
  expand_limits(y = 0) +
  ggtitle("Identifications by run") +
  theme_bw()

# Save the plot
ggsave(paste0(output_dir, "TMT IDs by run.png"), width = plot_width, height = plot_height)
```

## Mass error
Plot of mass error
```{r precursor mass error, echo = FALSE}
# Run ANOVA for proportion for mass error by batch
# Summarise data - arbitrary significance from large sample size
mass_er_summary <- PSMs %>%
  group_by(Run, Batch) %>%
  summarise(
   mz = mean(`Deltam/z [Da]`),
   ppm = mean(`DeltaM [ppm]`),
   .groups = "drop"
  )

# Run ANOVA for m/z
mz_anova <- mass_er_summary %>%
  anova_test(mz ~ Batch)
mz_anova$p

# Run ANOVA for ppm
ppm_anova <- mass_er_summary %>%
  anova_test(ppm ~ Batch)
ppm_anova$p

# Summaries
mean(PSMs$`Deltam/z [Da]`)
sd(PSMs$`Deltam/z [Da]`)
mean(PSMs$`DeltaM [ppm]`)
sd(PSMs$`DeltaM [ppm]`)

# Mean absolute errors
mean(abs(PSMs$`Deltam/z [Da]`))
mean(abs(PSMs$`DeltaM [ppm]`))
mean(abs(PSMs$`DeltaM [ppm]`) < 1)

# Density plot of precursor mass error - PSM level
PSMs %>%
  ggplot(aes(x = `Deltam/z [Da]`, colour = Batch)) +
  geom_density(alpha = 0.2) +
  scale_x_continuous(
    position = "bottom",
    sec.axis = dup_axis(name = "Fraction", labels = NULL)) +
  labs(
    title = "Mass error - PSM level by run, TMT data",
    x = expression(~Delta ~ "m/z (Da)"),
    y = "Density") +
  facet_wrap("Fraction", scales = "free_y") +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x.top = element_blank(),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )

# Save the plot
ggsave(paste0(output_dir, "TMT PSM mass error by run.png"), width = plot_width, height = plot_height)

# Density plot of precursor mass error - PSM level - by Batch - ppm
PSMs %>%
  ggplot(aes(x = `DeltaM [ppm]`, colour = Batch)) +
  geom_density(alpha = 0.2) +
  scale_x_continuous(
    position = "bottom",
    sec.axis = dup_axis(name = "Fraction", labels = NULL)) +
  labs(
    title = "Mass error - PSM level by run, TMT data",
    x = expression(~Delta ~ "m/z (ppm)"),
    y = "Density") +
  facet_wrap("Fraction", scales = "free_y") +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x.top = element_blank()
  )

# Save the plot
ggsave(paste0(output_dir, "TMT PSM mass error by run ppm.png"), width = plot_width, height = plot_height)
```

## UpSet plots - replication
Use UpSet plots to look at set overlap in TMT replicates, too many for a venn diagram
```{r UpSet plot, include = FALSE}
# prepare the data for UpSet plots. Rows represent proteins, columns represent runs. 1 and 0 are used to represent whether the protein was detected in that sample or not.

# Pivot wider to create the matrix
UpSet_matrix <- Proteins %>%
  mutate(present = 1) %>%
  pivot_wider(
    id_cols = Accession,
    names_from = Run,
    values_from = present,
    values_fill = 0
  )

# Make summaries of identifications by run
fraction_overlaps <- data.frame()

# Loop through fractions
for(fraction in unique(Run_annotations$Fraction)) {
  # Make a list of runs to include
  selected_runs <- Run_annotations %>%
    filter(Fraction == fraction) %>%
    pull(Run) %>%
    as.character()
  
  # Filter the matrix
  filtered_mat <- UpSet_matrix[,selected_runs]
  
  # Calculate proportion overlap in all runs
  prop_all_runs <- mean(rowSums(filtered_mat) == length(selected_runs))
  
  # Add to the data
  fraction_overlaps <- bind_rows(
    fraction_overlaps,
    data.frame(
      Fraction = fraction,
      n_runs = length(selected_runs),
      prop_all_runs = prop_all_runs
    )
  )
}

# Present the range
range(fraction_overlaps$prop_all_runs)

# Make the plot per fraction
for (i in unique(Run_annotations$Fraction)) {
  # Make a list of runs to include
  selected_runs <- Run_annotations %>%
    filter(Fraction == i) %>%
    pull(Run) %>%
    as.character()

  # make the plot
  upset(
    UpSet_matrix,
    selected_runs,
    name = "Run intersections",
    width_ratio = 0.1,
    height_ratio = 0.3,
    n_intersections = 15,
    base_annotations = list(
      "Protein identifications" = intersection_size(text_mapping = aes(label = !!upset_text_percentage()))
    ),
    min_degree = 1
  )

  # Save the plot
  ggsave(paste0(output_dir, "Fraction ", i, " UpSet plot.png"), width = plot_width * 2, height = plot_height * 2)
}
```

## Missing values
Rather than reporting a heatmap of runs (expected lots of missing values because of fractions), report proportion with missing reporter ions values
```{r}
# Create the heatmap
# Create the presence/absence matrix
heatmap_data <- MSstatsTMT_processed$ProteinLevelData %>%
  filter(!(Channel %in% c("126", "135N"))) %>%
  merge(
    mutate(Channel_annotations, Sample = paste(Donor, Condition)),
    by = "Channel") %>%
  dplyr::select(Sample, Protein) %>%
  distinct(Sample, Protein) %>%
  mutate(Present = TRUE) %>%
  complete(Sample, Protein, fill = list(Present = FALSE))

# Convert to wide format for clustering
wide_matrix <- heatmap_data %>%
  pivot_wider(names_from = Sample, values_from = Present, values_fill = FALSE) %>%
  tibble::column_to_rownames("Protein") %>%
  as.matrix()

# Perform hierarchical clustering on proteins (rows)
protein_dist <- dist(wide_matrix * 1, method = "binary")  # binary distance for TRUE/FALSE
protein_hclust <- hclust(protein_dist)
protein_order <- rownames(wide_matrix)[protein_hclust$order]

# Get number of missing values
signif(mean(!heatmap_data$Present) * 100, digits = 2)

# Plot with ordered proteins
heatmap_data %>%
  mutate(Protein = factor(Protein, levels = protein_order)) %>%
  ggplot(aes(x = Sample, y = Protein, fill = Present)) +
  geom_tile() +
  scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white")) +
  theme_bw() +
  labs(
    x = "Sample",
    y = "Protein",
    fill = "Identified",
    title = "TMT reporter ion missing values") +
  theme(
    axis.text.x = element_text(angle = 90),
    axis.text.y = element_blank(),
    panel.grid = element_blank(),
    axis.ticks.y = element_blank(),
    axis.ticks.x = element_blank()
  )

# Save the plot
ggsave(paste0(output_dir, "TMT missing values heatmap.png"), width = plot_width, height = plot_height)
```

## Normalisation
```{r Normalisation}
# Protein level normalisation plots
# TMT run intensity plot
rbind(
  mutate(
    MSstatsTMT_processed$ProteinLevelData,
    Normalisation = "Normalised"
  ),
  mutate(
    MSstatsTMT_processed_no_normalisation$ProteinLevelData,
    Normalisation = "Original"
  )
) %>%
  ggplot(aes(x = Run, y = Abundance, fill = Run)) +
  geom_boxplot() +
  stat_compare_means(
    method = "anova",
    aes(label = paste0("Anova, p = ", signif(after_stat(p), 3)))) +
  facet_wrap("Normalisation") +
  labs(
    x = "Run",
    y = "Protein-level abundance",
    title = "Protein level TMT protein abundances by run",
    fill = "Run"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "TMT protein level run intensities.png"), width = plot_width, height = plot_height)
```

## Precursor charge
```{r Precursor charge}
# Format the data
charge_data <- PSMs %>%
  mutate(Charge = as.factor(Charge)) %>%
  group_by(Run, Charge) %>%
  summarise(count = n()) %>%
  group_by(Charge) %>%
  summarise(Proportion = sum(count), SD = sd(count)) %>%
  mutate(
    Proportion = Proportion / nrow(PSMs),
    SD = SD / nrow(PSMs))

# Add empty row for charge = 1
charge_data <- bind_rows(
  tibble(Charge = factor("1"), Proportion = 0, SD = 0),
  charge_data
) %>%
  mutate(Charge = factor(Charge, levels = sort(as.numeric(levels(Charge)))))

# Make the plot
charge_data %>%
  ggplot(aes(x = Charge, y = Proportion, ymin = Proportion - SD, ymax = Proportion + SD, fill = Charge)) +
  geom_col() +
  geom_errorbar(data = . %>% filter(Charge != 1), width = 0.2) +
  scale_y_continuous(
    limits = c(0, 0.7),
    labels = function(x) paste0(x * 100, "%")
  ) +
  labs(
    x = "Precursor charge - label free quantification runs",
    y = "Proportion of PSMs",
    title = "Precursor charge distribution"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "TMT precursor charge.png"), width = plot_width, height = plot_height)
```

## Peptide length
```{r}
# Summarise data
peptide_lengths <- PSMs %>%
  mutate(
    Length = as.factor(nchar(gsub("^\\[.*?\\].|.\\[.*?\\]$", "", `Annotated Sequence`)))) %>%
  group_by(Run, Length) %>%
  summarise(count = n()) %>%
  group_by(Length) %>%
  summarise(Proportion = sum(count), SD = sd(count)) %>%
  mutate(Proportion = Proportion / nrow(PSMs), SD = SD / nrow(PSMs))

# Present most common length
peptide_lengths[which.max(peptide_lengths$Proportion),]
sum(as.numeric(peptide_lengths$Length) * peptide_lengths$Proportion, na.rm = TRUE)

# Make a plot
peptide_lengths %>%
  ggplot(aes(x = Length, y = Proportion, ymin = Proportion - SD, ymax = Proportion + SD, fill = Length)) +
  geom_col() +
  geom_errorbar(width = 0.2) +
  scale_y_continuous(
    labels = function(x) paste0(x * 100, "%")
  ) +
  labs(
    x = "Peptide length",
    y = "Proportion of PSMs",
    title = "Peptide length - TMT runs"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "TMT peptide lengths.png"), width = plot_width, height = plot_height)
```

## Unique peptides
```{r}
# Get some stats
mean(Proteins$`# Unique Peptides` == 0) * 100
mean(Proteins$`# Unique Peptides` == 1) * 100

# Summarise the data
unique_peptides <- Proteins %>%
  group_by(Run, `# Unique Peptides`) %>%
  mutate(`# Unique Peptides` = as.factor(
    ifelse(as.numeric(`# Unique Peptides`) >= 20, "20+", as.character(`# Unique Peptides`))
  )) %>%
  summarise(count = n()) %>%
  group_by(`# Unique Peptides`) %>%
  summarise(Proportion = sum(count), SD = sd(count)) %>%
  mutate(Proportion = Proportion / nrow(Proteins), SD = SD / nrow(Proteins))

# Make a plot
unique_peptides %>%
  ggplot(aes(x = factor(`# Unique Peptides`, levels = c(as.character(0:19), "20+")), y = Proportion, ymin = Proportion - SD, ymax = Proportion + SD, fill = `# Unique Peptides`)) +
  geom_col() +
  geom_errorbar(width = 0.2) +
  scale_y_continuous(
    labels = function(x) paste0(x * 100, "%")
  ) +
  labs(
    x = "Number of unique peptides per protein",
    y = "Proportion of identified proteins",
    title = "Protein unique peptide counts - TMT runs"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "TMT unique peptides.png"), width = plot_width, height = plot_height)
```

## Protein coverage
```{r}
mean(Proteins$`Coverage [%]`)
mean(Proteins$`Coverage [%]` > 3.74)

# Density plot of protein coverage
Proteins %>%
  ggplot(aes(x = `Coverage [%]`)) +
  geom_density() +
  labs(
    x = "Protein sequence coverage (%)",
    y = "Density",
    title = "Protein sequence coverage - TMT runs"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

# Save the plot
ggsave(paste0(output_dir, "TMT protein coverage.png"), width = plot_width, height = plot_height)
```

## Sequest score
```{r}
mean(Proteins$`Score Sequest HT: Sequest HT`)

# Density plot of protein coverage
Proteins %>%
  ggplot(aes(x = `Score Sequest HT: Sequest HT`)) +
  geom_density() +
  scale_x_continuous(trans = "log2") +
  labs(
    x = "Sequest HT score (log 2 scale)",
    y = "Density",
    title = "Sequest HT scores - TMT runs"
  ) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  )

# Save the plot
ggsave(paste0(output_dir, "TMT SEQUEST scores.png"), width = plot_width, height = plot_height)
```
