# Script to process images and analyse them for apical fluid volume
# Sam Siljee
# 6th September 2024

---
title: "Apical fluid analysis"
author: "Sam Siljee"
date: '2022-09-06'
output: html_document
---

Note that I've had to change two intensity threshold values for the pre-wash 4x data. These values caused convergence to fail when fitting the sigmoid function to the fail to converge.
The values I've changed are
21-052_1_270 from 297 to 273
and
22-069_6_270 from 362 to 337

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Packages
library(dplyr) # Data manipulation
library(ggplot2) # Plotting
library(stringr) # String manipulation in annotations
library(tiff) # Read in .tiff files
library(png) # Write .png images
library(drc) # For 4PL model

# Files and directories
input_dir <- paste0(getwd(), "/data/")
output_dir <- paste0(getwd(), "/results/")

# User variables
min_offset <- -90
max_offset <-5
magnification <- "4x" # "10X"
wash <- "pre" # "pre wash"

# Load custom functions
source("Apical_fluid_analysis_functions.R")
```

# Apical fluid analysis
Run main loop processing the data, iterating over each image
```{r Run processing}
# List input files filtering using the variables from setup
input_files <- list.files(input_dir) %>%
  grep(pattern = ".tif$", value = TRUE) %>%
  grep(pattern = magnification, value = TRUE) %>%
  grep(pattern = wash, value = TRUE)

# Create annotations from file names
annotations <- data.frame(file_name = input_files) %>%
  mutate(
    Donor = substr(file_name, start = 1, stop = 6),
    Well = str_extract(file_name, "(?<=well )\\d(?= )"),
    Rotation = str_extract(file_name, "(?<= wash )\\d{2,3}(?= well )"),
    ID = paste(Donor, Well, Rotation, sep = "_"),
    Condition = ifelse(Well %in% c(1, 2, 3), "Control", "p53 KD")
  )

# Read in intensity curve cutoffs
intensity_cutoffs <- read.csv(paste0(input_dir, "curve_cutoffs_", magnification, "_", wash,"_wash.csv"), header = TRUE)

# Initialise list for results
results <- list()

# Start loop through images
for (i in 1:nrow(annotations)) {
  # Set some variables
  file_name <- annotations$file_name[i]
  ID <- annotations$ID[i]
  
  # Get manual intensity cutoff value
  intensity_cutoff <- intensity_cutoffs[intensity_cutoffs$ID == ID, 2]
  
  # Print progress check
  print(paste0("Starting on ", ID, ", ", i, " of ", nrow(annotations)))

  # Load the image
  image_matrix <- readTIFF(paste0(input_dir, file_name), as.is = TRUE)

  # Convert to grey-scale image - sum all channels
  image_matrix <- image_matrix[, , 1] + image_matrix[, , 2] + image_matrix[, , 3]
  
  # Get number of rows for other functions
  image_height <- nrow(image_matrix)

  # Flip on vertical axis if image taken at 90 degrees position
  if (annotations$Rotation[i] == 90) {
    image_matrix <- image_matrix[, ncol(image_matrix):1]
  }

  # Take the average of the middle fifth as a reference row
  reference_row <- colMeans(image_matrix[floor(nrow(image_matrix) * 2 / 5):ceiling(nrow(image_matrix) * 3 / 5), ])

  # Run function to estimate best offsets
  offset_estimates <- get_alignment_offsets(image_matrix, reference_row, min_offset, max_offset)
  
  # Run quadratic model to estimate offset curve
  offset_model <- lm(offset ~ poly(row, 2), data = data.frame(row = 1:image_height, offset = offset_estimates))
  
  # Model the offsets
  offsets <- predict(offset_model, newdata = data.frame(row = 1:image_height))
  
  # Plot the offsets
  rbind(
    data.frame(
      Row = 1:image_height,
      Offset = offset_estimates,
      Method = "Image alignment"),
    data.frame(
      Row = 1:image_height,
      Offset = offsets,
      Method = "Quadratic model")
  ) %>%
    ggplot(aes(x = Row, y = Offset, col = Method)) +
    geom_line() +
    coord_flip() +
    labs(
      title = paste("Calcualted offset values for image ", ID),
      x = "Image row",
      y = "Row offset value"
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
    )
  
  # Save the plot
  ggsave(paste0(output_dir, "offset_plots/", ID, "_offset_plot.png"), width = 8, height = 5.5)

  # Align rows, saving an image of the aligned matrix and make a column means vector
  aligned_matrix <- align_matrix(image_matrix, -offsets)
  
  # Make an image with a red line for the intensity cutoff
  aligned_image <- array(dim = c(dim(aligned_matrix), 3))
  aligned_image[,,1:3] <- normalise_image(aligned_matrix) # All three channels the same
  # Add in red line
  aligned_image[,(intensity_cutoff - 2) : (intensity_cutoff + 2),1] <- 1
  aligned_image[,(intensity_cutoff - 2) : (intensity_cutoff + 2),2:3] <- 0
  
  # Save as a PNG image - grayscale
  writePNG(aligned_image, paste0(output_dir, "aligned_images/", ID, "_aligned_image.png"))
  
  # Get intensity values
  intensity_vector <- colMeans(aligned_matrix)
  
  # Trim by manual cutoff
  trimmed_intensity <- intensity_vector[intensity_cutoff:length(intensity_vector)]
  
  # Get intensity values for model - including exceptions for pre-wash 4x samples 21-052_1_270 and 22-069_6_270
  model_trimmed_intensity <- if (magnification == "10x" | wash == "post") {
    trimmed_intensity
  } else if (ID == "21-052_1_270") {
    intensity_vector[273:length(intensity_vector)]
  } else if (ID == "22-069_6_270") {
    intensity_vector[337:length(intensity_vector)]
  } else {
    trimmed_intensity
  }

  # Fit 4PL curve model with data trimmed by manual intensity cutoff
  model_data <- data.frame(
    Intensity = max(model_trimmed_intensity) - model_trimmed_intensity,
    Radial_distance = 1:length(model_trimmed_intensity)
    )
  
  fit <- drm(Intensity ~ Radial_distance, data = model_data, fct = LL.3())
  
  # Predict concentrations of samples
  predicted_fit <- predict(fit, newdata = data.frame(Radial_distance = 1:length(trimmed_intensity)))
  
  # Calculate intensity cutoff as per Harvey et. al. method - where intensity is at its maximum.
  Max_intensity_cutoff <- which.min(intensity_vector)
  
  # Make a plot of the intensity values
  data.frame(
    Radial_distance = 1:length(intensity_vector),
    Intensity = intensity_vector,
    Model = c(rep(NA, intensity_cutoff - 1), max(trimmed_intensity) - predicted_fit)
  ) %>%
    ggplot(aes(x = Radial_distance, y = Intensity)) +
    geom_line() +
    geom_line(aes(x = Radial_distance, y = Model), col = "darkgreen") +
    geom_vline(xintercept = intensity_cutoff, col = "red") +
    geom_vline(xintercept = Max_intensity_cutoff, col = "blue") +
    scale_y_reverse() +
    labs(
      title = paste("Aligned column mean intensity:", ID),
      x = "Radial distance (pixels)",
      y = "Mean intensity"
    ) +
    theme_bw() +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
      legend.position = "none"
    )
  
  # Save the plot
  ggsave(paste0(output_dir, "intensity_plots/", ID, "_intensity_plot.png"), width = 8, height = 5.5)

  # calculate area under the curve according to the 4PL model using numerical integration
  sigmoid_func <- function(x) {
    predict(fit, newdata = data.frame(Radial_distance = x))
  }
  
  # Calculate area under curve
  sigmoid_area <- integrate(sigmoid_func, lower = 0, upper = 2000)
  
  # Calculate the area under the curve by adding the intensity values - inverted for greyscale mapping
  sum_area <- sum(max(trimmed_intensity) - trimmed_intensity)
  
  # Add to results list
  results[[ID]] <- list(
    "sum_area" = sum_area,
    "sigmoid_area" = sigmoid_area$value,
    "sigmoid_model" = fit,
    "quadratic_model" = offset_model
  )
}

# Save the resulting area data
save(results, file = paste0(output_dir, magnification, "_", wash, "_results.rda"))
```
