---
title: "MS phosphoproteomics QC (Proteome Discoverer)"
author: "Sam Siljee"
output:
  pdf_document: default
  html_document: default
---

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr) # Data wrangling
library(tidyr) # For `pivot_longer` in labelling efficiency graph
library(stringr) # For annotating labeling efficiency and channel abundances
library(ggplot2) # Plotting
library(vroom) # Fast reading of data
library(ggVennDiagram) # New package to make venn diagrams
library(rawrr) # For loading raw data for chromatograms
library(Peptides) # for calculating GRAVY scores

# Set some variables
load_from_rda <- TRUE
run_chromatograms <- FALSE
plot_width <- 8
plot_height <- 5.5
input_dir <- paste0(getwd(), "/data/Phos/")
output_dir <- paste0(getwd(), "/../produced_plots/MS_QC/Phos/")
channels_vector <- c(
  "126",
  paste0(rep(127:134, each = 2), c("N", "C")),
  "135N"
)
```

## Load the data
```{r load data, include = FALSE}
# Load custom functions
source(paste0(input_dir, "../../../Functions.R"))

# Load the data
if(load_from_rda) {
  # load files from .rda files for faster loading
  rda_files <- c(
    "Run_annotations.rda",
    "MSMSSpectrumInfo.rda",
    "ProteinGroups.rda",
    "PSMs.rda",
    "QuanSpectra.rda",
    "PeptideGroups.rda",
    "Proteins.rda",
    "ModificationSites.rda"
  )
  for (i in rda_files) {
    load(paste0(input_dir, i))
  }
  
  # Load chromatogram data if required
  if(run_chromatograms) {
    load(paste0(input_dir, "chrom_data.rda"))
  }
} else {
  # Load PD LFQ output from txt files
  files_list <- list.files(input_dir, full.names = TRUE)
  
  # Load run summary to use as annotations file, add "Path" column to merge with other datasets
  Run_annotations <- read.delim(paste0(input_dir, "run_annotations.tsv"), header = TRUE) %>%
    # Convert some columns to factors for faster computation
    mutate(
      across(
        c(
          Batch,
          Run
        ),
        as.factor
      )
    )
  save(Run_annotations, file = paste0(input_dir, "Run_annotations.rda"))
  
  # Load different types of data
  MSMSSpectrumInfo <- import_txt_data("MSMSSpectrumInfo", input_dir, files_list, Run_annotations)
  save(MSMSSpectrumInfo, file = paste0(input_dir, "MSMSSpectrumInfo.rda"))
  
  ProteinGroups <- import_txt_data("ProteinGroups", input_dir, files_list, Run_annotations)
  save(ProteinGroups, file = paste0(input_dir, "ProteinGroups.rda"))
  
  PSMs <- import_txt_data("PSMs", input_dir, files_list, Run_annotations)
  save(PSMs, file = paste0(input_dir, "PSMs.rda"))
  
  QuanSpectra <- import_txt_data("QuanSpectra", input_dir, files_list, Run_annotations)
  save(QuanSpectra, file = paste0(input_dir, "QuanSpectra.rda"))
  
  PeptideGroups <- import_txt_data("PeptideGroups", input_dir, files_list, Run_annotations)
  save(PeptideGroups, file = paste0(input_dir, "PeptideGroups.rda"))
  
  Proteins <- import_txt_data("Proteins", input_dir, files_list, Run_annotations)
  save(Proteins, file = paste0(input_dir, "Proteins.rda"))
  
  ModificationSites <- import_txt_data("ModificationSites", input_dir, files_list, Run_annotations)
  save(ModificationSites, file = paste0(input_dir, "ModificationSites.rda"))
  
  # Load data if running chromatograms
  if (run_chromatograms) {
    # Load chromatogram data
    raw_files <- list.files(input_dir, full.names = TRUE) %>%
      grep(pattern = ".raw$", value = TRUE)
    
    # Initialise data.frame
    chrom_data <- data.frame()
    
    # Read in raw files and extract TIC values to data.frame
    for (i in raw_files) {
      # Open file
      tic <- rawrr::readChromatogram(rawfile = i, type = "tic")
      bpc <- rawrr::readChromatogram(rawfile = i, type = "bpc")
      
      chrom_data <- rbind(
        chrom_data,
        data.frame(
          TIC = tic$intensities,
          BP = bpc$intensities,
          RT = as.numeric(tic$times),
          raw_path = i
        )
      )
    }
    
    # Merge with run overview data for annotations
    chrom_data <- merge(
      chrom_data,
      mutate(
        Run_annotations,
        # Add column to match up to protein groups files
        raw_path = paste0(input_dir, "/", Run, ".raw")
      ),
      by = "raw_path",
      all.x = TRUE
    )
    
    # Save as R data for faster laoding in future
    save(chrom_data, file = paste0(input_dir, "chrom_data.rda"))
  }
}

# Add a column to describe fully or partially tagged peptides
PSMs <- PSMs %>%
  mutate(
    # Create a trimmed sequence to count lysines - ignoring lysines before and after
    `Trimmed sequence` = gsub("^\\[.*?\\].|.\\[.*?\\]$", "", `Annotated Sequence`),
    `Lysine count` = str_count(tolower(`Trimmed sequence`), "k"),
    `Lysine tag count` = str_count(Modifications, "K\\d{1,2}\\(TMTpro\\)"),
    `N-Term tagged` = grepl("N-Term(TMTpro)", Modifications, fixed = TRUE),
    `Lysine tagged` = ifelse(
      `Lysine count` != 0, # Only count sequences containing lysine
      ifelse(
        `Lysine tag count` == `Lysine count`,
        TRUE, # All lysines are tagged
        FALSE # Not all lysines are tagged
      ),
      NA
    ), # Return NA if no lysine in the sequence
    `Completely tagged` = `N-Term tagged` & (`Lysine tagged` | is.na(`Lysine tagged`)),
    "GRAVY score" = hydrophobicity(`Trimmed sequence`) # Calculate hydrophobicity
  )

# Parse phosphosite data
PSMs <- PSMs %>%
  rowwise() %>%
  mutate(parsed = list(parse_phos_confidence(`ptmRS: Best Site Probabilities`))) %>%
  unnest_wider(parsed)
```

## Chromatograms
```{r Create chromatograms}
if(run_chromatograms) {
  # Make the plot TIC chromatograms
  chrom_data %>%
    ggplot(aes(x = RT, y = TIC, col = Run)) +
    geom_line() +
    facet_wrap("Run", ncol = 1) +
    labs(
      title = "Total ion current chromatograms for phosphoproteomics runs",
      x = "Retention time"
    ) +
    theme_bw() +
    theme(legend.position = "none")
  
  # Save the plot standard size
  ggsave(paste0(output_dir, "TIC phos chromatograms.png"), width = plot_width, height = plot_height)
  
  # Make the plot BP chromatograms
  chrom_data %>%
    ggplot(aes(x = RT, y = BP, col = Run)) +
    geom_line() +
    facet_wrap("Run", ncol = 1) +
    labs(
      title = "Base peak chromatograms for phosphoproteomics runs",
      x = "Retention time"
    ) +
    theme_bw() +
    theme(legend.position = "none")
  
  # Save the plot standard size
  ggsave(paste0(output_dir, "BP phos chromatograms.png"), width = plot_width, height = plot_height)
}
```

## PSMs By RT
Plot of peptide-spectrum matches by retention time. Coloured by run or condition
```{r PSMs by retention time plot, echo = FALSE}
# Make the plot by sample with facet grid by batch/sample
PSMs %>%
  ggplot(aes(x = `RT [min]`, color = Run)) +
  geom_density() +
  facet_wrap("Run", ncol = 1) +
  labs(
    title = "PSMs by retention time for phosphoproteomics runs",
    x = "Retention time",
    y = "PSM identifications"
  ) +
  theme_bw() +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    axis.ticks = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
    )

# Save the plot standard size
ggsave(paste0(output_dir, "Phos PSMs by RT.png"), width = plot_width, height = plot_height)
```

## ID numbers
```{r}
# Summarise counts from PSM data
PSMs %>%
  dplyr::select(c(Run, `Master Protein Accessions`)) %>%
  unique() %>%
  group_by(Run) %>%
  summarise(Protein_count = n())

# Count distinct phosphosites
ModificationSites %>%
  dplyr::select(`Protein Accession`, Position) %>%
  distinct() %>%
  nrow()

# Count number of phosphosites per run
ModificationSites %>%
  dplyr::select(Run, `Protein Accession`, Position) %>%
  group_by(Run) %>%
  summarise("Phosphosites" = n())
```

## Modification numbers by protein
```{r}
# Summarise the data
Phosphosite_summary <- ModificationSites %>%
  group_by(Run, `Protein Accession`) %>%
  summarise(Phosphosites = n(), .groups = 'drop') %>% # Count number of phosphosite per protein
  group_by(Run, Phosphosites) %>%
  summarise(Count = n(), .groups = 'drop') %>% # Count phosphosites per protein
  group_by(Run) %>%
  mutate(Proportion = Count / sum(Count)) %>%
  mutate(Phosphosites = as.factor(
    ifelse(
      as.numeric(Phosphosites) >= 20,
      "20+",
      as.character(Phosphosites)))) %>%
  group_by(Phosphosites) %>%
  summarise(
    SD = sd(Proportion, na.rm = TRUE),
    Proportion = mean(Proportion),
    .groups = 'drop'
  )

# Display proportion of singly phosphorylated proteins
Phosphosite_summary %>%
  filter(Phosphosites == 1) %>%
  .$Proportion

# Make the plot
Phosphosite_summary %>%
  ggplot(aes(
    x = factor(Phosphosites, levels = c(as.character(1:19), "20+")),
    y = Proportion,
    ymin = Proportion - SD,
    ymax = Proportion + SD,
    fill = factor(Phosphosites, levels = c(as.character(1:19), "20+")))) +
  geom_col() +
  geom_errorbar(width = 0.2) +
  scale_y_continuous(
    labels = function(x) paste0(x * 100, "%")
  ) +
  labs(
    x = "Phosphosite counts by protein",
    y = "Proportion of identified proteins",
    title = "Protein phosphosite count"
  ) +
  theme_bw() +
  theme(
    legend.position = "none"
  )

# Save the plot
ggsave(paste0(output_dir, "Phosphosite counts.png"), width = plot_width, height = plot_height)
```

## Confidence scores
```{r}
# Present proportion of PSMs meeting confidence scores
mean(PSMs$probability > 75, na.rm = TRUE)

# Make the plot by sample with facet grid by batch/sample
PSMs %>%
  ggplot(aes(x = probability)) +
  geom_density(bw = 0.1) +
  scale_x_continuous(
    labels = function(x) paste0(x, "%")
  ) +
  labs(
    title = "Phosphosite confidence scores",
    x = "Site localisation confidence score (IMP-ptmRS)",
    y = "Density"
  ) +
  theme_bw()

# Save the plot standard size
ggsave(paste0(output_dir, "Phosphosite confidence scores.png"), width = plot_width, height = plot_height)
```

## Phosphorylation rate
Using data from dynamic phosphrylation rate.
```{r}
mean(grepl("Phospho", PSMs$Modifications))
```

# LEGACY CODE
Code no longer needed currently, from previous analyses

## Summarise details of the runs
```{r Run summary table}
# Count number of proteins per run
PG_summary <- ProteinGroups %>%
  group_by(Run) %>%
  summarise("Protein groups" = length(unique(`Protein Groups Protein Group ID`)))

# Count number of PSMs per run
PSM_summary <- PSMs %>%
  group_by(Run) %>%
  summarise("PSMs" = n())

# Count number of spectra per run
MSMSSpectrumInfo_summary <- MSMSSpectrumInfo %>%
  group_by(Run) %>%
  summarise("Spectra" = n())

# Count number of proteins per run
Proteins_summary <- Proteins %>%
  group_by(Run) %>%
  summarise("Proteins" = n())

# Count number of phosphosites per run
Phosphosites_summary <- ModificationSites %>%
  group_by(Run) %>%
  summarise("Phosphosites" = n())

# Combine with Run_annotations and select desired columns
Run_summary <- merge(
  PG_summary,
  PSM_summary
) %>%
  merge(MSMSSpectrumInfo_summary) %>%
  merge(Proteins_summary) %>%
  merge(Phosphosites_summary) %>%
  merge(Run_annotations) %>%
  mutate("Identification rate" = PSMs / Spectra) %>%
  dplyr::select(Run, Batch, Spectra, PSMs, `Identification rate`, `Protein groups`, Proteins, Phosphosites)

# Save as .csv file
write.csv(Run_summary, paste0(output_dir, "Run_summary.csv"))

# Summarise some results
range(Run_summary$Proteins)
mean(Run_summary$Proteins)
```

## Venn diagrams
Overview of protein overlaps between replicates of the same sample
```{r venn diagrams}
# List off unique proteins (by group description) for each fraction
# venn_list_1 <- ProteinGroups %>%
#   filter(Batch == 1) %>%
#   split(.$Fraction) %>%
#   lapply(function(df) {
#     df %>%
#       dplyr::select(`Group Description`) %>%
#       pull()
#   })
# 
# # make the plot for several combinations of fractions
# ggVennDiagram(venn_list_1[1:4]) +
#   theme_void() +
#   ggtitle("Fraction identification overlap - fractions 1 to 4") +
#   ggplot2::scale_fill_gradient(low="white",high = "red")
# 
# # Save the plot
# ggsave(paste0(output_dir, "Batch 1 fraction replication venn diagram 1 to 4.png"), width = plot_width, height = plot_height)
# 
# # make the plot for several combinations of fractions
# ggVennDiagram(venn_list_1[3:6]) +
#   theme_void() +
#   ggtitle("Fraction identification overlap - fractions 3 to 6") +
#   ggplot2::scale_fill_gradient(low="white",high = "red")
# 
# # Save the plot
# ggsave(paste0(output_dir, "Batch 1 fraction replication venn diagram 3 to 6.png"), width = plot_width, height = plot_height)
# 
# # make the plot for several combinations of fractions
# ggVennDiagram(venn_list_1[5:8]) +
#   theme_void() +
#   ggtitle("Fraction identification overlap - fractions 5 to 8") +
#   ggplot2::scale_fill_gradient(low="white",high = "red")
# 
# # Save the plot
# ggsave(paste0(output_dir, "Batch 1 fraction replication venn diagram 5 to 8.png"), width = plot_width, height = plot_height)
# 
# # Venn diagrams for fraction overlap by batch
# for (i in unique(Run_annotations$Fraction)) {
#   venn_list <- ProteinGroups %>%
#     filter(Fraction == i) %>%
#     split(.$Batch) %>%
#     lapply(function(df) {
#       df %>%
#         dplyr::select(`Group Description`) %>%
#         pull()
#     })
# 
#   # make the plot for several combinations of fractions
#   ggVennDiagram(venn_list) +
#     theme_void() +
#     ggtitle(paste("Batch identification overlap - fraction", i)) +
#     ggplot2::scale_fill_gradient(low="white",high = "red")
# 
#   # Save the plot
#   ggsave(paste0(output_dir, "Fraction ", i, " batch replication venn diagram.png"), width = plot_width, height = plot_height)
# }
# 
# # Top 3 runs Venn diagrams
# 
# # Venn diagrams for fraction overlap by top 3 batches - for fractions 1 - 6
# for (i in 1:6) {
#   venn_list <- list(
#     "1" = filter(ProteinGroups, Fraction == i, Batch == 1)$`Group Description`,
#     "3" = filter(ProteinGroups, Fraction == i, Batch == 3)$`Group Description`,
#     "4" = filter(ProteinGroups, Fraction == i, Batch == 4)$`Group Description`
#   )
# 
#   # make the plot for several combinations of fractions
#   ggVennDiagram(venn_list) +
#     theme_void() +
#     ggtitle(paste("Batch identification overlap - fraction", i)) +
#     ggplot2::scale_fill_gradient(low="white",high = "red")
# 
#   # Save the plot
#   ggsave(paste0(output_dir, "Fraction ", i, " batch replication venn diagram top 3 runs.png"), width = plot_width, height = plot_height)
# }
# 
# # Venn diagrams for fraction overlap by top 3 batches - for fraction 7
# venn_list <- list(
#   "1" = filter(ProteinGroups, Fraction == 7, Batch == 1)$`Group Description`,
#   "2" = filter(ProteinGroups, Fraction == 7, Batch == 2)$`Group Description`,
#   "4" = filter(ProteinGroups, Fraction == 7, Batch == 4)$`Group Description`
# )
# 
# # make the plot for several combinations of fractions
# ggVennDiagram(venn_list) +
#   theme_void() +
#   ggtitle("Batch identification overlap - fraction 7") +
#   ggplot2::scale_fill_gradient(low="white",high = "red")
# 
# # Save the plot
# ggsave(paste0(output_dir, "Fraction 7 batch replication venn diagram top 3 runs.png"), width = plot_width, height = plot_height)
# 
# # Venn diagrams for fraction overlap by top 3 batches - for fraction 8
# venn_list <- list(
#   "2" = filter(ProteinGroups, Fraction == 8, Batch == 2)$`Group Description`,
#   "3" = filter(ProteinGroups, Fraction == 8, Batch == 3)$`Group Description`,
#   "4" = filter(ProteinGroups, Fraction == 8, Batch == 4)$`Group Description`
# )
# 
# # make the plot for several combinations of fractions
# ggVennDiagram(venn_list) +
#   theme_void() +
#   ggtitle("Batch identification overlap - fraction 8") +
#   ggplot2::scale_fill_gradient(low="white",high = "red")
# 
# # Save the plot
# ggsave(paste0(output_dir, "Fraction 8 batch replication venn diagram top 3 runs.png"), width = plot_width, height = plot_height)
```
