---
title: "Combined transduction efficiency"
author: "Sam Siljee"
date: '2024-08-19'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Packages
library(dplyr)
library(tidyr) # For pivot_longer and separate_wider function in data import
library(ggplot2)
library(stringr)
library(rstatix) # ANOVA testing

# Files and directories
input_dir <- paste0(getwd(), "/data/")
output_dir <- paste0(getwd(), "/../produced_plots/Transduction_efficiency/")

# Read in data processed in the Data_processing script
load(paste0(input_dir, "incucyte_data.rda"))

# Correct elapsed times
incucyte_data <- incucyte_data %>%
  mutate(Elapsed = round(Elapsed + Elapsed_offset))

# Custom colour scheme for plotting
custom_colours <- read.csv(paste0(getwd(), "/../colours.csv"))$Colour
names(custom_colours) <- read.csv(paste0(getwd(), "/../colours.csv"))$Name

# Subset the data to get final ALI data
ALI_data <- incucyte_data %>%
  filter(
      Ratio < 1.5 & # Remove outliers
      Donor %in% c(
        "21-052",
        "21-073",
        "21-120",
        "21-161",
        "21-166",
        "21-189",
        "22-069",
        "22-114"
      ) &
      Well_old %in% (paste0("Image.", c(7:9, 12:14, 17:19))) # Only use central images
  ) %>%
  filter(!(Donor == "21-161" & Vector == "pLVUHshp53" & vessel_name == "SS 21_161 transduced" & Elapsed == "336")) %>% # Filter out data from replaced scans
  mutate(
    Day = round(Elapsed / 24 / 7) * 7 + 7, # Round to the nearest week, and add add a week between seeding and first scan
    Sample = paste(Donor, Vector, sep = "_")
  )

# Make summary data for stats
ALI_data_summarised <- ALI_data %>%
  group_by(Donor, Vector, Day) %>%
  summarise(Ratio_mean = mean(Ratio), .groups = "drop")
```

### Some details on the objectives and fields of view:
OBJECTIVES

 10x: Image Resolution: 1.24 microns/pixel
Image Size: 1408 x 1040 pixels
Field of View: 1.75 x 1.29 mm

 4x: Image Resolution: 2.82 microns/pixel
Image Size: 1536 x 1152 pixels
Field of View: 4.34 x 3.25 mm

## Plot dose response curves, HEK and primary
Plot to show primary cells similar to HEK 293 T17 dose responses.
```{r Create the plot}
HEK_21_097_data <- incucyte_data %>%
  filter(vessel_name %in% c("SS 21-097 transduced 268/1/Z PLL3.7/pLVUHshp53", "SS HEK293 T17 transduced 268/1/Z PLL3.7/pLVUHshp53")) %>%
  group_by(Donor, Vector, Dose) %>%
  summarise(SD = sd(Ratio), Ratio = mean(Ratio))

# Run some stats
pLL3.7_test <- t.test(
  x = HEK_21_097_data[HEK_21_097_data$Donor == "21-097" & HEK_21_097_data$Vector == "pLL3.7", ]$Ratio,
  y = HEK_21_097_data[HEK_21_097_data$Donor == "HEK293 T/17" & HEK_21_097_data$Vector == "pLL3.7", ]$Ratio,
  paired = TRUE
)
pLL3.7_test

pLVUHshp53_test <- t.test(
  x = HEK_21_097_data[HEK_21_097_data$Donor == "21-097" & HEK_21_097_data$Vector == "pLVUHshp53", ]$Ratio,
  y = HEK_21_097_data[HEK_21_097_data$Donor == "HEK293 T/17" & HEK_21_097_data$Vector == "pLVUHshp53", ]$Ratio,
  paired = TRUE
)
pLVUHshp53_test

# Create plot all points
HEK_21_097_data %>%
  ggplot(aes(
    x = Dose,
    y = Ratio,
    ymin = Ratio - SD,
    ymax = Ratio + SD,
    col = Donor
  )) +
  geom_point(size = 3) +
  geom_errorbar(width = 0.1) +
  geom_line() +
  scale_x_continuous(trans = "log2", breaks = c(500, 200, 100, 50, 20, 10, 5, 1)) +
  facet_wrap("Vector") +
  labs(
    x = "Volume lentivirus media added (log scale)",
    y = "Ratio of area GFP+",
    #title = "Transduction efficiency at 72 hours"
  ) +
  theme_bw() +
  scale_colour_manual(name = "Donor", values = custom_colours)

ggsave(paste0(output_dir, "HEK 293T17 21_097 dose response.png"), width = 8, height = 5.5)
```

## eGPF expression over time
Plot demonstrating maintenance of eGFP expression by 2D ALI cultures over the 28 day culture period. Also run ANOVA to test changes over time
```{r}
# Run the ANOVA
anova_pLL3.7 <- anova_test(
  data = ALI_data_summarised[ALI_data_summarised$Vector == "pLL3.7",],
  dv = Ratio_mean,
  wid = Donor,
  within = Day
)
anova_pLL3.7

anova_pLVUHshp53 <- anova_test(
  data = ALI_data_summarised[ALI_data_summarised$Vector == "pLVUHshp53",],
  dv = Ratio_mean,
  wid = Donor,
  within = Day
)
anova_pLVUHshp53

# Get the means
ALI_data_summarised %>%
  group_by(Vector, Day) %>%
  summarise(Ratio = mean(Ratio_mean))

# Make a plot
ALI_data %>%
  ggplot(aes(x = Donor, y = Ratio, fill = Sample)) +
  geom_boxplot() +
  facet_grid(Vector ~ Day) +
  labs(
    #title = "eGFP expression during 2D ALI culture",
    x = "Donor",
    y = "Ratio of area GFP+"
  ) +
  theme_bw() +
  scale_fill_manual(name = "Sample", values = custom_colours) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave(paste0(output_dir, "2D ALI eGFP maintenance.png"), width = 8, height = 5.5)
```
