---
title: "Incidence and mortality"
author: "Sam Siljee"
date: "2026-01-06"
output: html_document
---

Code for figure 1 in the introduction - incidence and mortality for lung cancer in New Zealand

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load libraries
library(readxl)
library(dplyr)
library(ggplot2)
library(tidyr)
library(survminer) # Kaplan-Meier plots
library(survival) # Survival analysis

# Set directories
input_dir <- paste0(getwd(), "/data/")
output_dir <- paste0(getwd(), "/../produced_plots/Incidence_mortality/")
excel_summary <- "cancer-historical-summary-1948-2017_final.xlsx"
reload_TCGA_data <- FALSE
COSMIC_file <- "Census_allFri Jan 16 03_06_32 2026.csv"
```

# Load and reshape NZ cancer stats data
```{r Load and reshape the data}
# Sheets to read in
sheets_to_read <- paste(rep(c("Regs", "Deaths"), each = 3), c("Total", "Male", "Female"), sep = "_")

# Initialise dataframe
cancer_data <- data.frame()

for (i in sheets_to_read) {
  # Read in data
  dat <- read_xlsx(paste0(input_dir, excel_summary), sheet = i, trim_ws = TRUE)

  # Get list of cancers for new column names
  cancer_names <- dat[3, ] %>%
    unlist() %>%
    na.omit()

  # Separate data by metric, convert to numeric, and add extra columns
  count_dat <- dat[dat[4, ] %in% c("Year", "No.")] %>%
    mutate_all(as.numeric) %>%
    mutate(Metric = "Count")
  rate_dat <- dat[dat[4, ] %in% c("Year", "Rate")] %>%
    mutate_all(as.numeric) %>%
    mutate(Metric = "Rate")

  # Combine datasets again
  dat <- bind_rows(count_dat, rate_dat)

  # Correct column names
  colnames(dat) <- c("Year", cancer_names, "Metric")

  # Add sheet column, filter out innapropriate rows, and add to dataset
  cancer_data <- bind_rows(
    cancer_data,
    filter(mutate(dat, Sheet = i), !is.na(Year))
  )
}

# Pivot from wide format to long format
cancer_data <- cancer_data %>%
  pivot_longer(
    col = -c("Year", "Metric", "Sheet"),
    names_to = "Cancer_type",
    values_to = "Value"
  ) %>%
  mutate(Cancer_type = as.factor(sub(" \\(C.*", "", Cancer_type)))
```

# Incidence plot
```{r Incidence plot}
cancer_data %>%
  filter(
    Sheet == "Regs_Total" &
      Year == 2017 &
      Metric == "Rate" &
      !is.na(Value) &
      Cancer_type != "All cancers"
  ) %>%
  ggplot(aes(x = reorder(Cancer_type, -Value), y = Value, fill = Cancer_type == "Lung & Trachea")) +
  geom_col() +
  labs(
    #title = "New Zealand cancer incidence rates, 2017",
    x = "Cancer type",
    y = "Age corrected cancer registration rate"
  ) +
  theme_bw() +
  scale_fill_manual(values = c("grey", "red")) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5),
    axis.title.x = element_blank(),
    legend.position = "none"
  )

# Save plot
ggsave(paste0(output_dir, "2017 Cancer incidence.png"), width = 8, height = 5.5)
```

# Mortality plot
```{r Incidence plot}
cancer_data %>%
  filter(
    Sheet == "Deaths_Total" &
      Year == 2017 &
      Metric == "Rate" &
      !is.na(Value) &
      Cancer_type != "All cancers"
  ) %>%
  ggplot(aes(x = reorder(Cancer_type, -Value), y = Value, fill = Cancer_type == "Lung & Trachea")) +
  geom_col() +
  labs(
    #title = "New Zealand cancer mortality rates, 2017",
    x = "Cancer type",
    y = "Age corrected cancer mortality rate"
  ) +
  theme_bw() +
  scale_fill_manual(values = c("grey", "red")) +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5),
    axis.title.x = element_blank(),
    legend.position = "none"
  )

# Save plot
ggsave(paste0(output_dir, "2017 Cancer mortality.png"), width = 8, height = 5.5)
```

# Load TCGA data
Code to load clinical data and TP53 expression (RNA seq) data from TCGA
```{r TCGA data}
# Load fresh form TCGA if selected, otherwise load from .rda file
if (reload_TCGA_data) {
  # Load required libraries
  library(TCGAbiolinks) # Download TCGA data
  library(SummarizedExperiment) # Manage RNA count data
  library(DESeq2) # VST of count matrix
  library(tibble) # For rownames_to_column function

  # Download clinical data for lung squamous cell carcinoma patients
  LUSC_clin <- GDCquery_clinic("TCGA-LUSC", "clinical")

  # Select appropriate columns and make new ones
  clin_data <- LUSC_clin %>%
    select(submitter_id, vital_status, days_to_last_follow_up, days_to_death) %>%
    mutate(
      deceased = ifelse(vital_status == "Alive", FALSE, TRUE),
      overall_survival = ifelse(vital_status == "Alive", days_to_last_follow_up, days_to_death)
    )

  # Query to get gene expression data
  LUSC_gene_query <- GDCquery(
    project = "TCGA-LUSC",
    data.category = "Transcriptome Profiling",
    experimental.strategy = "RNA-Seq",
    workflow.type = "STAR - Counts",
    data.type = "Gene Expression Quantification",
    sample.type = "Primary Tumor",
    access = "open"
  )

  # Download the RNA seq data
  GDCdownload(LUSC_gene_query)

  # Extract count data and prepare count matrix
  LUSC_counts <- GDCprepare(LUSC_gene_query, summarizedExperiment = TRUE)
  LUSC_matrix <- assay(LUSC_counts, "unstranded")

  # extract gene and sample metadata from summarizedExperiment object
  gene_metadata <- as.data.frame(rowData(LUSC_counts))
  coldata <- as.data.frame(colData(LUSC_counts))

  # vst transform counts to be used in survival analysis
  # Setting up countData object
  dds <- DESeqDataSetFromMatrix(
    countData = LUSC_matrix,
    colData = coldata,
    design = ~1
  )

  # Removing genes with sum total of 10 reads across all samples
  dds <- dds[(rowSums(counts(dds)) >= 10), ]

  # vst
  vsd <- vst(dds, blind = FALSE)
  LUSC_matrix_vst <- assay(vsd)

  # Get data for TP53 gene and add gene metadata information to it
  LUSC_TP53 <- LUSC_matrix_vst %>%
    as.data.frame() %>%
    rownames_to_column(var = "gene_id") %>%
    gather(key = "case_id", value = "counts", -gene_id) %>%
    left_join(., gene_metadata, by = "gene_id") %>%
    filter(gene_name == "TP53")

  # get median value
  median_value <- median(LUSC_TP53$counts)

  # Denote which cases have higher or lower expression than median count
  LUSC_TP53$TP53_Expression <- ifelse(LUSC_TP53$counts >= median_value, "High", "Low")

  # Add clinical information
  LUSC_TP53$case_id <- gsub("-01.*", "", LUSC_TP53$case_id)
  LUSC_TP53 <- merge(LUSC_TP53, clin_data, by.x = "case_id", by.y = "submitter_id")

  # Save the data
  save(LUSC_TP53, file = paste0(input_dir, "LUSC_TP53.Rdata"))
} else {
  load(file = paste0(input_dir, "LUSC_TP53.Rdata"))
}
```

# Kaplan-Meier plots
Make Kaplan-Meier plots from TCGA data, split by TP53 expression
```{r}
# Change TP53 survival from days to years
LUSC_TP53$overall_survival_years <- LUSC_TP53$overall_survival / 365

# fitting survival curve
fit <- survfit(Surv(overall_survival_years, deceased) ~ TP53_Expression, data = LUSC_TP53)
fit

ggsurvplot(fit,
  data = LUSC_TP53,
  pval = TRUE,
  risk.table = FALSE,
  #title = "Overall survival by TP53 expression (RNA seq)",
  xlab = "Survival time (years)",
  legend = "right",
  legend.title = "TP53 expression",
  legend.labs = c("High", "Low")
)

# Save plot
ggsave(paste0(output_dir, "Kaplan-Meier by TP53.png"), width = 8, height = 5.5)
```

# Driver gene count
Load COSMIC data to get information on driver mutations.
Data sourced here: https://cancer.sanger.ac.uk/census
Due to messy dataset, there's no way I can see to separate LSCC from other lung cancer types
```{r Driver gene count}
# Load COSMIC data
COSMIC_data <- read.csv(paste0(input_dir, COSMIC_file))

# Get list of tissue types
COSMIC_types <- unique(COSMIC_data$Tumour.Types.Somatic.)

# Filter to lung cancer types - include all
lung_types <- grep(pattern = "lung|SCLC", COSMIC_types, value = TRUE)

# Filter data
COSMIC_filtered <- COSMIC_data %>%
  filter(Tumour.Types.Somatic. %in% lung_types)

# Count overall driver mutations
nrow(COSMIC_filtered)

# Count hallmark genes
sum(COSMIC_filtered$Hallmark == "Yes")

# Select columns and export for use in introduction chapter table
COSMIC_filtered %>%
  filter(Hallmark == "Yes") %>%
  select(
    Gene.Symbol,
    Name,
    Mutation.Types,
    Tumour.Types.Somatic.,
    Role.in.Cancer
  ) %>%
  write.csv(paste0(output_dir, "COSMIC_summary.csv"), row.names = FALSE)
```
